### 2024-01-11 report.Rmd
Add files via upload

---
title: "Fabryka report"
date: "`r Sys.Date()`"
output: html_document
editor_options:
  markdown:
    wrap: 72
params:
  n: NA
---

```{r setup, echo = FALSE, message = FALSE, include = FALSE, warning = FALSE}

suppressPackageStartupMessages({
  library(ggplot2)
})

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

```

This report is made thanks to Fabryka 1.0.

# Method

**Classical method**

The fabric is a statistic made of the orientation and inclination of
elongated sedimentary particles. It provides information about the
origin of the sediments, the nature of the geological processes involved
in the formation of the deposit and their taphonomic impact on the
archaeological remains. The statistical analyses involved are not easily
accessible to non-specialists and are particularly time-consuming. In
addition, the interpretation of the fabric of geological deposits
depends on their comparison with actualist or archaeological references
and therefore requires a good knowledge of geological formation
processes. Recently, a new "spatialised" method has been published and
made reproducible in an R script. The application of this new method to
several paleolithic site assemblages of south west France highlights the
need to create new, better-adapted models. After presenting the classic
statistical tools generally used in the analysis of fabrics, we present
new methods and a new tool to facilitate this analysis. Based on
geoarchaeological studies of several Palaeolithic sites in south-west
France, we propose new models taking in consideration both, the spatial
properties of elongated remains and their nature. Finally, an open
access web application is being made available to make the fabric
analysis easy, quick and reproducible.

Traditionnelement, deux principales méthodes statistiques sont mises en
œuvre afin de caractériser la fabrique des ensembles considérés (Bertran
et Lenoble 2002).

La première s'intéresse à l'orientation des mesures uniquement. Le
calcul du taux d'orientation « r » (ou « L » lorsque exprimé en
pourcentage, Curray 1956) teste l'hypothèse d'une distribution unimodale
des orientations. Si la valeur « p » renvoyée par le test de Rayleigh
est inférieure à un seuil de 0,05, l'hypothèse nulle d'une orientation
aléatoire depuis une distribution uniforme est rejetée. La formule du
calcul du taux d'orientation « r » (ou « L » exprimé en pourcentage) où
« x » est l'orientation en radian et « n » l'effectif de
l'échantillon est la suivante :

(1)$$ r = \sqrt{\sum_{i=\alpha}^n sin{(2\alpha)} + \sum_{i=\alpha}^n cos{(2\alpha)}}  $$

(1.1)$$L =\frac{r*100}{n} $$

La valeur « p » renvoyée par le test de Rayleigh vaut :

(2)$$p = e^{(-(n*L^2))*10^-4}$$

La méthode consistant à doubler la valeur des orientations avant la
réalisation du calcul du taux d'orientation L et du test de Rayleigh
telle que proposée par Krumbein (1939) teste l'hypothèse d'une
distribution bimodale des orientations de période 90°. L'hypothèse d'une
distribution bimodale de période 90° d'un échantillon d'orientations est
retenu lorsque p \< 0,05 (Curray 1956) et lorsque le gain (L -- L) entre
les deux valeurs calculées du taux d'orientation « L » est supérieur à
30%.

La sensibilité du test de Rayleigh à l'effectif « n » empêche la
comparaison des échantillons entre eux lorsque « n » varie (Curray 1956,
Bertran et Lenoble 2002). Par exemple, lorsque le taux d'orientation
« L » est de 25%, le test de Rayleigh ne donne des valeurs de p en
dessous d'un seuil critique de 0,05 que pour des séries dont l'effectif
est supérieur à 50 mesures. Plus généralement, pour un taux
d'orientation "L" équivalent, lorsque « n » augmente la valeur p du test
de Rayleigh diminue (Fig. 1).

```{r echo = FALSE, message = FALSE, fig.cap="Evolution de la valeur p renvoyée par le test de Rayleigh en fonction du taux d’orientation L et de l’effectif n "}

require(ggplot2)
L = c(0:60)

rayleigh <- function(L,n){
  p = exp(-(n*L^2)*10^-4)
  p_value <- p
  p_value
}

p_value_sim_30 <- rep(x = NA, times = length (L))
for (i in 1:61) {
  p_value_sim_30[i]<- rayleigh(i,30)
}

p_value_sim_40 <- rep(x = NA, times = length (L))
for (i in 1:61) {
  p_value_sim_40[i]<- rayleigh(i,40)
}

p_value_sim_50 <- rep(x = NA, times = length (L))
for (i in 1:61) {
  p_value_sim_50[i]<- rayleigh(i,50)
}

p_value_sim_100 <- rep(x = NA, times = length (L))
for (i in 1:61) {
  p_value_sim_100[i]<- rayleigh(i,100)
}

seuil_0.05 = c(rep(0.05, length(L)))

rayleigh_table <- data.frame(L, p_value_sim_30, p_value_sim_40, p_value_sim_50, p_value_sim_100,seuil_0.05)

ggplot(rayleigh_table, aes(x=L)) + 
    geom_line(aes(y = p_value_sim_100, colour = "n = 100"), size = 1) +
    geom_line(aes(y = p_value_sim_30, colour = "n =  30"), size= 1) +
    geom_line(aes(y = p_value_sim_40, colour = "n =  40"), size= 1) + 
    geom_line(aes(y = p_value_sim_50, colour = "n =  50"), size= 1) +
    geom_line(aes(y = seuil_0.05, colour = "seuil de 0.05"), size= 0.3) +
    xlab("Taux d'orientation L (%)") +
    ylab("Valeur p du test de Rayleigh") + 
    scale_color_manual(values=c("#0073C2", "#EFC000", "#868686", "#3B3B3B", "red")) +         scale_x_continuous(breaks = c(0, 10, 20, 30, 40, 50, 60)) +
    scale_y_continuous(breaks = c(0, 0.10, 0.20, 0.30, 0.40, 0.50, 0.60, 0.70, 0.80, 0.90, 1)) +
  theme_minimal()

```

La seconde prend en compte l'orientation et l'inclinaison des vestiges.
Les valeurs propres normalisées E1, E2 et E3 sont calculées à partir du
cosinus que forme l'axe des débris avec les trois axes de l'espace
(Watson 1965, Bertran et Lenoble 2002). Les indices de Benn (1994)
décrivent la forme du nuage de point de la fabrique d'un échantillon.
L'indice d'élongation EL est égal à 1-(E2/E1) et l'indice d'isotropie IS
vaut E3/E1. Lorsque EL est élevé, les axes des objets sont regroupés
autour d'une direction. Lorsque IS est élevé la fabrique est dite
isotrope, les axes sont dispersés. Enfin, lorsque la valeur 1-EL-IS est
élevée la fabrique est dite planaire, les axes des objets sont regroupés
autour d'un plan.

Afin de distinguer différentes populations statistiques, une méthode
d'exploration spatiale proposée par S. McPherron (2018) est utilisée
(fig. 2). Les indices de Benn sont calculés pour chaque vestige, prenant
en compte ses n plus proches voisins en 2 dimensions (McPherron 2018).

**Spatial analysis**

A la différence de S. McPherron (2018), le calcul des indices de Benn
est réalisé à partir de l'orientation et de l'inclinaisons des 40 plus
proches voisins (distance euclidienne) de chaque objet dans un espace en
3D et non en 2D. De plus, comme déjà proposé dans l'une des fonctions du
script associé à l'article publié en 2018, le taux d'orientation « L »
et les tests statistiques sur la distribution des orientations sont
calculés pour chaque série de mesure.

En spatialisant la fabrique des vestiges, la méthode proposée par
McPherron (2018) s'affranchit de l'échantillonnage arbitraire
classiquement réalisé par carré ou par décapage (Bertran et Lenoble
2002). Il est possible de localiser ou de globaliser le signal
statistique d'une série en jouant sur le nombre « n » de mesures
échantillonnées pour calculer les indices de Benn. Le choix porté ici à
40 mesures (chaque vestige et ses 39 plus proches voisins) permet
d'observer des particularités locales, et surtout, il permet la
réalisation du test de Rayleigh dans son domaine d'application
(stabilité du test entre 40 et 50 mesures, selon Bertran et Lenoble
2002). Pour chaque UL, j'ai calculé une moyenne du taux d'orientation
local « Ll » et une moyenne de la valeur locale « pl » pour chacun des
tests effectués (Rayleigh et Rayleigh angles doubles). L'instabilité du
test de Rayleigh ainsi contournée, le taux de rejet Tr de l'hypothèse
testée décrit de manière plus fiable la distribution des orientations
d'un échantillon. De plus, il autorise les comparaisons entre
échantillons malgré des effectifs variables.

(3) $$Tr = \frac{(n times p < 0,05)* 100}{n}$$

Pour chaque échantillon la moyenne locale des indices de Benn (ELl et
ISl pour chaque série de 40 mesures) est distinguée des indices de Benn
globaux (ELg et ISg), calculés sur la totalité de l'échantillon.

L'application de la méthode du calcul des indices de Benn locaux sur des
échantillons de mesures provenant de plusieurs sites archéologiques m'a
permis d'observer l'émergence de différences parfois notables entre
d'une part la distribution générale et la distribution moyenne des
indices de Benn locaux et d'autre part, les indices de Benn globaux. La
fabrique locale est souvent plus proche du pôle linéaire que la fabrique
globale (fig. 2). Plus l'isotropie générale des échantillons en question
est forte, plus les différences semblent marquées. Cela peut s'expliquer
par le fait que des entités proches dans l'espace ont plus de chance de
se ressembler que des entités plus distantes. En termes de fabrique,
cela s'exprime par une plus forte linéarité de la fabrique locale.
Toutefois, une isotropie globale peut parfois masquer une anisotropie
locale. Cela s'exprime soit, par une fabrique localement plus linéaire,
soit, par une fabrique planaire avec une orientation fortement bimodale.
La fabrique « globale » est tout de même calculée pour tester
l'adéquation de l'échantillon avec les processus caractérisés dans le
modèle géoarchéologique proposé par P. Bertran et A. Lenoble (2002, cf.
§ II. a). Deux jeux de données retranscrivent les résultats de ces
analyses statistiques.

L'un s'intéresse aux séries de mesures spatialisées. Ainsi, à l'échelle
du vestige (i.e. l'isobarycentre de la série des 39 plus proches mesures
voisines) les colonnes suivantes apparaissent :

-   L'ensemble de colonnes composées des données brutes déjà présentes
    en amont de l'analyse (tab. 1) ;

-   Orientation des vestiges [0 ; 180] ;

-   Valeur doublée de l'orientation [0 ; 180] ;

-   Indices de Benn locaux « ISl, ELl, EPl » ;

-   Codes RGB correspondant à chaque pôle « rouge, vert, bleu » ;

-   Les coordonnées « X1, Y1, Z1, X2, Y2, Z2 » sont calculées à partir
    de la fabrique et de la position des vestiges. Elles sont
    réutilisées sous QGIS et MMQGIS pour matérialiser l'orientation des
    vestiges par des batônnets. Le code RGB est réutilisé pour
    matérialiser la forme de la fabrique locale ;

-   Valeur « Ll » du taux d'orientation local et la valeur «pl » du test
    de Rayleigh local ;

-   Valeur « Ll double » du taux d'orientation local et la valeur « pl
    double » du test de Rayleigh local calculée via la méthode des
    angles doubles ;

L'autre synthétise les résultats à l'échelle de l'échantillon. Il se
compose des colonnes suivantes :

-   UL investiguée ; Nombre de mesures « n » ; Indices de Benn globaux «
    IS, EL » ;

-   La moyenne des indices de Benn locaux « ISl, ELl » et leurs
    écarts-types « sdIS l » et « sdEL l » ;

-   Orientation moyenne ;

-   Inclinaison moyenne ;

-   Valeur moyenne « pl » du test de Rayleigh local ;

-   Taux d'orientation local moyen « Ll » et son écart-type « sdL l » ;

-   « Tr », soit le taux de rejet de l'hypothèse nulle du test de
    Rayleigh local ;

-   Valeur moyenne « pl » du test de Rayleigh local pour les angles
    doubles ;

-   Taux d'orientation local moyen « Ll double » pour les angles doubles
    et son écart-type « sdL l double » ;

-   « Tr », soit le taux de rejet de l'hypothèse nulle du test de
    Rayleigh local pour les angles doubles.

# Classical method

## Benn diagram

```{r, echo = FALSE, message = FALSE, include = FALSE, warning = FALSE}

# A plot of `params$n` random points. # qu'on peut mettre dans le texte.

# plot(rnorm(params$n), rnorm(params$n))
```

```{r, fig.cap = "Benn diagram"}
print(benn_c)
```

# Spatialised method

## Benn diagram

```{r, fig.cap = "Benn diagram"}

print(benn_sm)

```

```{r, fig.cap = "XY projection"}

print(xy_projection)

```

```{r, fig.cap = "XZ projection"}

print(xz_projection)

```

```{r, fig.cap = "YZ projection"}

print(yz_projection)

```

### 2025-04-13 fabryka/report_3.Rmd
project and git control version creation

---
title: "Fabryka report"
date: "`r Sys.Date()`"
output: pdf_document
editor_options:
  markdown:
    wrap: 72
params:
  n: NA
---

```{r setup, echo = FALSE, message = FALSE, include = FALSE, warning = FALSE}

suppressPackageStartupMessages({
  library(ggplot2)
})

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

```

This report is made thanks to Fabryka 1.0.

# Tutorial
## Uploading your data

First, upload your .csv file in the left sidebar panel. Pay attention to your decimal and column delimiters.

Fabryka supports three angle measurement methods:

- angles taken with a compass and an inclinometer
- angles taken with a DistoX2
- angles taken with two shots at the total station.

Five cases (i.e. forms of data) are covered by the application:

- check 'Case 1: Only angles' if you have in your dataset the columns 'id', 'sample', 'code', direction' and 'dip'.
- check 'Case 2: Only angles from DistoX2' if you have in your dataset the same columns that above but your data comes from a DistoX2 and needs to be converted.
- check 'Case 3: Angles from DistoX2 with coordinates' if you have in your dataset the same columns that above (Case 2, data from DistoX2) with a single set of coordinates (x, y, z).
- check 'Case 4: Angles and coordinates' if your data has the same form that above (Case 3) but is coming from a compass and a clinometer.
- check 'Case 5: Two shots data without angles' if you have the columns 'id', 'sample', 'code' and the angles have been measured with a total station (i.e. two shots data).

If you miss a context column, please create it by repeating a same character.

According to your method of angle measurement and the presence of coordinates you have to select the right case. Please refer to the example files in the left side bar panel. It contains a dataset coming from Le Moustier Lower shelter (Dordogne, France). The exmample files contain directly the good column names. You can use them directly in the application. Whatever the names of the columns in your file, you can select the appropriate columns in the column selector. Please keep only complete lines (i.e. no empty modalities, so no line without a fabric measurement). When a data table with new columns appears below the column selector, you are ready to use the application.

If your data comes from a DistoX2, the columns 'direction' (= 'bearing', 0 to 360° angles) and 'dip' (= 'plunge') are automatically corrected in the new table. Whatever the form of your dataset, two new columns 'orientation_pi' (0 to 180° angles) and 'angle_double' (see Krumbein 1939) appear. If the angles were measured using a total station (i.e. two shots data), the bearing and the plunge are also calculated. If only one set of coordinates is provided with angles, a second set of coordinates is calculated and displayed in the new table. You can download the new data by pushing the 'copy', the 'csv' or the 'pdf' buttons.

## Classical method

The classical method is available in all cases (i.e. source of data and absence or presence of coordinates). You can control the data you plot and the data appearing in the figures and the summary data table by changing the checkboxes filters (samples and code) and the slider (i.e. minimum number of measures) in the left sidebar panel. Certain geological processes can have a different impact on objects depending on their physical properties (e.g. silex versus bones in a water flow as in the example files provided where bones have a strong preferential orientation and are much more 'linear' in the Benn diagram, see Thomas et al., 2019, Texier et al., 2020). The natural processes models that you can add to the Benn diagram are calculated using data from Bertran & Lenoble (2002). In the rose diagram, orientations are considered (and not directions) as the Rayleigh test is perform on orientations (see Curray 1956) in the summary table. You can download all the figures by selecting a file format and pressing the 'download' button.

In this last table:

-  'n' stands for number of measures in the samples
- 'E1', 'E2' and 'E3' are the eigenvalues (see Woodcock 1977). They correspond to the sum of the cosines formed by the axis of the objects and the three axis of the space. They are used to compute the Benn indices.
- 'IS', 'PL' and 'EL' are calculated thanks to the eigenvalues (Benn 1994). These are the Benn indices, ratios standing respectively for, 'isotropy' with IS = E3/E1, 'elongation' with EL = 1-(E2/E1) and 'planar' PL = 1-IS-EL
- 'L' is the vector magnitude (strength of the preferred artefact orientation) and 'R.p' is the p-value result of the Rayleigh test.
- 'L.double' and 'R.p.double' are the same statistics that above for double angles (see Krumbein 1939).

Curray's formula to calculate 'L' is (1.1)

(1.1)$$L =\frac{r*100}{n} $$

with (1.2)

(1.2)$$ r = \sqrt{\sum_{i=\alpha}^n sin{(2\alpha)} + \sum_{i=\alpha}^n cos{(2\alpha)}}  $$

where 'n' is the number of measures and 'α', is the orientation (0 to 180°)

The Rayleigh test is perform thanks to the formula (1.3)

(2)$$p = e^{(-(n*L^2))*10^-4}$$

## Spatialised method

In order to distinguish different statistical populations, a spatial analysis method proposed by McPherron (2018) is used. Thanks to this methods it is possible to calculate the Benn indices and to test the orientation rate L (formula (1.1) modified R McPherron 2018 script computed for orientations and double angles) for each remains, taking into account its n nearest neighbors in 3 dimensions (R McPherron 2018 script, modified , formula (2.)). The analysis of the fabric as proposed by McPherron (2018) has the advantage of spatializing the measurements and thus avoiding the arbitrary sampling classically carried out by square or by décapage. It is possible to localize the statistical signal of a series by varying the number of measurements sampled to calculate the Benn indices. The choice of about 40 to 50 measurements (each vestige and its 40 to 50 closest neighbors) to calculate the Benn indices allows local particularities to be observed, and above all, it allows the Rayleigh test to be carried out in its field of application (stability of the test between 40 and 50 measurements according to Bertran and Lenoble 2002). Please note that the Rayleigh test is sample size sensible (fig. 1). 
Several modes of representation are used in the 'Benn diagram & projections' subpanel.

The Benn indices (1994) calculated for each series of remains are projected within a ternary diagram in which each pole (isotropic, linear and planar) is colored respectively in red, green and blue. The RGB color code is reused to materialize the fabric of the remains on spatial projections of the material along the axes of the excavation grid. Projections (xy, xz and yz axes) are performed. On the latter, the orientation of the objects can be materialized by sticks (by uncecking the corresponding box in the left sidebar panel) which color is representative of the fabric of the series formed by the vestige and its 50 closest neighbors. In the left sidebar panel, you can control the samples taken into account in the analyses by selecting the sample explored or by distinguishing the code used (e.g. flint versus bone). You can also use the slider to change the number of neighbours searched for around each point. Be careful, this will alter the results obtained in the Benn diagram and Rayleigh tests.

At the spatial series scale (i.e. each points and n neighbors), statistics are provided in a table (Spatial series data table subpanel) with the Benn indices, the orientation rates L, L.double and p-values returned by the Rayleigh tests. At the sample scale, mean and standard deviations values of Benn indices, mean L and its corresponding Rayleigh tests means (mean values of all spatial series in the sample) and rejections ratios 'Tr' (formula 2.) are given in the subpanel 'Summary table of samples'.

The rejection ratio 'Tr' is perform thanks to the formula (2.)

(2) $$Tr = \frac{(n times p < 0,05)* 100}{n}$$

## Spatial exploration

In this panel, you can perform a spatialised analysis of your data starting from a Benn diagram or from projections. In both cases, you need to select the data you want to explor more locally in a new analysis. From the projections, you can select the points you want to explore directly. In the Benn diagram, you need to select the lasso selector. Pay attention to the number of points selected and check that the new sample contains enough points to carry out the new analysis. If this is not the case, you can lower the slider in the left sidebar panel to perform the analysis with fewer neighbours. You can dowload the new samples you have selected from the corresponding subpanels.

## Markdown report

This panel is in progress. It will be available in the next version of the application. By pushing the download button you will find only a beta test with the tutorial.

```{r echo = FALSE, message = FALSE, fig.cap="Evolution of the p-value returned by the Rayleigh test as a function of the orientation rate L and the number of measures n."}

require(ggplot2)
L = c(0:60)

rayleigh <- function(L,n){
  p = exp(-(n*L^2)*10^-4)
  p_value <- p
  p_value
}

p_value_sim_30 <- rep(x = NA, times = length (L))
for (i in 1:61) {
  p_value_sim_30[i]<- rayleigh(i,30)
}

p_value_sim_40 <- rep(x = NA, times = length (L))
for (i in 1:61) {
  p_value_sim_40[i]<- rayleigh(i,40)
}

p_value_sim_50 <- rep(x = NA, times = length (L))
for (i in 1:61) {
  p_value_sim_50[i]<- rayleigh(i,50)
}

p_value_sim_100 <- rep(x = NA, times = length (L))
for (i in 1:61) {
  p_value_sim_100[i]<- rayleigh(i,100)
}

seuil_0.05 = c(rep(0.05, length(L)))

rayleigh_table <- data.frame(L, p_value_sim_30, p_value_sim_40, p_value_sim_50, p_value_sim_100,seuil_0.05)

ggplot(rayleigh_table, aes(x=L)) + 
    geom_line(aes(y = p_value_sim_100, colour = "n = 100"), size = 1) +
    geom_line(aes(y = p_value_sim_30, colour = "n =  30"), size= 1) +
    geom_line(aes(y = p_value_sim_40, colour = "n =  40"), size= 1) + 
    geom_line(aes(y = p_value_sim_50, colour = "n =  50"), size= 1) +
    geom_line(aes(y = seuil_0.05, colour = "seuil de 0.05"), size= 0.3) +
    xlab("Taux d'orientation L (%)") +
    ylab("Valeur p du test de Rayleigh") + 
    scale_color_manual(values=c("#0073C2", "#EFC000", "#868686", "#3B3B3B", "red")) +         scale_x_continuous(breaks = c(0, 10, 20, 30, 40, 50, 60)) +
    scale_y_continuous(breaks = c(0, 0.10, 0.20, 0.30, 0.40, 0.50, 0.60, 0.70, 0.80, 0.90, 1)) +
  theme_minimal()

```

### 2025-04-13 fabryka/fabryka_beta_1.0.R

@@ -1,0 +1,4329 @@
+### --- R packages ---
+
+library(shiny)
+library(dplyr)
+library(ggplot2)
+library(plotly)
+library(ggsci)
+library(ggtern)
+library(RColorBrewer)
+library(circular)
+library(CircStats)
+library(Ternary)
+library(cowplot)
+library(RStoolbox)
+library(shinyjs)
+library(rmarkdown)
+library(rsconnect)
+
+
+### --- Model data from Bertran & Lenoble 2002 ---
+# model <- read.csv2("C:/Users/Surface User/Documents/articles/fabriques/fabryka/model.csv")
+# ternary_model <- model %>%
+#   dplyr::filter(area != "NA")
+
+### --- Model data from Bertran & Lenoble 2002 ---
+ternary_model <- data.frame(
+  Type = c("Snow Avalanche", "Snow Avalanche", "Snow Avalanche", "Snow Avalanche", "Snow Avalanche", "Snow Avalanche", "Snow Avalanche", "Snow Avalanche", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Mudslides", "Mudslides", "Runoff Shallow", "Runoff Shallow", "Runoff Shallow", "Runoff Shallow", "Runoff Shallow", "Runoff Shallow", "Runoff Shallow", "Runoff Shallow", "Runoff Shallow", "Runoff Shallow", "Runoff Shallow", "Runoff Shallow", "Runoff Shallow", "Runoff Shallow", "Runoff Shallow", "Runoff Shallow", "Runoff Steep", "Runoff Steep", "Runoff Steep", "Runoff Ste
[…]

### 2025-04-13 fabryka/fabryka_beta_2.0.R

@@ -1,4 +1,5 @@
 ### --- R packages ---
+
 library(shiny)
 library(dplyr)
 library(ggplot2)
@@ -14,22 +15,551 @@ library(RStoolbox)
 library(shinyjs)
 library(rmarkdown)
 library(rsconnect)
+library(shinythemes)
+library(reactlog)
+library(ggalt)
 
 
-### --- Model data from Bertran & Lenoble 2002 ---
+### --- Model data from Bertran & Lenoble 2002 (McPherron 2018)---
 # model <- read.csv2("C:/Users/Surface User/Documents/articles/fabriques/fabryka/model.csv")
 # ternary_model <- model %>%
 #   dplyr::filter(area != "NA")
 
 ### --- Model data from Bertran & Lenoble 2002 ---
+# ternary_model <- data.frame(
+#   Type = c("Snow Avalanche", "Snow Avalanche", "Snow Avalanche", "Snow Avalanche", "Snow Avalanche", "Snow Avalanche", "Snow Avalanche", "Snow Avalanche", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Mudslides", "Mudslides", "Runoff Shallow", "Runoff Shallow", "Runoff Shallow", "Runoff Shallow", "Runoff Shallow", "Runoff Shallow", "Runoff Shallow", "Runoff Shallow", "Runoff Shallow", "Runoff Shallow", "Runoff Shallow", "Runoff Shallow", "Runoff Shallow", "Runoff Shallow", "Runoff Shallow", "Runoff Shallow", "Runoff Steep", "Runoff Steep", "Runoff Steep",
[…]

### 2025-04-13 fabryka/fabryka_beta_3.0.R
correcting errors in app text

@@ -805,7 +805,7 @@ ui <- tagList(
   useShinyjs(),
   navbarPage(
     theme = shinytheme("yeti"),
-    p(strong("Fabryka"), "1.0"),
+    p(strong("Fabryka \\(\\beta\\) 3.0")),
     tabPanel(
       "Presentation",
       mainPanel(
@@ -818,7 +818,7 @@ ui <- tagList(
           align = "justify",
           h2("Presentation"),
           hr(),
-          p("Fabryka is a R shiny application dedicated to the fabric analysis of archaeological remains. This is the \\(\\beta\\)-test version 3.0. The application allows a deep exploration of the fabric data. An article, a tutorial video and the source code are available via the links respectively", strong("here,"), strong("here"), "and", a(href = "https://gist.github.com/4211337", strong("here")), "If you use fabryka, please cite the article."),
+          p("Fabryka is a R shiny application dedicated to the fabric analysis of archaeological remains. The app' is developped by Marc THOMAS", a(href = "https://traces.univ-tlse2.fr/accueil/equipes-et-ateliers/smp3c-societes-et-milieux-des-populations-de-chasseurs-cueilleurs-collecteurs/marc-thomas", strong("lab' web page.")), "This is the \\(\\beta\\)-test version 3.0. The application allows a deep exploration of the fabric data.", "An article and the source code are available via the links respectively", strong("here,"), "and", a(href = "https://github.com/marchaeologist/fabryka/tree/main", strong("here.")), "If you use fabryka, please cite the article."),
           tags$br(),
           p(strong("The 'Upload data'"), "panel allows the user to upload different data sources."),
           tags$br(),
@@ -846,7 +846,7 @@ ui <- tagList(
           tags$br(),
           p(strong("The 'Bibliography'"), "panel contains the bibliography cited in the application."),
           tags$br(),
-          p(strong("The 'Contributors'"), "contains acknowledgements from contributors to the creation of the application."),
+          p(strong("The 'Contributors'"), "contains acknowledgements and the list of contributors to the creation of the application."),
           tags$br(),
           p(strong("The 'Citation'"), "describes how to cite the application, the methods and datasets used."),
           tags$br(),
@@ -1792,8 +1792,7 @@ ui <- tagList(
                h2("Data contributions"),
                hr(),
                tags$ul(
[…]

### 2025-04-13 fabryka_pci/article.Rmd
Bib done. To be checked.

---
title: '**Fabryka**'
subtitle: A web application for easily analysing & exploring the fabric of archaeological assemblages
output:
  pdf_document: default
header-includes:
- \usepackage{fancyhdr}
- \pagestyle{fancy}
- \fancyhead[L]{Fabryka}
- \fancyhead[R]{R web application for fabric analysis}
- \fancyfoot[C]{\thepage}
- \usepackage{float}
- \floatplacement{figure}{H}
editor_options:
  markdown:
    wrap: 72
  chunk_output_type: console
bibliography: references.bib
link-citations: true
---

```{=html}
<style>
body {
text-align: justify}
</style>
```

Marc THOMAS ^1\*^ \bigbreak \bigbreak

^1^ TRACES, UMR 5608 - Université Toulouse Jean Jaurès, 5 Allée Antonio
Machado, FR-31058 Toulouse Cedex 9 \smallbreak

^\*^ Corresponding author \smallbreak

E-mail: [marcthomas1\@hotmail.fr](mailto:marcthomas1@hotmail.fr){.email}
\bigbreak \bigbreak

**Key words** : R shiny web application, fabric, site formation
processes, spatial analyses, orientation and inclination \bigbreak
\bigbreak

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Abstract

This article presents Fabryka, an open access R shiny web application
dedicated to the fabric analysis of archaeological or geological
assemblages. The fabric is a statistic made of the orientation and
inclination of elongated sedimentary particles and archaeological
objects. It provides information about the source of sediments, the
nature of the geological processes involved in the formation of the
deposits and their impact on the preservation of the archaeological
assemblages. These statistical analyses are not easily accessible to
non-specialists and are particularly time-consuming, sometimes requiring
the use of several software, some of which are chargeable. Fabryka uses
classical analytical procedures (Benn, rose and Schmidt diagrams and
statistical tests on orientations) for analysing the fabrics of objects
and includes a spatial method. New statistical methods and new ways of
exploration are proposed in the application and this present article.
Also, a call for contributions is proposed to construct a new reference
model with spatial information. \smallbreak

You can use Fabryka $\beta$ 3.0 by following this link
<https://marchaeologist.shinyapps.io/fabryka/>. The version will be
updated to Fabryka 1.0 after the review process of this article. The
Fabryka application files, made in R [@rcoreteam2025] by using *shiny*
package [@chang2024], and this article, written in rmarkdown
[@allaire2014; @xie2018; @xie2020], are both available online in a
public github repository
(<https://github.com/marchaeologist/fabryka/tree/main>). When reviewing
this article, you can either clone the repository, or push on the
changes via github.\pagebreak

# Introduction

Fabric is made of the orientation and inclination of elongated
sedimentary particles [@bertran1997; @bertran2002; @lenoble2004]. It
is a statistic (*i.e.*, it needs as much measurements as possible) that
enables the identification of sediment sources that make up the deposits
[@texier2020], sedimentary processes involved in their formation
[@bertran1997; @bertran2002; @lenoble2004] and their impact on
archaeological assemblages *e.g.* [@isaac1967; @schick1986;
@thomas2019]. Objects abandoned by prehistoric people can be considered
as sedimentary particles and are affected by the same processes
[@schiffer1983]. As geological particles are often not elongated
enough, the fabric analysis made of the sedimentary deposits depends
mainly on the archaeological material (bones and lithics), which enables
them to be inserted into the formation dynamics of the site and to
obtain information on their representativeness for the reconstruction of
past activities.

One of the first archaeological application probably dates to the
experiments carried out by the South African G. Isaac (1967) on the
shores of Lake Magadi in Kenya. G. Isaac wanted to explain the role
played by hydrological processes on the bone accumulations found in the
Great Rift Valley. In November 1964, he placed several concentrations of
experimental material near a river branch. A few months later, heavy
rains had reactivated the valley's rivers. In addition to the
displacement or even loss of all or part of the material, G. Isaac
revealed a reorientation that was mainly perpendicular and, to a lesser
extent, parallel to the direction of flow, with a slight upstream
inclination. Since G. Isaac's pioneering work, there has been an
abundant literature devoted to the analysis of fabrics *e.g.*
[@benn1994; @lenoble2004; @mcpherron2018] allowing this method to be
increasingly performed on archaeological sites. Several software allow
data processing, some of which are not freely accessible, do not allow
spatial exploration of the data or do not offer interfaces to facilitate
the task of non-programmer users *e.g* [@mcpherron2018].

The following sections explain all the selected approaches and our
proposals for processing the fabrics thanks to fabryka $\beta$ 3.0, from
taking measurements in the field, through the statistical methods and
graphical modes of representation used in the application (fig.
\ref{fig:figure_presentation}). \bigbreak

\begin{figure} [h]
\centering
  \includegraphics[width=16cm]{figure/presentation_applications.png}
  \caption{Fabryka creation scheme and organisation of the article}
\label{fig:figure_presentation}
\end{figure}

# Field data acquisition

During excavation, the fabric of elongated remains is generally
systematically measured. The direction (= bearing) of the lowest point
is measured in the range [0; 360] with North as the reference (i.e.
direction = 0°). The inclination (= plunge, dip) is measured in the
range [0; 90] with horizontal as a reference. The direction [0; 360] can
be converted into an orientation measured in the range [0; 180] by
subtracting 180° from the measurements in the range ]180; 360].

Traditionally, the direction of the objects lowest point is measured
using a compass and the inclination is measured using an inclinometer.
The aim is to measure the direction and inclination of the physical axis
of elongation of objects ignoring its shape. To do this, the measuring
device must be placed parallel to the general axis of elongation of the
piece without touching it. Measuring the fabric from its imprint or
visible surface is biased by its shape, making impossible to assess its
true inclination. Since the theodolite was introduced to excavations in
the 1990s, the fabrics is sometimes measured by recording the
coordinates of the ends of objects. In the latter case, the rules of
trigonometry are applied to calculate directions and inclinations.

Recently, speleologist have created a digital inclinometer (Leica
DistoX310 modified to DistoX2; <https://paperless.bheeb.ch/>). The
device stores and associates data with the order of measurement. When
the device is tilted downwards, the direction of the lowest point is
measured and, the recorded inclination is negative. When the device is
tilted upwards, the direction of the highest point is recorded, and the
inclination is positive. To obtain data that is suitable for statistical
analysis, in the first case the absolute value of the inclination must
be retained. In the second case (*i.e.* positive inclination), if the
measured direction is in the interval [0; 180], 180° must be added to
the direction, and if the measured direction is in the interval ]180;
360], 180° must be subtracted from the direction.

When possible, preferring the use of a digital inclinometer avoids many
errors during measurement and data recording. On the other hand, for the
reasons given above, the use of a theodolite to measure fabrics is not
recommended as it results in inaccurate inclination measurements. As
well, measuring orientations and inclinations in sections leads to
sampling biases.

A sample corresponds to all the direction and inclination measurements
taken within a Lithostratigraphic Unit (UL). UL seems to be the most
coherent entity which needs to be explored because they are defined
according to sedimentary criterion, which a priori guarantee a certain
coherence in terms of the formation dynamics we are seeking to approach.
In Fabryka, analyses are constrained within samples unless the user
decides otherwise in the spatial exploration panel. \bigbreak

# Mathematics

## Classical method

Two main statistical methods are used:

1.  The first considers the orientation only. The calculation of the
    orientation rate "r" (or "L" when expressed as a percentage, Curray
    formula [@curray1956] tests the hypothesis of a unimodal
    distribution of orientations. If the "p" value returned by the
    Rayleigh test is less than a threshold of 0.05, the null hypothesis
    of a random orientation from a uniform distribution is rejected. The
    formula for calculating the orientation rate "r" where '$\alpha$' is
    the orientation in radians and "n" is the sample size is as follows:

Curray's formula to calculate 'L' is (**1.**)

$$L = \frac{r*100}{n}$$ with (**2.**)

$$r = \sqrt{\sum_{i=\alpha}^n sin{(2\alpha)} + \sum_{i=\alpha}^n cos{(2\alpha)}}$$
\smallbreak where 'n' is the number of measures and '$\alpha$', is the
orientation (0 to 180°) \smallbreak

The Rayleigh test is performed thanks to the formula (**3.**)

$$p = e^{(-(n*L^2))*10^-4}$$

The method consisting of doubling the value of the orientations before
calculating the orientation rate L and the Rayleigh test, as proposed by
Krumbein [@krumbein1939], tests the hypothesis of a bimodal
distribution of orientations with a period of 90°. The hypothesis of a
90° period bimodal distribution of a sample of orientations is retained
when p \< 0.05 [@curray1956]. The sensitivity of the Rayleigh test to
the sample size 'n' prevents samples from being compared with each other
when 'n' varies [@curray1956; @bertran2002]. For example, when the
orientation rate "L" is 25%, the Rayleigh test gives p-values below a
critical threshold of 0.05 for series with more than 50 measurements
(fig. \ref{fig:rayleigh}).

2.  The second considers the direction and inclination of the objects.
    The normalised eigenvalues E1, E2 and E3 are calculated from the
    cosine formed by the axis of the objects with the three axes of
    space [@watson1966]. The Benn indices [@benn1994] describe the
    fabric shape of a sample. The elongation index EL is equal to
    1-(E2/E1) and the isotropy index IS is E3/E1. When EL is high, the
    axes of the objects are grouped around a direction. When IS is high,
    the fabric is said to be isotropic meaning the axes of the objects
    show no order. Finally, when the value 1-EL-IS is high, the fabric
    is said to be planar, with the axes of the objects grouped around a
    plane. \bigskip

```{r rayleigh, echo = FALSE, warning = FALSE, message = FALSE, fig.cap="Evolution of the p-value returned by the Rayleigh test as a function of the orientation rate 'L' and the sample size 'n'", fig.height = 3.5, fig.pos="f"}

require(ggplot2)
L = c(0:60)

rayleigh <- function(L,n){
  p = exp(-(n*L^2)*10^-4)
  p_value <- p
  p_value
}

p_value_sim_30 <- rep(x = NA, times = length (L))
for (i in 1:61) {
  p_value_sim_30[i]<- rayleigh(i,30)
}

p_value_sim_40 <- rep(x = NA, times = length (L))
for (i in 1:61) {
  p_value_sim_40[i]<- rayleigh(i,40)
}

p_value_sim_50 <- rep(x = NA, times = length (L))
for (i in 1:61) {
  p_value_sim_50[i]<- rayleigh(i,50)
}

p_value_sim_100 <- rep(x = NA, times = length (L))
for (i in 1:61) {
  p_value_sim_100[i]<- rayleigh(i,100)
}

seuil_0.05 = c(rep(0.05, length(L)))

rayleigh_table <- data.frame(L, p_value_sim_30, p_value_sim_40, p_value_sim_50, p_value_sim_100,seuil_0.05)

ggplot(rayleigh_table, aes(x=L)) + 
    geom_line(aes(y = p_value_sim_100, colour = "n = 100"), size = 1) +
    geom_line(aes(y = p_value_sim_30, colour = "n =  30"), size= 1) +
    geom_line(aes(y = p_value_sim_40, colour = "n =  40"), size= 1) + 
    geom_line(aes(y = p_value_sim_50, colour = "n =  50"), size= 1) +
    geom_line(aes(y = seuil_0.05, colour = "threshold 0.05"), size= 0.5) +
    xlab("orientation rate 'L' (%)") +
    ylab("p-value") + 
    scale_color_manual(values=c("#0073C2", "#EFC000", "#868686", "#3B3B3B", "red")) +         scale_x_continuous(breaks = c(0, 10, 20, 30, 40, 50, 60)) +
scale_y_continuous(breaks = c(0, 0.10, 0.20, 0.30, 0.40, 0.50, 0.60, 0.70, 0.80, 0.90, 1)) +
  theme_minimal() +
  theme(legend.position = "bottom") + 
  labs(colour = "")

```

## Spatialised method

To distinguish different statistical populations, a spatial method
proposed by McPherron (2018) is used. Thanks to this methods it is
possible to calculate the Benn indices and to test the orientation rate
'L' (formula (**3.**) modified R [@mcpherron2018] script) for each
remains, and its n nearest neighbours in 3 dimensions (R McPherron 2018
script, modified , formula (**4.**)). The analysis of the fabric as
proposed by McPherron (2018) has the advantage of spatialising the
measurements and thus avoiding the arbitrary sampling classically
carried out by square or by décapage. It is possible to localize the
statistical signal of fabrics by varying the number of measurements
sampled to calculate the Benn indices. The choice of about 40 to 60
measurements (each remain and its 40 to 60 nearest neighbours) to
calculate the Benn indices allows local particularities to be observed,
and above all, it allows the Rayleigh test to be carried out in its
field of application (stability of the test between around 40 to 50
measurements after [@bertran2002]).

By avoiding the instability of the Rayleigh test, the rejection rate
'Tr' of the null hypothesis provides a more reliable description of the
distribution of orientations in a sample. In addition, it allows
comparisons between samples despite varying sample size.

The rejection ratio 'Tr' is performed thanks to the formula (**4.**)

$$Tr = \frac{(n~times~p.value < 0.05)* 100}{n}$$

# Graphical representation

Several methods of representation are used. Rose diagrams summaries the
orientation and inclination measurements for each sample. Benn indices
are projected in a Benn diagram and measures (directions and plunge) are
projected into Schmidt diagrams. These figures are available for both
'classical' and 'spatialised' methods.

As proposed by McPherron (2018), a spatial investigation of the fabric
is proposed in a dedicated panel. Benn indices (1994) calculated for
each spatial series of remains are projected within a Benn diagram in
which each pole (isotropic, linear, and planar) is coloured respectively
in red, green and blue. The RGB colour code is reused to materialize the
fabric of the remains on spatial projections of the material along the
axes of the excavation grid. Projections (xy, xz and yz axes) are
performed. \bigbreak

# Interface and user interactivity

The next sections provide a description and a tutorial of Fabryka.

## Upload your data

First, upload your .csv file in the left sidebar panel (fig.
\ref{fig:figure_upload_your_data}A). Pay attention to your decimal and
column delimiters (fig. \ref{fig:figure_upload_your_data}B and C).

Fabryka supports three angle measurement methods:

-   angles taken with a compass and an inclinometer
-   angles taken with a DistoX2
-   angles taken with two shots at the total station.

Five cases (i.e. forms of data) are covered by the application (fig.
\ref{fig:figure_upload_your_data}D):

-   check 'Case 1: Only angles' if you have in your dataset the columns
    'id', 'sample', 'code', direction' and 'dip'.
-   check 'Case 2: Only angles from DistoX2' if you have in your dataset
    the same columns that above but your data comes from a DistoX2 and
    needs to be converted.
-   check 'Case 3: Angles from DistoX2 with coordinates' if you have in
    your dataset the same columns that above (Case 2, data from DistoX2)
    with a single set of coordinates (x, y, z).
-   check 'Case 4: Angles and coordinates' if your data has the same
    form that above (Case 3) but is coming from a compass and a
    clinometer.
-   check 'Case 5: Two shots data without angles' if you have the
    columns 'id', 'sample', 'code' and the angles have been measured
    with a total station (*i.e.* two shots data).

\begin{figure} [H]
\centering
  \includegraphics[width=17cm]{figure/upload_your_data.jpg}
  \caption{A. Browse your .csv file, B. choose your column separator, C. choose your decimal separator, D. check the right form of your dataset, E. to do so, have a look of example files, F. select the column needed in the application, G. a new dataset with new columns appears, your are ready to explore your data.}
\label{fig:figure_upload_your_data}
\end{figure}

If you miss a context column, please create it by repeating a same
character.

According to your method of angle measurement and the presence of
coordinates you must select the right case. Please refer to the example
files in the left side bar panel (fig.
\ref{fig:figure_upload_your_data}E). It contains a dataset coming from
Le Moustier Lower shelter (Dordogne, France). The example files directly
contain the good column names. You can use them directly in the
application. Whatever the names of the columns in your file, you can
select the appropriate columns in the column selector (fig.
\ref{fig:figure_upload_your_data}F). Please keep only complete lines
(*i.e.* no empty modalities, so no line without a fabric measurement).
When a data table with new columns appears below the column selector,
you are ready to use the application.

If your data comes from a DistoX2, the columns 'direction' (= 'bearing',
0 to 360° angles) and 'dip' (= 'plunge') are automatically corrected in
the new table. Whatever the form of your dataset, two new columns
'orientation_pi' (0 to 180° angles) and 'angle_double' (see Krumbein
1939) appear. If the angles were measured using a total station (*i.e.*
two shots data), the bearing and the plunge are also calculated. If only
one set of coordinates is provided with angles, a second set of
coordinates is calculated and displayed in the new table (fig.
\ref{fig:figure_upload_your_data}G). You can download the new data by
clicking the 'copy', the 'csv' or the 'pdf' buttons. \bigbreak

## Classical method

The classical method (fig. \ref{fig:figure_classical_method}) can be
computed for all sources of data and absence/presence of coordinates
(*i.e.* all cases described above).

This panel includes:

-   Benn diagrams with possible integration of Bertran and Lenoble
    (2002) model (fig. \ref{fig:figure_classical_method}D)
-   Rose diagrams for bearing and plunge (fig.
    \ref{fig:figure_classical_method}G)
-   Schmidt diagrams (fig. \ref{fig:figure_classical_method}H)
-   Summary data table with Benn indices, additional statistics, and
    orientation statistical tests.

In this last table (fig. \ref{fig:figure_classical_method}I):

-   'n' stands for number of measures in the samples
-   'E1', 'E2' and 'E3' are the eigenvalues (see Woodcock 1977). They
    correspond to the sum of the cosines formed by the axis of the
    objects and the three axis of the space. They are used to compute
    the Benn indices
-   'IS', 'PL' and 'EL' are calculated thanks to the eigenvalues (Benn
    1994). These are the Benn indices, ratios standing respectively for,
    'isotropy' with IS = E3/E1, 'elongation' with EL = 1-(E2/E1) and
    'planar' PL = 1-IS-EL
-   'L' is the vector magnitude (strength of the preferred artefact
    orientation) and 'R.p' is the p-value result of the Rayleigh test.
-   'L.double' and 'R.p.double' are the same statistics that above for
    double angles (Krumbein 1939).

You can control the data you plot and the data appearing in the figures
and the summary data table by changing the checkboxes filters (samples
and code) and the slider (*i.e.* minimum number of measures) in the left
sidebar panel (fig. \ref{fig:figure_classical_method}A, B and C).
Certain geological processes can have a different impact on objects
depending on their physical properties (*e.g.* silex versus bones in a
water flow as in the example files provided where bones have a strong
preferential orientation and are much more 'linear' in the Benn diagram,
see [@thomas2019] and [@texier2020]). The natural processes models that
you can add to the Benn diagram are calculated using data mainly from
[@bertran1997; @bertran2002; @bertran2006; @lenoble2004]. In the rose
diagram, orientations are considered (and not directions) as the
Rayleigh test is perform on orientations (see Curray 1956) in the
summary table. You can download all the figures by selecting a file
format (.pdf, .jpeg and .png) and pressing the 'download' button (fig.
\ref{fig:figure_classical_method}E and F).

\begin{figure} [H]
\centering
  \includegraphics[width=17cm]{figure/classical_method.jpg}
  \caption{A. filter by sample, B. if you want to filter by codes, uncheck 'no code distinction', C. manage the minimum sample size to perform analyses, D. buld your model thanks to Bertran and Lenoble (2002) data, E. and F. download figures by selecting a file format and pressing the 'download' button, G. rose diagrams subpanel, H. Schmidt diagram subpanel, I. summary table with additional statistics}
\label{fig:figure_classical_method}
\end{figure}

## Spatialised method

This panel includes (fig. \ref{fig:figure_spatialised_method}):

-   Benn diagrams with Benn indices computed with nearest neighbours
    points (McPherron 2018)
-   Spatial projections of the points according to their fabric shape
-   Summary data tables with orientation statistical tests, one at the
    scale of each spatial series (each object and its n nearest
    neighbours = one line) and one at the scale of the sample (fig.
    \ref{fig:figure_spatialised_method}F and G).

In these tables:

-   'n' stands for number of measures in the samples
-   'E1', 'E2' and 'E3' are the eigenvalues [@woodcock1977]. They
    correspond to the sum of the cosines formed by the axis of the
    objects and the three axes of the space. They are used to compute
    the Benn indices.
-   'IS', 'PL' and 'EL' are calculated thanks to the eigenvalues (Benn
    1994). These are the Benn indices, ratios standing respectively for,
    'isotropy' with IS = E3/E1, 'elongation' with EL = 1-(E2/E1) and
    'planar' PL = 1-IS-EL
-   'L' is the vector magnitude (strength of the preferred artefact
    orientation) and 'R.p' is the p-value result of the Rayleigh test.
-   'L.double' and 'R.p.double' are the same statistics that above for
    double angles (see Krumbein 1939).

At the spatial series scale (*i.e.* each points and n neighbours),
statistics are provided in a table (Spatial series data table subpanel)
with the Benn indices, the orientation rates L, L.double and p-values
returned by the Rayleigh tests. At the sample scale, mean and standard
deviations values of Benn indices, mean L and its corresponding Rayleigh
tests means (mean values of all spatial series in the sample) and
rejections ratios 'Tr' (formula **4.**) are given in the subpanel
'Summary table of samples'.

In the projections, the orientation of the objects can be materialized
by sticks (by unchecking the corresponding box in the left sidebar
panel, fig. \ref{fig:figure_spatialised_method}D) which colour is
representative of the fabric of the series formed by the object and its
n nearest neighbours. You can also add orthophoto in the background of
projections (fig. \ref{fig:figure_spatialised_method}E). In the left
sidebar panel, you can control the samples considered in the analyses by
selecting the sample explored or by distinguishing the code used (*e.g.*
flint versus bone, fig. \ref{fig:figure_spatialised_method}A and B). You
can also use the slider to change the number of neighbors searched for
around each point (fig. \ref{fig:figure_spatialised_method}C). Be
careful, this will alter the results obtained in the Benn diagram and
Rayleigh tests. \bigbreak

## Spatial exploration

This panel includes (fig. \ref{fig:figure_spatial_exploration}):

-   A spatial exploration of your data starting from a Benn diagram
-   A spatial exploration of your data starting from the spatial
    projections (fig. \ref{fig:figure_spatial_exploration}J)
-   Summary data tables with orientation statistical tests (fig.
    \ref{fig:figure_spatial_exploration}I and K).

\begin{figure} [H]
\centering
  \includegraphics[width=17cm]{figure/spatialised_method.jpg}
  \caption{A. filter by sample, B. if you want to filter by code, uncheck 'no code distinction', C. choose the n nearest neighbors to perform the statistics, D. make sticks in spatial projections, E. upload a georeferenced orthophoto and check 'include' orthophoto to display it, F. new dataset at the spatial series scale, G. summary table at the sample scale.}
\label{fig:figure_spatialised_method}
\end{figure}

In this panel, you can perform a spatialised analysis of your data
starting from a Benn diagram or from projections. In both cases, you
need to select the data you want to explore more locally in a new
analysis. From the projections, you can select the points you want to
explore directly. In the Benn diagram, you need to select the lasso
selector (fig. \ref{fig:figure_spatial_exploration}D). Pay attention to
the number of points selected and check that the new sample contains
enough points to carry out the new analysis. If this is not the case,
you can lower the slider in the left sidebar panel to perform the
analysis with fewer neighbours (fig.
\ref{fig:figure_spatial_exploration}E). You can download the new samples
you have selected with spatial statistics from the corresponding
subpanels (fig. \ref{fig:figure_spatial_exploration}I and K). \bigbreak

## Tutorial

The 'Tutorial' panel contains a summary of the mathematics, statistics
and methods used in Fabryka and a short tutorial directly accessible
from the application. \bigbreak

\begin{figure} [H]
\centering
  \includegraphics[width=17cm]{figure/spatial_exploration.jpg}
  \caption{A. if you want to filter by sample, uncheck 'No sample distinction' B. if you want to filter by code, uncheck 'no code distinction', C. choose the n nearest neighbors to perform the statistics (first Benn diagram), E. choose the n nearest neighbors to perform the statistics on the sampled data (second Benn diagram), F. make sticks in spatial projections, G. you can upload a georeferenced orthophoto and check 'include' orthophoto to display it, H. a new spatial exploration of the fabric is performed, I. new dataset at the selected spatial series scale, I. a similar spatial exploration is possible from the projections, K. new dataset at the selected spatial series scale.}
\label{fig:figure_spatial_exploration}
\end{figure}

\pagebreak

## New models

The 'spatialised' method generates often more linear fabrics in
comparison with the 'classical' method. This can be explained by the
fact that objects that are close in space are more likely to resemble
each other than objects that are further in space. \bigbreak

Also, the structure of the point pattern is evocative of its fabric.
Thus, the dispersion of the points depends on the shape of the fabric.
Two extremes appear: the most isotropic samples are those with the most
dispersed point patterns (*i.e.* the standard deviations are high) while
the most linear samples are those with the least dispersed point
patterns. Planar samples highlight point patterns with intermediate
dispersion. The explanation of this phenomenon is intuitive: the more
similar the fabrics of objects are, the closer they are in the Benn
diagram and the more linear they are. To reflect this phenomenon, the
standard deviations of the Benn indices are calculated and aim to
describe the dispersion point patterns in the Benn diagrams.

Global planar or isotropic fabrics can sometimes mask local anisotropy.
Within certain samples, groups of objects form a dense cluster in the
point pattern. This is expressed either by a locally more linear fabric,
or by a planar fabric with a strong bimodal orientation.

For the reasons given above, we wish to create new models that
considered the particularities of the spatial method. A panel was
therefore set up to collect new data describing the fabric of
sedimentary processes in order to build a new "spatialised" reference
model (*cf.* the "classical" model proposed by Bertran and Lenoble 2002).
Some processes are already available. \bigbreak

## Contributors

The contributors panel aims at citing all contributors of this project
which is open to everyone interested in the fabric analysis methods or
to contribute in building a new spatial model by sending reference data.
If any request or contribution, please contact the author of this
article. \bigbreak

# Acknowledgements

The writing of the script was greatly facilitated by the work published
by Shannon McPherron (2018). It includes some functions published by the
author, with a few modifications. This work benefited from discussions
with Jean-Pierre Texier, Arnaud Lenoble, Pascal Bertran and Dominique
Todisco on issues related to site formation processes. It also benefited
from the help of Rémi Lemoy and François and Victor Baleux on
mathematical issues. A first version of the application was tested by
Emmanuel Discamps and Aurélien Royer. I thanks them for their advice and
support. \bigbreak

# Util links

-   use Fabryka $\beta$ 3.0: <https://marchaeologist.shinyapps.io/fabryka/> \smallbreak
-   source codes: <https://github.com/marchaeologist/fabryka/tree/main> \smallbreak
-   after the review process, create a doi (https://zenodo.org/login/?next=%2Faccount%2Fsettings%2Fgithub%2F) for each app' version. 
    \pagebreak

# Bibliography

The 'Bibliography' panel in the application contains the bibliography
cited in the application. This article contains the same bibliography
with some additional references. \bigbreak

### 2025-04-13 fabryka_pci/article.Rmd
last changes in the rmarkdown article

@@
 
-Marc THOMAS ^1\*^ \bigbreak \bigbreak
+\begin{figure} [H]
+\centering
+  \includegraphics[width=4cm]{figure/hex_Fabryka4.png}
+\label{}
+\end{figure}
+
+Marc THOMAS ^1\*^ (https://orcid.org/0000-0002-8160-1910) \smallbreak
 
@@
 E-mail: [marcthomas1\@hotmail.fr](mailto:marcthomas1@hotmail.fr){.email}
-\bigbreak \bigbreak
+\bigbreak
 
@@
 processes, spatial analyses, orientation and inclination \bigbreak
-\bigbreak
 
@@
 updated to Fabryka 1.0 after the review process of this article. The
-Fabryka application files, made in R [@rcoreteam2025] by using *shiny*
-package [@chang2024], and this article, written in rmarkdown
-[@allaire2014; @xie2018; @xie2020], are both available online in a
-public github repository
-(<https://github.com/marchaeologist/fabryka/tree/main>). When reviewing
-this article, you can either clone the repository, or push on the
-changes via github.\pagebreak
+Fabryka application files, made in R [@rcoreteam2025] by using *shiny* package [@chang2024], and this article, written in rmarkdown [@allaire2014; @xie2018; @xie2020], are both available online in a public github repository (<https://github.com/marchaeologist/fabryka/tree/main>).
+When reviewing this article you can fork and clone the fabryka project repository on github to make suggestions for modifications.\pagebreak
 
@@
 Fabric is made of the orientation and inclination of elongated
-sedimentary particles [@bertran1997; @bertran2002; @lenoble2004]. It
-is a statistic (*i.e.*, it needs as much measurements as possible) that
+sedimentary particles [@bertran1997; @bertran2002; @lenoble2004]. It is
+a statistic (*i.e.*, it needs as much measurements as possible) that
 enables the identification of sediment sources that make up the deposits
@@
 [@bertran1997; @bertran2002; @lenoble2004] and their impact on
-archaeological assemblages *e.g.* [@isaac1967; @schick1986;
-@thomas2019]. Objects abandoned by prehistoric people can be considered
-as sedimentary particles and are affected by the same processes
-[@schiffer1983]. As geological particles are often not elongated
-enough, the fabric analysis made of the sedimentary deposits depends
-mainly on the archaeological material (bones and lithics), which enables
-them to be inserted into the formation dynamics of the site and to
-obtain information on their representativeness for the reconstruction of
-past activities.
+archaeological assemblages [@isaac1967; @schick1986; @thomas2019].
+Objects abandoned by prehistoric people can be considered as sedimentary
+particles and are affected by the same processes [@schiffer1983]. As
+geological particles are often not elongated enough, the fabric analysis
+made of the sedimentary deposits depends mainly on the archaeological
+material (bones and lithics), which enables them to be inserted into the
+formation dynamics of the site and to obtain information on their
+representativeness for the reconstruction of past activities.
 
@@
 inclination. Since G. Isaac's pioneering work, there has been an
-abundant literature devoted to the analysis of fabrics *e.g.*
-[@benn1994; @lenoble2004; @mcpherron2018] allowing this method to be
-increasingly performed on archaeological sites. Several software allow
-data processing, some of which are not freely accessible, do not allow
+abundant literature devoted to the analysis of fabrics [@benn1994;
+@lenoble2004; @mcpherron2018] allowing this method to be increasingly
+performed on archaeological sites. Several software allow data
+processing, some of which are not freely accessible, do not allow
 spatial exploration of the data or do not offer interfaces to facilitate
-the task of non-programmer users *e.g* [@mcpherron2018].
+the task of non-programmer users [@mcpherron2018].
 
@@
 systematically measured. The direction (= bearing) of the lowest point
-is measured in the range [0; 360] with North as the reference (i.e.
+is measured in the range [0; 360] with North as the reference (*i.e.*
 direction = 0°). The inclination (= plunge, dip) is measured in the
@@
 Recently, speleologist have created a digital inclinometer (Leica
-DistoX310 modified to DistoX2; <https://paperless.bheeb.ch/>). The
-device stores and associates data with the order of measurement. When
-the device is tilted downwards, the direction of the lowest point is
-measured and, the recorded inclination is negative. When the device is
-tilted upwards, the direction of the highest point is recorded, and the
-inclination is positive. To obtain data that is suitable for statistical
-analysis, in the first case the absolute value of the inclination must
-be retained. In the second case (*i.e.* positive inclination), if the
-measured direction is in the interval [0; 180], 180° must be added to
-the direction, and if the measured direction is in the interval ]180;
-360], 180° must be subtracted from the direction.
+DistoX310 modified to DistoX2; <https://paperless.bheeb.ch/>) which can
+be used to measure fabrics [@discamps2024]. The device stores and
+associates data with the order of measurement. When the device is tilted
+downwards, the direction of the lowest point is measured and, the
+recorded inclination is negative. When the device is tilted upwards, the
+direction of the highest point is recorded, and the inclination is
+positive. To obtain data that is suitable for statistical analysis, in
+the first case the absolute value of the inclination must be retained.
+In the second case (*i.e.* positive inclination), if the measured
+direction is in the interval [0; 180], 180° must be added to the
+direction, and if the measured direction is in the interval ]180; 360],
+180° must be subtracted from the direction.
 
@@
 1.  The first considers the orientation only. The calculation of the
-    orientation rate "r" (or "L" when expressed as a percentage, Curray
-    formula [@curray1956] tests the hypothesis of a unimodal
-    distribution of orientations. If the "p" value returned by the
+    orientation rate 'r' (or 'L' when expressed as a percentage,
+    Curray's [-@curray1956] formula tests the hypothesis of a unimodal
+    distribution of orientations. If the 'p' value returned by the
     Rayleigh test is less than a threshold of 0.05, the null hypothesis
     of a random orientation from a uniform distribution is rejected. The
-    formula for calculating the orientation rate "r" where '$\alpha$' is
-    the orientation in radians and "n" is the sample size is as follows:
+    formula for calculating the orientation rate 'r' where '$\alpha$' is
+    the orientation in radians and 'n' is the sample size is as follows:
 
@@
 calculating the orientation rate L and the Rayleigh test, as proposed by
-Krumbein [@krumbein1939], tests the hypothesis of a bimodal
-distribution of orientations with a period of 90°. The hypothesis of a
-90° period bimodal distribution of a sample of orientations is retained
-when p \< 0.05 [@curray1956]. The sensitivity of the Rayleigh test to
-the sample size 'n' prevents samples from being compared with each other
-when 'n' varies [@curray1956; @bertran2002]. For example, when the
-orientation rate "L" is 25%, the Rayleigh test gives p-values below a
-critical threshold of 0.05 for series with more than 50 measurements
-(fig. \ref{fig:rayleigh}).
+Krumbein [-@krumbein1939], tests the hypothesis of a bimodal distribution
+of orientations with a period of 90°. The hypothesis of a 90° period
+bimodal distribution of a sample of orientations is retained when p \<
+0.05 [@curray1956]. The sensitivity of the Rayleigh test to the sample
+size 'n' prevents samples from being compared with each other when 'n'
+varies [@curray1956; @bertran2002]. For example, when the orientation
+rate "L" is 25%, the Rayleigh test gives p-values below a critical
+threshold of 0.05 for series with more than 50 measurements (fig. 2).
 
@@
 
-```{r rayleigh, echo = FALSE, warning = FALSE, message = FALSE, fig.cap="Evolution of the p-value returned by the Rayleigh test as a function of the orientation rate 'L' and the sample size 'n'", fig.height = 3.5, fig.pos="f"}
+```{r rayleigh, echo = FALSE, warning = FALSE, message = FALSE, fig.cap="Evolution of the p-value returned by the Rayleigh test as a function of the orientation rate 'L' and the sample size 'n' \\label{rayleigh}", fig.height = 3.5, fig.pos="f"}
 
@@
 'L' (formula (**3.**) modified R [@mcpherron2018] script) for each
-remains, and its n nearest neighbours in 3 dimensions (R McPherron 2018
-script, modified , formula (**4.**)). The analysis of the fabric as
-proposed by McPherron (2018) has the advantage of spatialising the
-measurements and thus avoiding the arbitrary sampling classically
-carried out by square or by décapage. It is possible to localize the
-statistical signal of fabrics by varying the number of measurements
-sampled to calculate the Benn indices. The choice of about 40 to 60
-measurements (each remain and its 40 to 60 nearest neighbours) to
-calculate the Benn indices allows local particularities to be observed,
-and above all, it allows the Rayleigh test to be carried out in its
-field of application (stability of the test between around 40 to 50
-measurements after [@bertran2002]).
+remains, and its n nearest neighbours in 3 dimensions (McPherron
+[-@mcpherron2018] R script, modified , formula (**4.**)). The analysis
+of the fabric as proposed by McPherron [-@mcpherron2018] has the
+advantage of spatialising the measurements and thus avoiding the
+arbitrary sampling classically carried out by square and décapage. It
+is possible to localize the statistical signal of fabrics by varying the
+number of measurements sampled to calculate the Benn indices. The choice
+of about 40 to 60 measurements (each remain and its 40 to 60 nearest
+neighbours) to calculate the Benn indices allows local particularities
+to be observed, and above all, it allows the Rayleigh test to be carried
+out in its field of application (stability of the test between around 40
+to 50 measurements after [-@bertran2002]).
 
@@
 in red, green and blue. The RGB colour code is reused to materialize the
-fabric of the remains on spatial projections of the material along the
-axes of the excavation grid. Projections (xy, xz and yz axes) are
-performed. \bigbreak
+fabric of the remains on spatial projections (plan and section views) of the material along the axes of the excavation grid. \bigbreak
 
@@
 
-## Upload your data
+## Upload data
+
+The simplest method to upload data in the application is to name the column names directly, as shown in the example files.
+If you miss a context column (*i.e.* 'id', 'code' or 'sample'), please create it by repeating a same character.
+By doing so, you do not need to select your variables in the main panel.
+The application correctly reads the data file directly.
 
@@
 
-Five cases (i.e. forms of data) are covered by the application (fig.
+Five cases (*i.e.* forms of data) are covered by the application (fig.
 \ref{fig:figure_upload_your_data}D):
@@
 
-\begin{figure} [H]
+\begin{figure} [h]
 \centering
-  \includegraphics[width=17cm]{figure/upload_your_data.jpg}
+  \includegraphics[width=16cm]{figure/upload_your_data.jpg}
   \caption{A.
@@
 \end{figure}
-
-If you miss a context column, please create it by repeating a same
-character.
 
@@
 
--   Benn diagrams with possible integration of Bertran and Lenoble
-    (2002) model (fig. \ref{fig:figure_classical_method}D)
--   Rose diagrams for bearing and plunge (fig.
-    \ref{fig:figure_classical_method}G)
+-   Benn diagrams with possible integration of Bertran and Lenoble (2002) model (fig. \ref{fig:figure_classical_method}D)
+-   Rose diagrams for bearing and plunge (fig. \ref{fig:figure_classical_method}G)
 -   Schmidt diagrams (fig. \ref{fig:figure_classical_method}H)
--   Summary data table with Benn indices, additional statistics, and
-    orientation statistical tests.
+-   Summary data table with Benn indices, additional statistics, and orientation statistical tests.
 
@@
 You can control the data you plot and the data appearing in the figures
-and the summary data table by changing the checkboxes filters (samples
-and code) and the slider (*i.e.* minimum number of measures) in the left
+and the summary data table by changing the checkboxes filters ('filter by sample'
+and 'filter by code') and the slider (*i.e.* minimum number of measures) in the left
 sidebar panel (fig. \ref{fig:figure_classical_method}A, B and C).
@@
 you can add to the Benn diagram are calculated using data mainly from
-[@bertran1997; @bertran2002; @bertran2006; @lenoble2004]. In the rose
-diagram, orientations are considered (and not directions) as the
-Rayleigh test is perform on orientations (see Curray 1956) in the
-summary table. You can download all the figures by selecting a file
-format (.pdf, .jpeg and .png) and pressing the 'download' button (fig.
-\ref{fig:figure_classical_method}E and F).
+[@bertran1997; @bertran2002; @bertran2006; @lenoble2004].
+You will find the reference model data (with corresponding litarature references) which is used in the application in the 'model reference data' subpanel.
+In the rose diagram, orientations are considered (and not directions) as the Rayleigh test is perform on orientations (Curray 1956) in the
+summary table. You can download all the figures by selecting a file format (.pdf, .jpeg and .png) and pressing the 'download' button (fig. \ref{fig:figure_classical_method}E and F).
 
-\begin{figure} [H]
+\begin{figure} [h]
 \centering
@@
 
--   Benn diagrams with Benn indices computed with nearest neighbours
-    points (McPherron 2018)
+-   Benn diagrams with Benn indices computed with nearest neighbours points (McPherron 2018)
 -   Spatial projections of the points according to their fabric shape
@@
     scale of each spatial series (each object and its n nearest
-    neighbours = one line) and one at the scale of the sample (fig.
+    neighbours fabric = one row) and one at the scale of the sample (fig.
     \ref{fig:figure_spatialised_method}F and G).
@@
 
-\begin{figure} [H]
+\begin{figure} [h]
 \centering
@@
 you can lower the slider in the left sidebar panel to perform the
-analysis with fewer neighbours (fig.
-\ref{fig:figure_spatial_exploration}E). You can download the new samples
-you have selected with spatial statistics from the corresponding
+analysis with fewer neighbours (fig. \ref{fig:figure_spatial_exploration}E). You can download the new samples you have selected with spatial statistics from the corresponding
 subpanels (fig. \ref{fig:figure_spatial_exploration}I and K). \bigbreak
@@
 
-\pagebreak
-
 ## New models
@@
 fact that objects that are close in space are more likely to resemble
-each other than objects that are further in space. \bigbreak
+each other than objects that are further in space.
 
@@
 sedimentary processes in order to build a new "spatialised" reference
-model (*cf.* the "classical" model proposed by Bertran and Lenoble 2002).
-Some processes are already available. \bigbreak
+model (*cf.* the "classical" model proposed by Bertran and Lenoble
+2002). Please, do not hesitate to send referential data to feed these new models ! \bigbreak
 
@@
 The contributors panel aims at citing all contributors of this project
-which is open to everyone interested in the fabric analysis methods or
+which is open to everyone interested in improving fabric analysis methods or
 to contribute in building a new spatial model by sending reference data.
@@
 article. \bigbreak
+
+# Util links
+
+-   use Fabryka $\beta$ 3.0:
+    <https://marchaeologist.shinyapps.io/fabryka/> \smallbreak
+-   source codes: <https://github.com/marchaeologist/fabryka/tree/main>
+    \smallbreak
+-   github reprository doi of Fabryka $\beta$ 3.0: \smallbreak
 
@@
 with Jean-Pierre Texier, Arnaud Lenoble, Pascal Bertran and Dominique
-Todisco on issues related to site formation processes. It also benefited
-from the help of Rémi Lemoy and François and Victor Baleux on
-mathematical issues. A first version of the application was tested by
-Emmanuel Discamps and Aurélien Royer. I thanks them for their advice and
+Todisco on issues related to site formation processes.
+The reference data used to build the models of the classical method were collected by P.
+Bertran throughout his career.
+P.
+Bertran provided us with all these data in order to feed the application.
+The developpement of the application also benefited from the help of Rémi Lemoy and François and Victor Baleux on mathematical issues.
+A first version of the application was tested by
+Emmanuel Discamps and Aurélien Royer. I would like to thank all these reaseachers for their advice and
 support. \bigbreak
-
-# Util links
-
--   use Fabryka $\beta$ 3.0: <https://marchaeologist.shinyapps.io/fabryka/> \smallbreak
--   source codes: <https://github.com/marchaeologist/fabryka/tree/main> \smallbreak
--   after the review process, create a doi (https://zenodo.org/login/?next=%2Faccount%2Fsettings%2Fgithub%2F) for each app' version. 
-    \pagebreak
 

### 2025-05-05 fabryka_pci/article.Rmd
Suggestion for various typos and biblio improvements

@@
 updated to Fabryka 1.0 after the review process of this article. The
-Fabryka application files, made in R [@rcoreteam2025] by using *shiny* package [@chang2024], and this article, written in rmarkdown [@allaire2014; @xie2018; @xie2020], are both available online in a public github repository (<https://github.com/marchaeologist/fabryka/tree/main>).
+Fabryka application files, made in R [@rcoreteam2025] by using the R package *shiny* [@chang2024], and this article, written in rmarkdown [@allaire2014; @xie2018; @xie2020], are both available online in a public github repository (<https://github.com/marchaeologist/fabryka/tree/main>).
 When reviewing this article you can fork and clone the fabryka project repository on github to make suggestions for modifications.\pagebreak
@@
 archaeological assemblages [@isaac1967; @schick1986; @thomas2019].
-Objects abandoned by prehistoric people can be considered as sedimentary
+Objects abandoned by prehistoric people can be treated as sedimentary
 particles and are affected by the same processes [@schiffer1983]. As
@@
 
-As proposed by McPherron (2018), a spatial investigation of the fabric
+As proposed by @mcpherron2018, a spatial investigation of the fabric
 is proposed in a dedicated panel. Benn indices (1994) calculated for
@@
 Le Moustier Lower shelter (Dordogne, France). The example files directly
-contain the good column names. You can use them directly in the
+contain the right column names. You can use them directly in the
 application. Whatever the names of the columns in your file, you can
@@
 
--   Benn diagrams with possible integration of Bertran and Lenoble (2002) model (fig. \ref{fig:figure_classical_method}D)
+-   Benn diagrams with possible integration of @bertran2002 model (fig. \ref{fig:figure_classical_method}D)
 -   Rose diagrams for bearing and plunge (fig. \ref{fig:figure_classical_method}G)
@@
 -   'L.double' and 'R.p.double' are the same statistics that above for
-    double angles (Krumbein 1939).
+    double angles [@krumbein1939].
 
@@
 preferential orientation and are much more 'linear' in the Benn diagram,
-see [@thomas2019] and [@texier2020]). The natural processes models that
+see @thomas2019 and @texier2020). The natural processes models that
 you can add to the Benn diagram are calculated using data mainly from
@@
 You will find the reference model data (with corresponding litarature references) which is used in the application in the 'model reference data' subpanel.
-In the rose diagram, orientations are considered (and not directions) as the Rayleigh test is perform on orientations (Curray 1956) in the
+In the rose diagram, orientations are considered (and not directions) as the Rayleigh test is perform on orientations [@curray1956] in the
 summary table. You can download all the figures by selecting a file format (.pdf, .jpeg and .png) and pressing the 'download' button (fig. \ref{fig:figure_classical_method}E and F).
@@
 manage the minimum sample size to perform analyses, D.
-buld your model thanks to Bertran and Lenoble (2002) data, E.
+buld your model thanks to @bertran2002 data, E.
 and F.
@@
 
--   Benn diagrams with Benn indices computed with nearest neighbours points (McPherron 2018)
+-   Benn diagrams with Benn indices computed with nearest neighbours points [@mcpherron2018]
 -   Spatial projections of the points according to their fabric shape
@@
     the Benn indices.
--   'IS', 'PL' and 'EL' are calculated thanks to the eigenvalues (Benn
-    1994). These are the Benn indices, ratios standing respectively for,
+-   'IS', 'PL' and 'EL' are calculated thanks to the eigenvalues [@benn1994]. These are the Benn indices, ratios standing respectively for,
     'isotropy' with IS = E3/E1, 'elongation' with EL = 1-(E2/E1) and
@@
 -   'L.double' and 'R.p.double' are the same statistics that above for
-    double angles (see Krumbein 1939).
+    double angles [@krumbein1939].
 
@@
 
-# Util links
+# Useful links
 
@@
     <https://marchaeologist.shinyapps.io/fabryka/> \smallbreak
--   source codes: <https://github.com/marchaeologist/fabryka/tree/main>
+-   source code: <https://github.com/marchaeologist/fabryka/tree/main>
     \smallbreak
--   github reprository doi of Fabryka $\beta$ 3.0: \smallbreak
+-   Zenodo repository of Fabryka $\beta$ 3.0: <https://doi.org/10.5281/zenodo.15207419> \smallbreak
 

### 2025-05-05 fabryka_pci/article.Rmd
Typos

@@
 To distinguish different statistical populations, a spatial method
-proposed by McPherron (2018) is used. Thanks to this methods it is
+proposed by McPherron (2018) is used. Thanks to this method it is
 possible to calculate the Benn indices and to test the orientation rate
@@
 out in its field of application (stability of the test between around 40
-to 50 measurements after [-@bertran2002]).
+to 50 measurements after @bertran2002).
 

### 2025-05-05 fabryka_pci/article.Rmd
Reword sentence

@@
 possible to calculate the Benn indices and to test the orientation rate
-'L' (formula (**3.**) modified R [@mcpherron2018] script) for each
+'L' (formula (**3.**) modified from @mcpherron2018 R script) for each
 remains, and its n nearest neighbours in 3 dimensions (McPherron
-[-@mcpherron2018] R script, modified , formula (**4.**)). The analysis
+[-@mcpherron2018] R script, modified, formula (**4.**)). The analysis
 of the fabric as proposed by McPherron [-@mcpherron2018] has the

### 2025-07-14 DEPENDENCIES.Rmd
correction of fabryka article and shiny application after first round of review in PCI archaeology

---
title: "fabryka"
Version: 1.1
output: github_document
---

```{=html}
<style>
body {
text-align: justify}
</style>
```

# [<img src="https://raw.githubusercontent.com/marchaeologist/fabryka/main/inst/www/hex_Fabryka4.png" height="250" align="center"/>](https://github.com/marchaeologist/fabryka)

[![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.15236695.svg)](https://doi.org/10.5281/zenodo.15236695)

# Package dependencies

Find below a list of all packages used in fabryka.

```{r library_fabryka}

fabryka_packages=c("shiny", "dplyr","ggplot2","plotly", "ggsci", "ggtern", "circular", "CircStats", "Ternary", 
                   "RStoolbox", "shinyjs", "rmarkdown", "shinythemes",  "ggalt", "DT", "raster")
print(fabryka_packages)

```

Find in the code chunk below a way to install all the packages used in fabryka.

```{r install_library, eval = FALSE, tidy=FALSE, width=120}

fabryka_packages=c("shiny", "dplyr","ggplot2","plotly", "ggsci", "ggtern", "circular", "CircStats", "Ternary", 
                   "RStoolbox", "shinyjs", "rmarkdown", "shinythemes",  "ggalt", "DT", "raster")
is_installed=sapply(fabryka_packages,require,character.only=T)
sapply(fabryka_packages[!is_installed],install.packages)

```

Find below a list of all packages dependencies used in fabryka.

```{r dependencies}
options(repos = c(CRAN = "https://cloud.r-project.org"))
dependencies <- tools::package_dependencies(fabryka_packages, db = available.packages())
print(dependencies)
```

Please note that there are currently conflicts between the functions of several packages. You will find a list below. This problem will be addressed more carefully in future versions of the application. For now, all functions have been prefixed by their namespace.

```{r conflicts}
library(conflicted)
conflicted::conflict_scout(fabryka_packages)
```

Below find the list of imported packaged by using the usethis::use_package() command.

```{r imports, eval=FALSE}

usethis::use_package("shiny")
usethis::use_package("dplyr")
usethis::use_package("ggplot2")
usethis::use_package("plotly")
usethis::use_package("ggsci")
usethis::use_package("ggtern")
usethis::use_package("circular")
usethis::use_package("CircStats")
usethis::use_package("Ternary")
usethis::use_package("RStoolbox")
usethis::use_package("shinyjs")
usethis::use_package("rmarkdown")
usethis::use_package("shinythemes")
usethis::use_package("ggalt")
usethis::use_package("DT")
usethis::use_package("raster")

```

### 2025-07-14 README.Rmd
correction of fabryka article and shiny application after first round of review in PCI archaeology

---
title: "fabryka"
Version: 1.1
Author: Marc THOMAS [aut, cre] 
output:
  github_document
---

```{=html}
<style>
body {
text-align: justify}
</style>
```

# [<img src="https://raw.githubusercontent.com/marchaeologist/fabryka/main/inst/www/hex_Fabryka4.png" height="250" align="center"/>](https://github.com/marchaeologist/fabryka)

[![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.15236695.svg)](https://doi.org/10.5281/zenodo.15236695)

## README

fabryka is an open access R *shiny* user-friendly web application dedicated to the fabric analysis of archaeological or geological assemblage. fabryka uses classical analytical procedures (Benn, rose, Schmidt and Woodcock diagrams and statistical tests on orientations) for analysing the fabrics of objects, an spatial method and new ways of spatial and statistical exploration. A descriptive article is available at <https://zenodo.org/records/15236695>.

## USE fabryka

The web application is available online at <https://marchaeologist.shinyapps.io/fabryka/>.

You can also use fabryka in RStudio. To do so :

* Install [R](https://www.r-project.org) and [Rstudio Desktop](https://posit.co/download/rstudio-desktop/).

* clone the fabryka repository from GitHub: <https://github.com/marchaeologist/fabryka> or download the code in "Code" and "Download ZIP" at the same address. This will give you the latest version of the source code. The application can be launched by opening the app.R file in the main folder and by running the following commands:

```{r, eval=FALSE}

pkgload::load_all(".")
fabryka()

```

This can be useful if you want to use fabryka in the field without an internet access.

## CONTACT

If you have problems, questions or suggestions, please contact me at marcthomas1@hotmail.fr or via the fabryka repository issues page at <https://github.com/marchaeologist/fabryka/issues>.
All contributors to the project are listed in the 'Contributors' panel in the application.

## SOURCE CODE

The source code is available on github: <https://github.com/marchaeologist/fabryka>. You will also find previous versions of the application archived there.

## TUTORIAL AND EXAMPLE DATASET

A tutorial is included in the application (*i.e.* 'Tutorial" panel) and can also be found in a dedicated article: <https://zenodo.org/records/15236695>.
'fabryka' includes an example file you can download from the application ('Upload data' subpanel).

## NOTICE

This project is licensed under the [MIT](https://choosealicense.com/licenses/mit/) License — Copyright (c) 2025 fabryka.

## CITATION

THOMAS, M. (2025). Fabryka: A web application for easily analysing & exploring the fabric of archaeological assemblages. In PCI archaeology (Version 1). Zenodo. https://doi.org/10.5281/zenodo.15236695

### 2025-07-14 fabryka_pci/article.Rmd
correction of fabryka article and shiny application after first round of review in PCI archaeology

--- 2025-05-05 fabryka_pci/article.Rmd
@@
 ---
-title: '**Fabryka**'
+title: '**fabryka**'
 subtitle: A web application for easily analysing & exploring the fabric of archaeological assemblages
@@
 header-includes:
+- \usepackage{amsmath}
+- \usepackage{lineno}
+- \linenumbers
 - \usepackage{fancyhdr}
 - \pagestyle{fancy}
-- \fancyhead[L]{Fabryka}
+- \fancyhead[L]{fabryka}
 - \fancyhead[R]{R web application for fabric analysis}
@@
 
+\renewcommand{\[}{\begin{equation}}
+\renewcommand{\]}{\end{equation}}
 ```{=html}
@@
 
-Marc THOMAS ^1\*^ (https://orcid.org/0000-0002-8160-1910) \smallbreak
+Marc THOMAS ^1\*^ (<https://orcid.org/0000-0002-8160-1910>) \smallbreak
 
@@
 **Key words** : R shiny web application, fabric, site formation
-processes, spatial analyses, orientation and inclination \bigbreak
+processes, spatial analysis, orientation and dip \bigbreak
 
@@
 
-This article presents Fabryka, an open access R shiny web application
-dedicated to the fabric analysis of archaeological or geological
-assemblages. The fabric is a statistic made of the orientation and
-inclination of elongated sedimentary particles and archaeological
-objects. It provides information about the source of sediments, the
-nature of the geological processes involved in the formation of the
-deposits and their impact on the preservation of the archaeological
-assemblages. These statistical analyses are not easily accessible to
-non-specialists and are particularly time-consuming, sometimes requiring
-the use of several software, some of which are chargeable. Fabryka uses
-classical analytical procedures (Benn, rose and Schmidt diagrams and
-statistical tests on orientations) for analysing the fabrics of objects
-and includes a spatial method. New statistical methods and new ways of
-exploration are proposed in the application and this present article.
-Also, a call for contributions is proposed to construct a new reference
-model with spatial information. \smallbreak
+This article presents fabryka 1.1, an open access R shiny web
+application dedicated to the fabric analysis of archaeological or
+geological assemblages. The fabric is a statistic made of the
+orientation and dip of elongated particles and archaeological objects.
+It provides information about the nature of the geological processes
+involved in the formation of the deposits and their impact on the
+preservation of the archaeological assemblages. These statistical
+analysis are not easily accessible to non-specialists and are
+particularly time-consuming, sometimes requiring the use of several
+software, some of which are chargeable. fabryka uses classical
+analytical procedures (Benn, rose, Schmidt and Woodcock diagrams and statistical
+tests on orientations) for analysing the fabrics of objects and includes
+a spatial method. New statistical methods and new ways of exploration
+are proposed in the application and this present article.\smallbreak
 
-You can use Fabryka $\beta$ 3.0 by following this link
-<https://marchaeologist.shinyapps.io/fabryka/>. The version will be
-updated to Fabryka 1.0 after the review process of this article. The
-Fabryka application files, made in R [@rcoreteam2025] by using the R package *shiny* [@chang2024], and this article, written in rmarkdown [@allaire2014; @xie2018; @xie2020], are both available online in a public github repository (<https://github.com/marchaeologist/fabryka/tree/main>).
-When reviewing this article you can fork and clone the fabryka project repository on github to make suggestions for modifications.\pagebreak
+You can use fabryka by following this link
+<https://marchaeologist.shinyapps.io/fabryka/> or by launching the
+application in Rstudio thanks to its script. The fabryka application
+files, made in R [@rcoreteam2025] by using the R package *shiny*
+[@chang2024], and this article, written in rmarkdown [@allaire2014;
+@xie2018; @xie2020], are both available online in a public github
+repository (<https://github.com/marchaeologist/fabryka/tree/main>).
+\pagebreak
 
@@
 
-Fabric is made of the orientation and inclination of elongated
-sedimentary particles [@bertran1997; @bertran2002; @lenoble2004]. It is
-a statistic (*i.e.*, it needs as much measurements as possible) that
-enables the identification of sediment sources that make up the deposits
-[@texier2020], sedimentary processes involved in their formation
-[@bertran1997; @bertran2002; @lenoble2004] and their impact on
+Fabric is made of the orientation and dip of elongated objects
+[@bertran1997; @bertran2002; @lenoble2004]. It is a statistic (*i.e.*,
+it needs as much measurements as possible) that enables the
+identification of sedimentary processes involved in the formation of the
+deposits [@bertran1997; @bertran2002; @lenoble2004] and their impact on
 archaeological assemblages [@isaac1967; @schick1986; @thomas2019].
-Objects abandoned by prehistoric people can be treated as sedimentary
+Objects abandoned by prehistoric people can be considered as sedimentary
 particles and are affected by the same processes [@schiffer1983]. As
-geological particles are often not elongated enough, the fabric analysis
-made of the sedimentary deposits depends mainly on the archaeological
-material (bones and lithics), which enables them to be inserted into the
-formation dynamics of the site and to obtain information on their
-representativeness for the reconstruction of past activities.
+geological particles are often not elongated enough, analysis of the
+fabric of sedimentary deposits depends mainly on archaeological material
+(bones and lithics). This makes it possible to identify the sedimentary
+processes to which they were subjected and to discuss their degree of
+spatial disorganization.
 
@@
 revealed a reorientation that was mainly perpendicular and, to a lesser
-extent, parallel to the direction of flow, with a slight upstream
-inclination. Since G. Isaac's pioneering work, there has been an
-abundant literature devoted to the analysis of fabrics [@benn1994;
-@lenoble2004; @mcpherron2018] allowing this method to be increasingly
-performed on archaeological sites. Several software allow data
-processing, some of which are not freely accessible, do not allow
-spatial exploration of the data or do not offer interfaces to facilitate
-the task of non-programmer users [@mcpherron2018].
+extent, parallel to the orientation of flow, with a slight upstream dip.
+Since G. Isaac's pioneering work, there has been an abundant literature
+devoted to the analysis of fabrics [@benn1994; @lenoble2004;
+@mcpherron2018] allowing this method to be increasingly performed on
+archaeological sites.
+
+Numerical methods have been developed to measure the orientation and dip
+of objects in the field [@mcpherron2005artifact; @discamps2024] or from
+field archives such as plans and photos [@de2013application;
+@sanchez2016assessment].
+
+Two main families of statistical method are commonly used to
+characterize the fabric of a given assemblage [@bertran2002;
+@mcpherron2018; @de2021new].
+
+The first methods focus on the orientation
+of the measurements only and aim to test the null hypothesis of a random
+orientation from a uniform distribution through several statistical
+tests [@krumbein1939; @curray1956; @rao1967large].
+
+The second statistical method take into account object orientation and
+dip. Three normalized eigenvalues (E1, E2 and E3) are calculated from
+the cosine formed by the axis of the artifact with the three axes of
+space [@watson1966]. @woodcock1977 and @benn1994 indices are used to
+describe the fabric shape. Several authors *e.g.*, [@bertran1997;
+@bertran2002; @lenoble2004] built models to compare the fabric of
+archaeological assemblages to natural sedimentary processes and human
+processes (*i.e.*, fabric of knapping spots). The null hypothesis of an
+absence of perturbation is a fabric shape compatible with that of
+experimental knapping spots. These methods are described in the
+Mathematics section of this article.
+
+Several software allow data processing, some of which are not freely
+accessible, do not allow spatial exploration of the data or do not offer
+interfaces to facilitate the task of non-programmer users
+[@mcpherron2018].
 
 The following sections explain all the selected approaches and our
-proposals for processing the fabrics thanks to fabryka $\beta$ 3.0, from
-taking measurements in the field, through the statistical methods and
-graphical modes of representation used in the application (fig.
+proposals for processing the fabrics thanks to fabryka, from taking
+measurements in the field, through the statistical methods and graphical
+modes of representation used in the application (fig.
 \ref{fig:figure_presentation}). \bigbreak
@@
   \includegraphics[width=16cm]{figure/presentation_applications.png}
-  \caption{Fabryka creation scheme and organisation of the article}
+  \caption{fabryka creation scheme and organisation of the article}
 \label{fig:figure_presentation}
@@
 
-During excavation, the fabric of elongated remains is generally
-systematically measured. The direction (= bearing) of the lowest point
-is measured in the range [0; 360] with North as the reference (*i.e.*
-direction = 0°). The inclination (= plunge, dip) is measured in the
-range [0; 90] with horizontal as a reference. The direction [0; 360] can
-be converted into an orientation measured in the range [0; 180] by
-subtracting 180° from the measurements in the range ]180; 360].
+During excavation, the orientation and dip of elongated remains (*i.e.* a > 2 * b and with a > 2 cm) is
+measured on a representative set of artefacts or on the whole
+assemblage. The orientation (= bearing) of the lowest point is measured
+in the range [0; 360] with North as the reference (*i.e.* orientation =
+0°). The dip (= plunge) is measured in the range [0; 90] with horizontal
+as a reference. The orientation [0; 360] can be converted into axial
+data measured in the range [0; 180] by subtracting 180° from the
+measurements in the range ]180; 360].
 
-Traditionally, the direction of the objects lowest point is measured
-using a compass and the inclination is measured using an inclinometer.
-The aim is to measure the direction and inclination of the physical axis
-of elongation of objects ignoring its shape. To do this, the measuring
-device must be placed parallel to the general axis of elongation of the
-piece without touching it. Measuring the fabric from its imprint or
-visible surface is biased by its shape, making impossible to assess its
-true inclination. Since the theodolite was introduced to excavations in
-the 1990s, the fabrics is sometimes measured by recording the
-coordinates of the ends of objects. In the latter case, the rules of
-trigonometry are applied to calculate directions and inclinations.
+Traditionally, the orientation of the objects lowest point is measured
+using a compass and the dip is measured using an inclinometer. The aim
+is to measure the orientation and dip of the physical axis of elongation
+of objects ignoring its shape. To do this, the measuring device must be
+placed parallel to the general axis of elongation of the object.
+Measuring the fabric from its imprint or visible surface is biased by
+its shape, making impossible to assess its true dip. Since the
+theodolite was introduced to excavations in the 1990s, the fabrics is
+sometimes measured by recording the coordinates of the ends of objects [@mcpherron2005artifact].
+In the latter case, the rules of trigonometry are applied to convert the
+pair of coordinates into orientations and dips.
 
@@
 associates data with the order of measurement. When the device is tilted
-downwards, the direction of the lowest point is measured and, the
-recorded inclination is negative. When the device is tilted upwards, the
-direction of the highest point is recorded, and the inclination is
-positive. To obtain data that is suitable for statistical analysis, in
-the first case the absolute value of the inclination must be retained.
-In the second case (*i.e.* positive inclination), if the measured
-direction is in the interval [0; 180], 180° must be added to the
-direction, and if the measured direction is in the interval ]180; 360],
-180° must be subtracted from the direction.
+downwards, the orientation of the lowest point is measured and, the
+recorded dip is negative. When the device is tilted upwards, the
+orientation of the highest point is recorded, and the dip is positive.
+To obtain data that is suitable for statistical analysis, in the first
+case the absolute value of the dip must be retained. In the second case
+(*i.e.* positive dip), if the measured orientation is in the interval
+[0; 180], 180° must be added to the orientation, and if the measured
+orientation is in the interval ]180; 360], 180° must be subtracted from
+the orientation.
 
@@
 reasons given above, the use of a theodolite to measure fabrics is not
-recommended as it results in inaccurate inclination measurements. As
-well, measuring orientations and inclinations in sections leads to
-sampling biases.
+recommended as it results in inaccurate dip measurements. As well,
+measuring orientations and dip in sections leads to sampling biases.
 
-A sample corresponds to all the direction and inclination measurements
-taken within a Lithostratigraphic Unit (UL). UL seems to be the most
-coherent entity which needs to be explored because they are defined
-according to sedimentary criterion, which a priori guarantee a certain
-coherence in terms of the formation dynamics we are seeking to approach.
-In Fabryka, analyses are constrained within samples unless the user
-decides otherwise in the spatial exploration panel. \bigbreak
+A sample corresponds to all the orientation and dip measurements taken
+within a Lithostratigraphic Unit (LU) or a pedostratigraphic unit. These
+units seem to be the most coherent entity which needs to be explored
+because they are defined according to geological criterion, which a
+priori guarantee a certain coherence in terms of the formation dynamics
+we are seeking to approach. In fabryka, analysis are constrained within
+samples unless the user decides otherwise in the spatial exploration
+panel. \bigbreak
 
@@
 
-Two main statistical methods are used:
+Two main statistical family methods are used:
 
 1.  The first considers the orientation only. The calculation of the
-    orientation rate 'r' (or 'L' when expressed as a percentage,
-    Curray's [-@curray1956] formula tests the hypothesis of a unimodal
-    distribution of orientations. If the 'p' value returned by the
+    Vector Magnitude $L$ (expressed as a percentage) by Curray's
+    [-@curray1956] formula tests the hypothesis of a unimodal
+    distribution of orientations. If the $p$ value returned by the
     Rayleigh test is less than a threshold of 0.05, the null hypothesis
     of a random orientation from a uniform distribution is rejected. The
-    formula for calculating the orientation rate 'r' where '$\alpha$' is
-    the orientation in radians and 'n' is the sample size is as follows:
+    formula for calculating the Vector Magnitude $L$ is as follows:
 
-Curray's formula to calculate 'L' is (**1.**)
+Curray's formula to calculate $L$ is:
 
-$$L = \frac{r*100}{n}$$ with (**2.**)
+$$L = \frac{r*100}{n}$$ with:
 
-$$r = \sqrt{\sum_{i=\alpha}^n sin{(2\alpha)} + \sum_{i=\alpha}^n cos{(2\alpha)}}$$
-\smallbreak where 'n' is the number of measures and '$\alpha$', is the
-orientation (0 to 180°) \smallbreak
+$$r = \sqrt{\sum_{i=\alpha}^n \sin{(2\alpha)} + \sum_{i=\alpha}^n \cos{(2\alpha)}}$$
+\smallbreak where $n$ is the number of measures and $\alpha$, is the
+axial data (0 to 180°) \smallbreak
 
-The Rayleigh test is performed thanks to the formula (**3.**)
+The Rayleigh test is performed thanks to the formula:
 
@@
 
-The method consisting of doubling the value of the orientations before
-calculating the orientation rate L and the Rayleigh test, as proposed by
-Krumbein [-@krumbein1939], tests the hypothesis of a bimodal distribution
-of orientations with a period of 90°. The hypothesis of a 90° period
-bimodal distribution of a sample of orientations is retained when p \<
-0.05 [@curray1956]. The sensitivity of the Rayleigh test to the sample
-size 'n' prevents samples from being compared with each other when 'n'
-varies [@curray1956; @bertran2002]. For example, when the orientation
-rate "L" is 25%, the Rayleigh test gives p-values below a critical
+The method consisting of doubling the value of the axial angles before
+calculating the Vector Magnitude $L$ and the Rayleigh test, as proposed by
+Krumbein [-@krumbein1939], tests the hypothesis of a bimodal
+distribution with a period of 90°. The hypothesis of a 90° period
+bimodal distribution of a sample of orientations is retained when $p$
+\<0.05 [@curray1956]. The sensitivity of the Rayleigh test to the sample
+size $n$ prevents samples from being compared with each other when $n$
+varies [@curray1956; @bertran2002]. For example, when the Vector
+Magnitude $L$ is 25%, the Rayleigh test gives $p-values$ below a critical
 threshold of 0.05 for series with more than 50 measurements (fig. 2).
 
-2.  The second considers the direction and inclination of the objects.
-    The normalised eigenvalues E1, E2 and E3 are calculated from the
+The Rao [-@rao1967large] spacing test assesses the uniformity of
+circular data. It determines whether the data show significant
+directionality (*i.e.* the null hypothesis is a uniform distribution).
+It is widely used to detect multimodal directionality.
+
+2.  The second considers the orientation and dip of the objects. The
+    normalised eigenvalues $E1$, $E2$ and $E3$ are calculated from the
     cosine formed by the axis of the objects with the three axes of
-    space [@watson1966]. The Benn indices [@benn1994] describe the
-    fabric shape of a sample. The elongation index EL is equal to
-    1-(E2/E1) and the isotropy index IS is E3/E1. When EL is high, the
-    axes of the objects are grouped around a direction. When IS is high,
-    the fabric is said to be isotropic meaning the axes of the objects
-    show no order. Finally, when the value 1-EL-IS is high, the fabric
-    is said to be planar, with the axes of the objects grouped around a
-    plane. \bigskip
+    space [@watson1966].
 
-```{r rayleigh, echo = FALSE, warning = FALSE, message = FALSE, fig.cap="Evolution of the p-value returned by the Rayleigh test as a function of the orientation rate 'L' and the sample size 'n' \\label{rayleigh}", fig.height = 3.5, fig.pos="f"}
+The fabric shape can be described by calculating the ratios $r1$ = ln(E1/E2), $r2$ = ln(E2/E3) and $K$ = r1/r2 [@woodcock1977].
+When 0 \< $K$ \< 1, the fabric is planar.
+When 1 \< $K$ \< $+\infty$, the fabric is linear.
+The parameter $C$ = ln(E1/E3) is a measure of the strength of the preferred orientation [@woodcock1977].
+These indices can be plotted into a Woodcock diagram.
+
+The Benn indices [@benn1994] describe the fabric shape of a sample and can be plotted in a ternary diagram (*i.e.* Benn diagram).
+The elongation index EL is equal to $1-(E2/E1)$ and the isotropy index $IS$ is $E3/E1$.
+When $EL$ is high, the axes of the objects are grouped around an orientation.
+When $IS$ is high, the fabric is said to be isotropic meaning the axes of the objects are uniformly distributed in all directions.
+Finally, when the value $1-EL-IS$ is high, the fabric is said to be planar, with the axes of the objects grouped around a plane.
+\bigskip
+
+```{r rayleigh, echo = FALSE, warning = FALSE, message = FALSE, fig.cap="Evolution of the p-value returned by the Rayleigh test as a function of the Vector Magnitude 'L' and the sample size 'n' \\label{rayleigh}", fig.height = 3.5, fig.pos="f"}
 
@@
     ylab("p-value") + 
-    scale_color_manual(values=c("#0073C2", "#EFC000", "#868686", "#3B3B3B", "red")) +         scale_x_continuous(breaks = c(0, 10, 20, 30, 40, 50, 60)) +
-scale_y_continuous(breaks = c(0, 0.10, 0.20, 0.30, 0.40, 0.50, 0.60, 0.70, 0.80, 0.90, 1)) +
+    scale_color_manual(values=c("#0073C2", "#EFC000", "#868686", "#3B3B3B", "red")) +         
+    scale_x_continuous(breaks = c(0, 10, 20, 30, 40, 50, 60), expand = c(0, 0)) +
+    scale_y_continuous(breaks = c(0, 0.10, 0.20, 0.30, 0.40, 0.50, 0.60, 0.70, 0.80, 0.90, 1), expand = c(0, 0)) +
     theme_minimal() +
@@
 To distinguish different statistical populations, a spatial method
-proposed by McPherron (2018) is used. Thanks to this method it is
-possible to calculate the Benn indices and to test the orientation rate
-'L' (formula (**3.**) modified R [@mcpherron2018] script) for each
-remains, and its n nearest neighbours in 3 dimensions (McPherron
-[-@mcpherron2018] R script, modified , formula (**4.**)). The analysis
-of the fabric as proposed by McPherron [-@mcpherron2018] has the
-advantage of spatialising the measurements and thus avoiding the
-arbitrary sampling classically carried out by square and décapage. It
-is possible to localize the statistical signal of fabrics by varying the
-number of measurements sampled to calculate the Benn indices. The choice
-of about 40 to 60 measurements (each remain and its 40 to 60 nearest
-neighbours) to calculate the Benn indices allows local particularities
-to be observed, and above all, it allows the Rayleigh test to be carried
-out in its field of application (stability of the test between around 40
-to 50 measurements after @bertran2002).
+proposed by McPherron [-@mcpherron2018] is used. Thanks to this method
+it is possible to calculate the Benn indices and the Vector Magnitude
+$L$ (formula (**3.**) modified from McPherron [-@mcpherron2018] R
+script) for each remains, and its $n$ nearest neighbors in 3 dimensions
+(McPherron [-@mcpherron2018] R script, modified, formula (**4.**)). The
+analysis of the fabric as proposed by McPherron [-@mcpherron2018] has
+the advantage of spatialising the measurements. It is possible to
+localize the statistical signal of fabrics by varying the number of
+measurements sampled to calculate the Benn indices. The choice of about
+40 to 60 measurements (each remain and its 40 to 60 nearest neighbors)
+to calculate the Benn indices allows local particularities to be
+observed, and above all, it allows the Rayleigh test to be carried out
+in its field of application (stability of the test between around 40 to
+50 measurements after @bertran2002).
 
 By avoiding the instability of the Rayleigh test, the rejection rate
-'Tr' of the null hypothesis provides a more reliable description of the
+$Tr$ of the null hypothesis provides a more reliable description of the
 distribution of orientations in a sample. In addition, it allows
@@
 
-The rejection ratio 'Tr' is performed thanks to the formula (**4.**)
+The rejection ratio $Tr$ is performed thanks to the formula:
 
@@
 Several methods of representation are used. Rose diagrams summaries the
-orientation and inclination measurements for each sample. Benn indices
-are projected in a Benn diagram and measures (directions and plunge) are
+orientation and dip measurements for each sample. Benn indices are
+projected in a Benn diagram and measures (orientation and dip) are
 projected into Schmidt diagrams. These figures are available for both
@@
 
-As proposed by @mcpherron2018, a spatial investigation of the fabric
-is proposed in a dedicated panel. Benn indices (1994) calculated for
-each spatial series of remains are projected within a Benn diagram in
-which each pole (isotropic, linear, and planar) is coloured respectively
-in red, green and blue. The RGB colour code is reused to materialize the
-fabric of the remains on spatial projections (plan and section views) of the material along the axes of the excavation grid. \bigbreak
+As submited by @mcpherron2018, a spatial investigation of the fabric is
+proposed in a dedicated panel. @benn1994 indices calculated for each
+spatial series of remains are projected within a Benn diagram in which
+each pole (isotropic, linear, and planar) is colored respectively in
+red, green and blue. The RGB color code is reused to materialize the
+fabric of the remains on spatial projections (plan and section views) of
+the material along the axes of the excavation grid. \bigbreak
 
@@
 
-The next sections provide a description and a tutorial of Fabryka.
+The next sections provide a description and a tutorial of fabryka. All
+the figures in this section were produced using the example data set
+provided in the application (*i.e.* "Case 3: Angles from DistoX2 with
+coordinates).”
 
@@
 
-The simplest method to upload data in the application is to name the column names directly, as shown in the example files.
-If you miss a context column (*i.e.* 'id', 'code' or 'sample'), please create it by repeating a same character.
-By doing so, you do not need to select your variables in the main panel.
-The application correctly reads the data file directly.
+The simplest method to upload data in the application is to name the
+column names directly, as shown in the example files. Enter your data in
+a csv file, adding the columns 'id', ‘code’ and 'sample'. If the columns
+don't exist, fill them in with the same letter. By doing so, you do not
+need to select your variables in the main panel. The application
+correctly reads the data file directly.
 
@@
 
-Fabryka supports three angle measurement methods:
+fabryka supports three angle measurement methods:
 
@@
 
--   check 'Case 1: Only angles' if you have in your dataset the columns
-    'id', 'sample', 'code', direction' and 'dip'.
--   check 'Case 2: Only angles from DistoX2' if you have in your dataset
-    the same columns that above but your data comes from a DistoX2 and
-    needs to be converted.
+-   check 'Case 1: Only angles' if you have in your data set the columns
+    'id', 'sample', 'code', 'orientation' and 'dip'.
+-   check 'Case 2: Only angles from DistoX2' if you have in your data
+    set the same columns that above but your data comes from a DistoX2
+    and needs to be converted.
 -   check 'Case 3: Angles from DistoX2 with coordinates' if you have in
-    your dataset the same columns that above (Case 2, data from DistoX2)
-    with a single set of coordinates (x, y, z).
+    your data set the same columns that above (Case 2, data from
+    DistoX2) with a single set of coordinates (x, y, z).
 -   check 'Case 4: Angles and coordinates' if your data has the same
-    form that above (Case 3) but is coming from a compass and a
-    clinometer.
+    form that above (Case 3) but is coming from a compass and an
+    inclinometer.
 -   check 'Case 5: Two shots data without angles' if you have the
@@
 
-\begin{figure} [h]
+According to your method of angle measurement and the presence of
+coordinates you must select the right case. Please refer to the example
+files in the left side bar panel (fig.
+\ref{fig:figure_upload_your_data}E). It contains a data set coming from
+the Lower shelter of Le Moustier [@thomas2019; @texier2020]. The example files
+directly contain the right column names. You can use them directly in
+the application. Whatever the names of the columns in your file, you can
+select the appropriate columns in the column selector (fig.
+\ref{fig:figure_upload_your_data}F). Please keep only complete lines
+(*i.e.* no empty modalities, so no line without a fabric measurement).
+When a data table with new columns appears below the column selector,
+you are ready to use the application.
+
+If your data comes from a DistoX2, the columns 'orientation' (=
+'bearing', *i.e.* 0 to 360°) and 'dip' (= 'plunge') are automatically
+corrected in the new table. Whatever the form of your data set, two new
+columns 'orientation_pi' (0 to 180° angles) and 'angle_double' (see
+Krumbein 1939) appear. If the angles were measured using a total station
+(*i.e.* two shots data or 2 by 2 coordinates of plotted pieces refits), the orientation and the dip are calculated. If only one set of coordinates is provided with angles, a second set of
+coordinates is calculated and displayed in the new table (fig.
+\ref{fig:figure_upload_your_data}G). You can download the new data by
+clicking the 'copy', the 'csv' or the 'pdf' buttons. \bigbreak
+
+## Classical method
+
+The classical method (fig. \ref{fig:figure_classical_method}) can be
+computed for all sources of data and absence/presence of coordinates
+(*i.e.* all cases described above).
+
+\begin{figure} [H]
 \centering
@@
 
-According to your method of angle measurement and the presence of
-coordinates you must select the right case. Please refer to the example
-files in the left side bar panel (fig.
-\ref{fig:figure_upload_your_data}E). It contains a dataset coming from
-Le Moustier Lower shelter (Dordogne, France). The example files directly
-contain the right column names. You can use them directly in the
-application. Whatever the names of the columns in your file, you can
-select the appropriate columns in the column selector (fig.
-\ref{fig:figure_upload_your_data}F). Please keep only complete lines
-(*i.e.* no empty modalities, so no line without a fabric measurement).
-When a data table with new columns appears below the column selector,
-you are ready to use the application.
-
-If your data comes from a DistoX2, the columns 'direction' (= 'bearing',
-0 to 360° angles) and 'dip' (= 'plunge') are automatically corrected in
-the new table. Whatever the form of your dataset, two new columns
-'orientation_pi' (0 to 180° angles) and 'angle_double' (see Krumbein
-1939) appear. If the angles were measured using a total station (*i.e.*
-two shots data), the bearing and the plunge are also calculated. If only
-one set of coordinates is provided with angles, a second set of
-coordinates is calculated and displayed in the new table (fig.
-\ref{fig:figure_upload_your_data}G). You can download the new data by
-clicking the 'copy', the 'csv' or the 'pdf' buttons. \bigbreak
-
-## Classical method
-
-The classical method (fig. \ref{fig:figure_classical_method}) can be
-computed for all sources of data and absence/presence of coordinates
-(*i.e.* all cases described above).
-
 This panel includes:
 
--   Benn diagrams with possible integration of @bertran2002 model (fig. \ref{fig:figure_classical_method}D)
--   Rose diagrams for bearing and plunge (fig. \ref{fig:figure_classical_method}G)
+-   Benn diagrams with possible integration of @bertran2002 model (fig.
+    \ref{fig:figure_classical_method}D)
+-   Rose diagrams for axial data and dip (fig.
+    \ref{fig:figure_classical_method}G)
 -   Schmidt diagrams (fig. \ref{fig:figure_classical_method}H)
--   Summary data table with Benn indices, additional statistics, and orientation statistical tests.
+-   Woodcock diagrams
+-   Summary data table with Benn indices, additional statistics, and
+    orientation statistical tests.
 
@@
 
--   'n' stands for number of measures in the samples
--   'E1', 'E2' and 'E3' are the eigenvalues (see Woodcock 1977). They
+-   $N$ stands for number of measures in the samples
+-   $E1$, $E2$ and $E3$ are the eigenvalues [@watson1966]. They
     correspond to the sum of the cosines formed by the axis of the
@@
     the Benn indices
--   'IS', 'PL' and 'EL' are calculated thanks to the eigenvalues (Benn
-    1994). These are the Benn indices, ratios standing respectively for,
-    'isotropy' with IS = E3/E1, 'elongation' with EL = 1-(E2/E1) and
-    'planar' PL = 1-IS-EL
--   'L' is the vector magnitude (strength of the preferred artefact
-    orientation) and 'R.p' is the p-value result of the Rayleigh test.
--   'L.double' and 'R.p.double' are the same statistics that above for
+-   $IS$ and $EL$ are the Benn indices ratios standing
+    respectively for, 'isotropy' with $IS = E3/E1$ and 'elongation' with
+    $EL = 1-(E2/E1)$ [@benn1994].
+-   $K$ and $C$ are the @woodcock1977 indices.
+-   $L$ is the Vector Magnitude (strength of the preferred orientation)
+    and $R.p$ is the p-value result of the Rayleigh test.
+-   $L.double$ and $R.p.double$ are the same statistics that above for
     double angles [@krumbein1939].
+-   $Rao_p$ is the p_value of the @rao1967large test.
 
 You can control the data you plot and the data appearing in the figures
-and the summary data table by changing the checkboxes filters ('filter by sample'
-and 'filter by code') and the slider (*i.e.* minimum number of measures) in the left
-sidebar panel (fig. \ref{fig:figure_classical_method}A, B and C).
-Certain geological processes can have a different impact on objects
-depending on their physical properties (*e.g.* silex versus bones in a
-water flow as in the example files provided where bones have a strong
-preferential orientation and are much more 'linear' in the Benn diagram,
-see @thomas2019 and @texier2020). The natural processes models that
-you can add to the Benn diagram are calculated using data mainly from
-[@bertran1997; @bertran2002; @bertran2006; @lenoble2004].
-You will find the reference model data (with corresponding litarature references) which is used in the application in the 'model reference data' subpanel.
-In the rose diagram, orientations are considered (and not directions) as the Rayleigh test is perform on orientations [@curray1956] in the
-summary table. You can download all the figures by selecting a file format (.pdf, .jpeg and .png) and pressing the 'download' button (fig. \ref{fig:figure_classical_method}E and F).
+and the summary data table by changing the checkboxes filters ('filter
+by sample' and 'filter by code') and the slider (*i.e.* minimum number
+of measures) in the left sidebar panel (fig.
+\ref{fig:figure_classical_method}A, B and C).
 
-\begin{figure} [h]
+\begin{figure} [H]
 \centering
-  \includegraphics[width=17cm]{figure/classical_method.jpg}
+  \includegraphics[width=16cm]{figure/classical_method.jpg}
   \caption{A.
@@
 if you want to filter by codes, uncheck 'no code distinction', C.
-manage the minimum sample size to perform analyses, D.
-buld your model thanks to @bertran2002 data, E.
+manage the minimum sample size to perform analysis, D.
+buld your model thanks to Bertran and Lenoble (2002) data, E.
 and F.
 download figures by selecting a file format and pressing the 'download' button, G.
-rose diagrams subpanel, H.
+Rose diagrams subpanel, H.
 Schmidt diagram subpanel, I.
-summary table with additional statistics}
+Summary table with additional statistics}
 \label{fig:figure_classical_method}
 \end{figure}
+
+Certain geological processes can have a different impact on objects depending on their
+physical properties (*e.g.* silex versus bones in a water flow as in the
+example files provided where bones have a strong preferential
+orientation and are much more 'linear' in the Benn diagram [@thomas2019; @texier2020]. The natural processes models that you can add to the Benn diagram are calculated using data mainly from
+[@bertran1997; @bertran2002; @bertran2006; @lenoble2004]. You will find
+the reference model data (with corresponding literature references)
+which is used in the application in the 'Model reference data' subpanel.
+In the rose diagram, axial data is considered (and not orientations) as
+the Rayleigh test is perform on axial data [@curray1956] in the summary
+table. You can download all the figures by selecting a file format
+(.pdf, .jpeg and .png) and pressing the 'download' button (fig.
+\ref{fig:figure_classical_method}E and F).
 
@@
 
--   Benn diagrams with Benn indices computed with nearest neighbours points [@mcpherron2018]
+-   Benn diagrams with Benn indices computed with nearest neighbors
+    points [@mcpherron2018]
 -   Spatial projections of the points according to their fabric shape
 -   Summary data tables with orientation statistical tests, one at the
-    scale of each spatial series (each object and its n nearest
-    neighbours fabric = one row) and one at the scale of the sample (fig.
+    scale of each spatial series (each object and its $n$ nearest
+    neighbors fabric = one row) and one at the scale of the sample (fig.
     \ref{fig:figure_spatialised_method}F and G).
 
-In these tables:
+A table similar to the one done in the Classical method is proposed and
+contains the same statistics .
 
--   'n' stands for number of measures in the samples
--   'E1', 'E2' and 'E3' are the eigenvalues [@woodcock1977]. They
-    correspond to the sum of the cosines formed by the axis of the
-    objects and the three axes of the space. They are used to compute
-    the Benn indices.
--   'IS', 'PL' and 'EL' are calculated thanks to the eigenvalues [@benn1994]. These are the Benn indices, ratios standing respectively for,
-    'isotropy' with IS = E3/E1, 'elongation' with EL = 1-(E2/E1) and
-    'planar' PL = 1-IS-EL
--   'L' is the vector magnitude (strength of the preferred artefact
-    orientation) and 'R.p' is the p-value result of the Rayleigh test.
--   'L.double' and 'R.p.double' are the same statistics that above for
-    double angles [@krumbein1939].
-
-At the spatial series scale (*i.e.* each points and n neighbours),
+At the spatial series scale (*i.e.* each points and $n$ neighbors),
 statistics are provided in a table (Spatial series data table subpanel)
-with the Benn indices, the orientation rates L, L.double and p-values
-returned by the Rayleigh tests. At the sample scale, mean and standard
-deviations values of Benn indices, mean L and its corresponding Rayleigh
-tests means (mean values of all spatial series in the sample) and
-rejections ratios 'Tr' (formula **4.**) are given in the subpanel
-'Summary table of samples'.
+with the Benn indices, the orientation rates $L$, $L.double$ and
+$p-values$ returned by the Rayleigh tests. At the sample scale, mean and
+standard deviations values of Benn indices, mean $L$ and its
+corresponding Rayleigh tests means (mean values of all spatial series in
+the sample) and rejections ratios $Tr$ (formula **4.**) are given in the
+subpanel 'Summary table of samples'.
 
@@
 by sticks (by unchecking the corresponding box in the left sidebar
-panel, fig. \ref{fig:figure_spatialised_method}D) which colour is
+panel, fig. \ref{fig:figure_spatialised_method}D) which color is
 representative of the fabric of the series formed by the object and its
-n nearest neighbours. You can also add orthophoto in the background of
+$n$ nearest neighbors. You can also add orthophoto in the background of
 projections (fig. \ref{fig:figure_spatialised_method}E). In the left
-sidebar panel, you can control the samples considered in the analyses by
+sidebar panel, you can control the samples considered in the analysis by
 selecting the sample explored or by distinguishing the code used (*e.g.*
@@
 
+In this panel, you can perform a spatial analysis of your data starting
+from a Benn diagram or from projections. In both cases, you need to
+select the data you want to explore more locally in a new analysis. From
+the projections, you can select the points you want to explore directly.
+In the Benn diagram, you need to select the lasso selector (fig.
+\ref{fig:figure_spatial_exploration}D). Pay attention to the number of
+points selected and check that the new sample contains enough points to
+carry out the new analysis. If this is not the case, you can lower the
+slider in the left sidebar panel to perform the analysis with fewer
+neighbors (fig. \ref{fig:figure_spatial_exploration}E). You can download
+the new samples you have selected with spatial statistics from the
+corresponding subpanels (fig. \ref{fig:figure_spatial_exploration}I and
+K). \bigbreak
+
 \begin{figure} [h]
 \centering
-  \includegraphics[width=17cm]{figure/spatialised_method.jpg}
+  \includegraphics[width=16cm]{figure/spatialised_method.jpg}
   \caption{A.
@@
 
-In this panel, you can perform a spatialised analysis of your data
-starting from a Benn diagram or from projections. In both cases, you
-need to select the data you want to explore more locally in a new
-analysis. From the projections, you can select the points you want to
-explore directly. In the Benn diagram, you need to select the lasso
-selector (fig. \ref{fig:figure_spatial_exploration}D). Pay attention to
-the number of points selected and check that the new sample contains
-enough points to carry out the new analysis. If this is not the case,
-you can lower the slider in the left sidebar panel to perform the
-analysis with fewer neighbours (fig. \ref{fig:figure_spatial_exploration}E). You can download the new samples you have selected with spatial statistics from the corresponding
-subpanels (fig. \ref{fig:figure_spatial_exploration}I and K). \bigbreak
-
 ## Tutorial
@@
 The 'Tutorial' panel contains a summary of the mathematics, statistics
-and methods used in Fabryka and a short tutorial directly accessible
+and methods used in fabryka and a short tutorial directly accessible
 from the application. \bigbreak
@@
 \centering
-  \includegraphics[width=17cm]{figure/spatial_exploration.jpg}
+  \includegraphics[width=16cm]{figure/spatial_exploration.jpg}
   \caption{A.
@@
 
-## New models
+## New reference models
 
-The 'spatialised' method generates often more linear fabrics in
-comparison with the 'classical' method. This can be explained by the
-fact that objects that are close in space are more likely to resemble
-each other than objects that are further in space.
+The 'spatialised' method generates often more
+linear fabrics in comparison with the 'classical' method. This can be
+explained by the fact that objects that are close in space are more
+likely to resemble each other than objects that are further in space. A difference between global fabric ("classical method", *i.e.*, one point per sample) and spatialised fabric 
+(series of n neighbors) in Benn diagrams marks a spatial heterogeneity that needs to be explored.
 
@@
 sedimentary processes in order to build a new "spatialised" reference
-model (*cf.* the "classical" model proposed by Bertran and Lenoble
-2002). Please, do not hesitate to send referential data to feed these new models ! \bigbreak
+model (*cf.* the "classical" model proposed by @bertran2002). This work
+is in progress. This panel will be updated in the next version of the
+application.
+
+\bigbreak
 
@@
 
-The contributors panel aims at citing all contributors of this project
-which is open to everyone interested in improving fabric analysis methods or
-to contribute in building a new spatial model by sending reference data.
-If any request or contribution, please contact the author of this
-article. \bigbreak
+The contributors panel aims at citing all contributors of this project. \bigbreak
 
@@
 
--   use Fabryka $\beta$ 3.0:
-    <https://marchaeologist.shinyapps.io/fabryka/> \smallbreak
--   source code: <https://github.com/marchaeologist/fabryka/tree/main>
+-   use fabryka online: <https://marchaeologist.shinyapps.io/fabryka/>
     \smallbreak
--   Zenodo repository of Fabryka $\beta$ 3.0: <https://doi.org/10.5281/zenodo.15207419> \smallbreak
+-   If you are preparing a mission in a field without Internet access:
+
+Install [R](https://www.r-project.org) and [Rstudio
+Desktop](https://posit.co/download/rstudio-desktop/).
+
+Clone the fabryka repository from GitHub:
+<https://github.com/marchaeologist/fabryka> or download the code in
+"Code" and "Download ZIP" at the same address. This will give you the
+latest version of the source code. The application can be launched by
+opening the app.R file in the main folder and by running the following
+commands:
+
+```{r, eval=FALSE}
+
+pkgload::load_all(".")
+fabryka()
+
+```
+
+If you are using the application for the first time, Rstudio will install several packages used in the application.
+Lists of all these packages and dependencies are available in a DESCRIPTION file and a DEPENDECIES file in the github repository.
+
+-   source code: <https://github.com/marchaeologist/fabryka/tree/main> \smallbreak
+-   Report bugs or any request: <https://github.com/marchaeologist/fabryka/issues> \smallbreak
+-   github reprository doi of fabryka:
+    <https://doi.org/10.5281/zenodo.15207419> \smallbreak
+-   Zenodo repository of this article:
+    <https://doi.org/10.5281/zenodo.15236695> \bigbreak
 
@@
 with Jean-Pierre Texier, Arnaud Lenoble, Pascal Bertran and Dominique
-Todisco on issues related to site formation processes.
-The reference data used to build the models of the classical method were collected by P.
-Bertran throughout his career.
-P.
-Bertran provided us with all these data in order to feed the application.
-The developpement of the application also benefited from the help of Rémi Lemoy and François and Victor Baleux on mathematical issues.
-A first version of the application was tested by
-Emmanuel Discamps and Aurélien Royer. I would like to thank all these reaseachers for their advice and
-support. \bigbreak
+Todisco on issues related to site formation processes. The reference
+data used to build the models of the classical method were collected by
+P. Bertran throughout his career. P. Bertran provided us with all these
+data in order to feed the application. The development of the
+application also benefited from the help of Rémi Lemoy and François and
+Victor Baleux on mathematical issues. A first version of the application
+was tested by Emmanuel Discamps and Aurélien Royer. This project is
+supported by the "Service Valorisarion de la Recherche" of the
+University Toulouse Jean Jaurès. Finally, the constructive comments of
+Pascal Bertran, recommender of this article, and, of Frédéric
+Santos, Nicolas Frerebeau and Alfonso Benito-Calvo, reviewers of this
+article, considerably improved the manuscript and the web application. I
+would like to thank all these reaseachers for their advice. \bigbreak
+
+# Conflict of interest disclosure
+
+The authors declare that they comply with the PCI rule of having no financial conflicts of interest in relation to the content of the article. \bigbreak
 

### 2025-07-14 R/Fabryka.R

@@ -1,0 +1,4 @@
+# shiny::shinyApp() function
+fabryka <- function(){
+    shiny::shinyApp(ui = app_ui, server = app_server)
+}

### 2025-07-14 R/app_server.R

@@ -1,0 +1,3120 @@
+### --- server ---
+app_server <- function(input, output, session) {
+  
+  # increase file upload size limit in shiny to 50MB
+  base::options(shiny.maxRequestSize = 350 * 1024^2)
+  
+  # get the user data
+  user_data <- reactive({
+    req(!base::is.null(input$user_data))
+    
+    inFile <- input$user_data
+    
+    if (input$decimal == ".") {
+      data <- utils::read.csv(inFile$datapath, sep = input$sep)
+    } else {
+      data <- utils::read.csv2(inFile$datapath, sep = input$sep)
+    }
+    base::return(data)
+  })
+  
+  output$user_data_head <- renderTable(
+    utils::head(user_data(), 1)
+  )
+  
+  output$id1 <- renderUI({
+    selectInput("id", 
+                strong("unique 'ID'"),
+                choices = c("NULL", base::names(user_data())),
+                selected = "NULL"
+    )
+  })
+  
+  output$code1 <- renderUI({
+    selectInput("code", 
+                strong("'Code' (e.g. silex, fauna)"),
+                choices = c("NULL", base::names(user_data())),
+                selected = "NULL"
+    )
+  })
+  
+  output$sample1 <- renderUI({
+    selectInput("sample", 
+                strong("'Sample' (e.g. US, UL)"),
+                choices = c("NULL", base::names(user_data())),
+                selected = "NULL"
+    )
+  })
+  
+  output$x1 <- renderUI({
+    selectInput("x", 
+                strong("'X'"),
+                choices = c("NULL", base::names(user_data())),
+                selected = "NULL"
+    )
+  })
+  
+  output$y1 <- renderUI({
+    selectInput("y", 
+                strong("'Y'"),
+                choices = c("NULL", base::names(user_data())),
+                selected = "NULL"
+    )
+  })
+  
+  output$z1 <- renderUI({
+    selectInput("z", 
+                strong("'Z'"),
+                choices = c("NULL", base::names(user_data())),
+                selected = "NULL"
+    )
+  })
+  
+  output$x2 <- renderUI({
+    selectInput("x2", 
+                strong("'X2'"),
+                choices = c("NULL", base::names(user_data())),
[…]

### 2025-07-14 R/app_ui.R

@@ -1,0 +1,1054 @@
+### --- ui ---
+app_ui <- function() {
+  
+  addResourcePath(
+    prefix = "www",
+    directoryPath = system.file("www", package = "fabryka")
+  )
+  
+  ui <- tagList(
+  shinyjs::useShinyjs(),
+  navbarPage(
+    theme = shinythemes::shinytheme("yeti"),
+    p(strong("fabryka 1.1")),
+    tabPanel(
+      "Presentation",
+      mainPanel(
+        column(12,
+               align = "center",
+               tags$img(src = "www/hex_Fabryka4.png", height = 200),
+               tags$br()
+        ),
+        column(12,
+               align = "justify",
+               h2("Presentation"),
+               hr(),
+               p("fabryka is a R shiny application dedicated to the fabric analysis of archaeological remains. The app' is developped by Marc THOMAS", a(href = "https://orcid.org/0000-0002-8160-1910", strong("orcid.")), "This is the version", strong("1.1")),
+               p("An article and the source code are available via the links respectively", a(href = "https://zenodo.org/records/15236695", strong("here")), "and", a(href = "https://github.com/marchaeologist/fabryka/tree/main", strong("here."))),
+               tags$br(),
+               p(strong("The 'Upload data'"), "panel allows the user to upload different data sources."),
+               tags$br(),
+               p(strong("The 'Classical method'"), "panel contains subpanels to perform:"),
+               tags$li("Benn diagrams with possible integration of Bertran & Lenoble model (2002)"),
+               tags$li("Rose diagrams for axial data and dip"),
+               tags$li("Schmidt diagrams"),
+               tags$li("Woodcock diagrams"),
+               tags$li("Summary data table with orientation statistical tests."),
+               tags$br(),
+               p(strong("The 'Spatialised method'"), "panel contains subpanels to perform:"),
+               tags$li("Benn diagrams with Benn indices computed with nearest neighbors points (McPherron 2018)"),
+               tags$li("Spatial projections of the points according to their fabric shape"),
+               tags$li("Summary data tables with orientation statistical tests."),
+               tags$br(),
+               p(strong("The 'Spatial exploration'"), "panel contains subpanels to perform:"),
+               tags$li("A spatial exploration of your data starting from a Benn diagram"),
+               tags$li("A spatial exploration of your data starting from the spatial projections"),
+               tags$li("Summary data tables with orientation statistical tests."),
+               tags$br(),
[…]

### 2025-07-14 R/global.R

@@ -1,0 +1,309 @@
+# Load packages used in fabryka
+fabryka_packages = c("shiny","dplyr","ggplot2","plotly", "ggsci", "ggtern", "circular", "CircStats", "Ternary", "RStoolbox", "shinyjs", "rmarkdown", "shinythemes", "ggalt", "DT", "raster")
+is_installed = sapply(fabryka_packages, require, character.only=T)
+sapply(fabryka_packages[!is_installed], install.packages)
+
+# Modified function from circular package to perform Rao test
+#' @export
+print.rao.spacing.test.modified <- function(x, digits=4, ...) {
+  U <- x$statistic
+  alpha <- x$alpha
+  n <- x$n
+  data(rao.table, package='circular', envir=sys.frame(which=sys.nframe()))
+  if (n <= 30)
+    table.row <- n - 3
+  else if (n <= 32)
+    table.row <- 27
+  else if (n <= 37)
+    table.row <- 28
+  else if (n <= 42)
+    table.row <- 29
+  else if (n <= 47)
+    table.row <- 30
+  else if (n <= 62)
+    table.row <- 31
+  else if (n <= 87)
+    table.row <- 32
+  else if (n <= 125)
+    table.row <- 33
+  else if (n <= 175)
+    table.row <- 34
+  else if (n <= 250)
+    table.row <- 35
+  else if (n <= 350)
+    table.row <- 36
+  else if (n <= 450)
+    table.row <- 37
+  else if (n <= 550)
+    table.row <- 38
+  else if (n <= 650)
+    table.row <- 39
+  else if (n <= 750)
+    table.row <- 40
+  else if (n <= 850)
+    table.row <- 41
+  else if (n <= 950)
+    table.row <- 42
+  else table.row <- 43
+
+  cat("\n")
+  cat("       Rao's Spacing Test of Uniformity", "\n", "\n")
+  cat("Test Statistic =", round(U, digits=digits), "\n")
+
+  if (alpha == 0) {
+    if (U > rao.table[table.row, 1])
+      cat("P-value < 0.001", "\n", "\n")
+    else if (U > rao.table[table.row, 2])
+      cat("0.001 < P-value < 0.01", "\n", "\n")
+    else if (U > rao.table[table.row, 3])
+      cat("0.01 < P-value < 0.05", "\n", "\n")
+    else if (U > rao.table[table.row, 4])
[…]

### 2025-07-14 app.R

@@ -1,0 +1,7 @@
+pkgload::load_all(".")
+
+# You can launch the app by using :
+fabryka()
+
+#or:
+shiny::runApp(fabryka())

### 2025-07-14 fabryka_pci/article.Rmd
correction of fabryka shiny application and description article after first round of review in PCI archaeology

@@
 -   Zenodo repository of this article:
-    <https://doi.org/10.5281/zenodo.15236695> \bigbreak
+    <https://doi.org/10.5281/zenodo.15879462> \bigbreak
 

### 2025-07-14 fabryka/fabryka_beta_1.0.R
correction of fabryka shiny application and description article after first round of review in PCI archaeology

@@ -1,4329 +1,0 @@
-### --- R packages ---
-
-library(shiny)
-library(dplyr)
-library(ggplot2)
-library(plotly)
-library(ggsci)
-library(ggtern)
-library(RColorBrewer)
-library(circular)
-library(CircStats)
-library(Ternary)
-library(cowplot)
-library(RStoolbox)
-library(shinyjs)
-library(rmarkdown)
-library(rsconnect)
-
-
-### --- Model data from Bertran & Lenoble 2002 ---
-# model <- read.csv2("C:/Users/Surface User/Documents/articles/fabriques/fabryka/model.csv")
-# ternary_model <- model %>%
-#   dplyr::filter(area != "NA")
-
-### --- Model data from Bertran & Lenoble 2002 ---
-ternary_model <- data.frame(
-  Type = c("Snow Avalanche", "Snow Avalanche", "Snow Avalanche", "Snow Avalanche", "Snow Avalanche", "Snow Avalanche", "Snow Avalanche", "Snow Avalanche", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Debris Flow", "Mudslides", "Mudslides", "Runoff Shallow", "Runoff Shallow", "Runoff Shallow", "Runoff Shallow", "Runoff Shallow", "Runoff Shallow", "Runoff Shallow", "Runoff Shallow", "Runoff Shallow", "Runoff Shallow", "Runoff Shallow", "Runoff Shallow", "Runoff Shallow", "Runoff Shallow", "Runoff Shallow", "Runoff Shallow", "Runoff Steep", "Runoff Steep", "Runoff Steep", "Runoff Ste
[…]

### 2025-07-14 README.Rmd
correction of fabryka shiny application and description article after first round of review in PCI archaeology
correction of fabryka article and shiny application after first round of review in PCI archaeology

@@
 ---
-
-```{=html}
-<style>
-body {
-text-align: justify}
-</style>
-```
 
@@
 This will give you the latest version of the source code.
-The application can be launched by opening the app.R file in the main folder and by running the following commands:
 
-```{r, eval=FALSE}
+* Open fabryka.Rproj
+
+* If you are using the application for the first time, you have to install several packages used in the application.
+Lists of all these packages and dependencies are available in a DESCRIPTION file and a DEPENDECIES file in the github repository.
+To install fabryka packages and dependencies, copy and paste the following code lines into the R console:
+
+```{r packages_dep, eval=FALSE}
+
+fabryka_packages = c("shiny", "dplyr","ggplot2","plotly", "ggsci", "ggtern", "circular", "CircStats", "Ternary", 
+                   "RStoolbox", "shinyjs", "rmarkdown", "shinythemes",  "ggalt", "DT", "raster")
+
+is_installed = sapply(fabryka_packages, require, character.only=T)
+
+sapply(fabryka_packages[!is_installed], install.packages)
+
+```
+
+The application can be launched by opening the app.R file in the main folder and pressing "Run App" or by running the following commands:
+
+```{r run_app, eval=FALSE}
 
 pkgload::load_all(".")
+
 fabryka()

### 2025-07-14 fabryka_pci/article.Rmd
correction of fabryka article and shiny application after first round of review in PCI archaeology

@@
 
-# Useful links
+# Useful links and tips
 
--   use fabryka online: <https://marchaeologist.shinyapps.io/fabryka/>
-    \smallbreak
--   If you are preparing a mission in a field without Internet access:
+-   use fabryka online: <https://marchaeologist.shinyapps.io/fabryka/> \smallbreak
+-   source code: <https://github.com/marchaeologist/fabryka/tree/main> \smallbreak
+-   Report bugs or any request: <https://github.com/marchaeologist/fabryka/issues> \smallbreak
+-   github reprository doi of fabryka: <https://doi.org/10.5281/zenodo.15207419> \smallbreak
+-   Zenodo repository of this article: <https://doi.org/10.5281/zenodo.15879462> \bigbreak
 
-Install [R](https://www.r-project.org) and [Rstudio
+**If you are preparing a mission in a field without Internet access:**
+
+1. Install [R](https://www.r-project.org) and [Rstudio
 Desktop](https://posit.co/download/rstudio-desktop/).
 
-Clone the fabryka repository from GitHub:
+2. Clone the fabryka repository from GitHub:
 <https://github.com/marchaeologist/fabryka> or download the code in
 "Code" and "Download ZIP" at the same address. This will give you the
-latest version of the source code. The application can be launched by
-opening the app.R file in the main folder and by running the following
-commands:
+latest version of the source code.
+
+3. Open fabryka.Rproj
+
+4.
+If you are using the application for the first time, you have to install several packages used in the application.
+Lists of all these packages and dependencies are available in a DESCRIPTION file and a DEPENDECIES file in the github repository.
+To install fabryka packages and dependencies, copy and paste the following code lines into the R console:
+
+```{r packages_dep, eval=FALSE}
+
+fabryka_packages = c("shiny", "dplyr","ggplot2","plotly", "ggsci", "ggtern", "circular", 
+                     "CircStats", "Ternary", "RStoolbox", "shinyjs", "rmarkdown", 
+                     "shinythemes",  "ggalt", "DT", "raster")
+
+is_installed = sapply(fabryka_packages, require, character.only=T)
+
+sapply(fabryka_packages[!is_installed], install.packages)
+
+```
+
+5. The application can be launched by opening the app.R file in the main folder and pressing "Run App" or by running the following commands:
 
@@
 pkgload::load_all(".")
+
 fabryka()
@@
 ```
-
-If you are using the application for the first time, Rstudio will install several packages used in the application.
-Lists of all these packages and dependencies are available in a DESCRIPTION file and a DEPENDECIES file in the github repository.
-
--   source code: <https://github.com/marchaeologist/fabryka/tree/main> \smallbreak
--   Report bugs or any request: <https://github.com/marchaeologist/fabryka/issues> \smallbreak
--   github reprository doi of fabryka:
-    <https://doi.org/10.5281/zenodo.15207419> \smallbreak
--   Zenodo repository of this article:
-    <https://doi.org/10.5281/zenodo.15879462> \bigbreak
 

### 2025-07-15 README.Rmd
correction of fabryka article and shiny application after first round of review in PCI archaeology

@@
 
-[![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.15236695.svg)](https://doi.org/10.5281/zenodo.15236695)
+[![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.15880861.svg)](https://doi.org/10.5281/zenodo.15880861)
 
@@
 fabryka uses classical analytical procedures (Benn, rose, Schmidt and Woodcock diagrams and statistical tests on orientations) for analysing the fabrics of objects, an spatial method and new ways of spatial and statistical exploration.
-A descriptive article is available at <https://zenodo.org/records/15236695>.
+A descriptive article is available at <https://zenodo.org/records/15880861>.
 
@@
 
-A tutorial is included in the application (*i.e.* 'Tutorial" panel) and can also be found in a dedicated article: <https://zenodo.org/records/15236695>.
+A tutorial is included in the application (*i.e.* 'Tutorial" panel) and can also be found in a dedicated article: <https://zenodo.org/records/15880861>.
 'fabryka' includes an example file you can download from the application ('Upload data' subpanel).
@@
 
-THOMAS, M. (2025). Fabryka: A web application for easily analysing & exploring the fabric of archaeological assemblages. In PCI archaeology (Version 1). Zenodo. https://doi.org/10.5281/zenodo.15236695
+THOMAS, M. (2025). Fabryka: A web application for easily analysing & exploring the fabric of archaeological assemblages. In PCI archaeology (Version 1). Zenodo. https://doi.org/10.5281/zenodo.15880861

### 2025-07-15 fabryka_pci/article.Rmd
correction of fabryka article and shiny application after first round of review in PCI archaeology

@@
 -   Report bugs or any request: <https://github.com/marchaeologist/fabryka/issues> \smallbreak
--   github reprository doi of fabryka: <https://doi.org/10.5281/zenodo.15207419> \smallbreak
+-   github reprository doi of fabryka: <https://doi.org/10.5281/zenodo.15881625> \smallbreak
 -   Zenodo repository of this article: <https://doi.org/10.5281/zenodo.15879462> \bigbreak

### 2025-07-15 fabryka_pci/article.Rmd
correction of fabryka article and shiny application after first round of review in PCI archaeology

@@
 -   Report bugs or any request: <https://github.com/marchaeologist/fabryka/issues> \smallbreak
--   github reprository doi of fabryka: <https://doi.org/10.5281/zenodo.15881625> \smallbreak
--   Zenodo repository of this article: <https://doi.org/10.5281/zenodo.15879462> \bigbreak
+-   github reprository doi of fabryka: <https://doi.org/10.5281/zenodo.15909666> \smallbreak
+-   Zenodo repository of this article: <https://doi.org/10.5281/zenodo.15880861> \bigbreak
 

### 2025-07-30 README.Rmd
minor changes in article text after second round of review in PCI archaeology

@@
 
-[![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.15880861.svg)](https://doi.org/10.5281/zenodo.15880861)
+[![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.16601423.svg)](https://doi.org/10.5281/zenodo.16601423)
 
@@
 fabryka uses classical analytical procedures (Benn, rose, Schmidt and Woodcock diagrams and statistical tests on orientations) for analysing the fabrics of objects, an spatial method and new ways of spatial and statistical exploration.
-A descriptive article is available at <https://zenodo.org/records/15880861>.
+A descriptive article is available at <https://zenodo.org/records/16601423>.
 
@@
 * clone the fabryka repository from GitHub: <https://github.com/marchaeologist/fabryka> or download the code in "Code" and "Download ZIP" at the same address.
+Unzip the folder.
 This will give you the latest version of the source code.
@@
 
-A tutorial is included in the application (*i.e.* 'Tutorial" panel) and can also be found in a dedicated article: <https://zenodo.org/records/15880861>.
+A tutorial is included in the application (*i.e.* 'Tutorial" panel) and can also be found in a dedicated article: <https://zenodo.org/records/16601423>.
 'fabryka' includes an example file you can download from the application ('Upload data' subpanel).
@@
 
-THOMAS, M. (2025). Fabryka: A web application for easily analysing & exploring the fabric of archaeological assemblages. In PCI archaeology (Version 1). Zenodo. https://doi.org/10.5281/zenodo.15880861
+THOMAS, M. (2025). Fabryka: A web application for easily analysing & exploring the fabric of archaeological assemblages. In PCI archaeology (Version 1). Zenodo. https://doi.org/10.5281/zenodo.16601423

### 2025-07-30 fabryka_pci/article.Rmd
minor changes in article text after second round of review in PCI archaeology

@@
 title: '**fabryka**'
-subtitle: A web application for easily analysing & exploring the fabric of archaeological assemblages
+subtitle: A web application for easily analysing & exploring the fabric of archaeological
+  assemblages
 output:
-  pdf_document: default
+  word_document: default
+  pdf_document:
+    keep_tex: true
 header-includes:
@@
 involved in the formation of the deposits and their impact on the
-preservation of the archaeological assemblages. These statistical
-analysis are not easily accessible to non-specialists and are
+preservation of the archaeological assemblages. This statistical
+analysis is not easily accessible to non-specialists and is
 particularly time-consuming, sometimes requiring the use of several
@@
 a spatial method. New statistical methods and new ways of exploration
-are proposed in the application and this present article.\smallbreak
+are proposed in the application and in this present article.\smallbreak
 
@@
 Objects abandoned by prehistoric people can be considered as sedimentary
-particles and are affected by the same processes [@schiffer1983]. As
+particles and are affected by natural processes [@schiffer1983]. As
 geological particles are often not elongated enough, analysis of the
-fabric of sedimentary deposits depends mainly on archaeological material
+fabric of sedimentary deposits often depends mainly on archaeological material
 (bones and lithics). This makes it possible to identify the sedimentary
 processes to which they were subjected and to discuss their degree of
-spatial disorganization.
+spatial reorganization.
+When possible, distinguishing sedimentary particles from archaeological objects in the fabric analysis could allow to test whether these two populations were affected by different processes (*e.g.*, the fabric of an undisturbed occupation covered by deposits from a minor flood should reveal a distinct fabric from that of the alluvial pebbles on which it rests).
 
@@
 characterize the fabric of a given assemblage [@bertran2002;
-@mcpherron2018; @de2021new].
-
-The first methods focus on the orientation
-of the measurements only and aim to test the null hypothesis of a random
-orientation from a uniform distribution through several statistical
-tests [@krumbein1939; @curray1956; @rao1967large].
-
-The second statistical method take into account object orientation and
+@mcpherron2018; @de2021new]. \newline
+The first methods focus on the orientation only and aim to test the null hypothesis of a random
+orientation of objects through several statistical
+tests [@krumbein1939; @curray1956; @rao1967large].\newline
+The second statistical method takes into account object orientation and
 dip. Three normalized eigenvalues (E1, E2 and E3) are calculated from
@@
 interfaces to facilitate the task of non-programmer users
-[@mcpherron2018].
-
-The following sections explain all the selected approaches and our
+[@mcpherron2018]. The following sections explain all the selected approaches and our
 proposals for processing the fabrics thanks to fabryka, from taking
@@
 \centering
-  \includegraphics[width=16cm]{figure/presentation_applications.png}
+  \includegraphics[width=15cm]{figure/presentation_applications.png}
   \caption{fabryka creation scheme and organisation of the article}
@@
 assemblage. The orientation (= bearing) of the lowest point is measured
-in the range [0; 360] with North as the reference (*i.e.* orientation =
+in the range [0; 360[ with North as the reference (*i.e.* orientation =
 0°). The dip (= plunge) is measured in the range [0; 90] with horizontal
-as a reference. The orientation [0; 360] can be converted into axial
-data measured in the range [0; 180] by subtracting 180° from the
-measurements in the range ]180; 360].
+as a reference. The orientation [0; 360[ can be converted into axial
+data measured in the range [0; 180[ by subtracting 180° from the
+measurements in the range [180; 360].
 
-Traditionally, the orientation of the objects lowest point is measured
-using a compass and the dip is measured using an inclinometer. The aim
-is to measure the orientation and dip of the physical axis of elongation
+Traditionally, the azimuth of the principal axis of an object is measured using a compass considering that the point with the lowest altitude indicates the orientation of the vector.
+The dip is measured using an inclinometer.
+The aim is to measure the orientation and dip of the elongaton axis of
 of objects ignoring its shape. To do this, the measuring device must be
-placed parallel to the general axis of elongation of the object.
+placed parallel to the elongation axis of the object.
 Measuring the fabric from its imprint or visible surface is biased by
@@
 units seem to be the most coherent entity which needs to be explored
-because they are defined according to geological criterion, which a
+because they are defined according to geological criteria, which a
 priori guarantee a certain coherence in terms of the formation dynamics
@@
     Vector Magnitude $L$ (expressed as a percentage) by Curray's
-    [-@curray1956] formula tests the hypothesis of a unimodal
+    [-@curray1956] formula tests the hypothesis of an unimodal
     distribution of orientations. If the $p$ value returned by the
-    Rayleigh test is less than a threshold of 0.05, the null hypothesis
-    of a random orientation from a uniform distribution is rejected. The
+    Rayleigh test is less than the threshold of 0.05, the null hypothesis
+    of a random sampling from a uniform distribution is rejected. The
     formula for calculating the Vector Magnitude $L$ is as follows:
@@
 example files provided where bones have a strong preferential
-orientation and are much more 'linear' in the Benn diagram [@thomas2019; @texier2020]. The natural processes models that you can add to the Benn diagram are calculated using data mainly from
+orientation and are much more 'linear' in the Benn diagram [@thomas2019; @texier2020]). The natural processes models that you can add to the Benn diagram are calculated using data mainly from
 [@bertran1997; @bertran2002; @bertran2006; @lenoble2004]. You will find
@@
 -   github reprository doi of fabryka: <https://doi.org/10.5281/zenodo.15909666> \smallbreak
--   Zenodo repository of this article: <https://doi.org/10.5281/zenodo.15880861> \bigbreak
+-   Zenodo repository of this article: <https://doi.org/10.5281/zenodo.16601423> \bigbreak
 
@@
 <https://github.com/marchaeologist/fabryka> or download the code in
-"Code" and "Download ZIP" at the same address. This will give you the
+"Code" and "Download ZIP" at the same address. Unzip the folder. This will give you the
 latest version of the source code.
@@
 data used to build the models of the classical method were collected by
-P. Bertran throughout his career. P. Bertran provided us with all these
+P. Bertran throughout his career. Pascal Bertran provided us with all these
 data in order to feed the application. The development of the
-application also benefited from the help of Rémi Lemoy and François and
+application also benefited from the help of Rémi Lemoy, François Baleux and
 Victor Baleux on mathematical issues. A first version of the application
@@
 article, considerably improved the manuscript and the web application. I
-would like to thank all these reaseachers for their advice. \bigbreak
+would like to thank all these reasearchers for their advice. \bigbreak
 

### 2025-07-30 app.R

@@ -1,7 +1,8 @@
+# load all dependencies, functions, data and application server and ui
 pkgload::load_all(".")
 
-# You can launch the app by using :
+# launch the app by running :
 fabryka()
 
-#or:
+# or:
 shiny::runApp(fabryka())

### 2025-08-07 fabryka_pci/article.Rmd
peer review completed (PCI archaeology)

@@
 ---
-title: '**fabryka**'
+title: "\\\n **fabryka**"
 subtitle: A web application for easily analysing & exploring the fabric of archaeological
@@
 output:
-  word_document: default
   pdf_document:
     keep_tex: true
+  word_document: default
 header-includes:
+- \usepackage{titling}
+- \pretitle{\begin{flushright}
+- \includegraphics[width=2in,height=2in]{logo.png}\LARGE\\}
+- \posttitle{\end{flushright}}
 - \usepackage{amsmath}
-- \usepackage{lineno}
-- \linenumbers
 - \usepackage{fancyhdr}
@@
   chunk_output_type: console
+geometry: margin=2.54cm
 bibliography: references.bib
@@
 
-\begin{figure} [H]
-\centering
-  \includegraphics[width=4cm]{figure/hex_Fabryka4.png}
-\label{}
-\end{figure}
-
-Marc THOMAS ^1\*^ (<https://orcid.org/0000-0002-8160-1910>) \smallbreak
+**Marc Thomas** ^1\*^ (<https://orcid.org/0000-0002-8160-1910>) \smallbreak
 
@@
 E-mail: [marcthomas1\@hotmail.fr](mailto:marcthomas1@hotmail.fr){.email}
-\bigbreak
+\smallbreak
 
-**Key words** : R shiny web application, fabric, site formation
+**Key words**: R shiny web application, fabric, site formation
 processes, spatial analysis, orientation and dip \bigbreak
@@
 preservation of the archaeological assemblages. This statistical
-analysis is not easily accessible to non-specialists and is
-particularly time-consuming, sometimes requiring the use of several
-software, some of which are chargeable. fabryka uses classical
-analytical procedures (Benn, rose, Schmidt and Woodcock diagrams and statistical
-tests on orientations) for analysing the fabrics of objects and includes
-a spatial method. New statistical methods and new ways of exploration
-are proposed in the application and in this present article.\smallbreak
+analysis is not easily accessible to non-specialists and is particularly
+time-consuming, sometimes requiring the use of several software, some of
+which are chargeable. fabryka uses classical analytical procedures
+(Benn, rose, Schmidt and Woodcock diagrams and statistical tests on
+orientations) for analysing the fabrics of objects and includes a
+spatial method. New statistical methods and new ways of exploration are
+proposed in the application and in this present article.\smallbreak
 
@@
 @xie2018; @xie2020], are both available online in a public github
-repository (<https://github.com/marchaeologist/fabryka/tree/main>).
+repository
+(<https://github.com/marchaeologist/fabryka/tree/main>).
 \pagebreak
@@
 geological particles are often not elongated enough, analysis of the
-fabric of sedimentary deposits often depends mainly on archaeological material
-(bones and lithics). This makes it possible to identify the sedimentary
-processes to which they were subjected and to discuss their degree of
-spatial reorganization.
-When possible, distinguishing sedimentary particles from archaeological objects in the fabric analysis could allow to test whether these two populations were affected by different processes (*e.g.*, the fabric of an undisturbed occupation covered by deposits from a minor flood should reveal a distinct fabric from that of the alluvial pebbles on which it rests).
+fabric of sedimentary deposits often depends mainly on archaeological
+material (bones and lithics). This makes it possible to identify the
+sedimentary processes to which they were subjected and to discuss their
+degree of spatial reorganization. When possible, distinguishing
+sedimentary particles from archaeological objects in the fabric analysis
+could allow to test whether these two populations were affected by
+different processes (*e.g.*, the fabric of an undisturbed occupation
+covered by deposits from a minor flood should reveal a distinct fabric
+from that of the alluvial pebbles on which it rests).
 
@@
 characterize the fabric of a given assemblage [@bertran2002;
-@mcpherron2018; @de2021new]. \newline
-The first methods focus on the orientation only and aim to test the null hypothesis of a random
-orientation of objects through several statistical
-tests [@krumbein1939; @curray1956; @rao1967large].\newline
-The second statistical method takes into account object orientation and
-dip. Three normalized eigenvalues (E1, E2 and E3) are calculated from
-the cosine formed by the axis of the artifact with the three axes of
-space [@watson1966]. @woodcock1977 and @benn1994 indices are used to
-describe the fabric shape. Several authors *e.g.*, [@bertran1997;
-@bertran2002; @lenoble2004] built models to compare the fabric of
-archaeological assemblages to natural sedimentary processes and human
-processes (*i.e.*, fabric of knapping spots). The null hypothesis of an
-absence of perturbation is a fabric shape compatible with that of
-experimental knapping spots. These methods are described in the
-Mathematics section of this article.
+@mcpherron2018; @de2021new]. \newline The first methods focus on the
+orientation only and aim to test the null hypothesis of a random
+orientation of objects through several statistical tests [@krumbein1939;
+@curray1956; @rao1967large].\newline The second statistical method takes
+into account object orientation and dip. Three normalized eigenvalues
+(E1, E2 and E3) are calculated from the cosine formed by the axis of the
+artifact with the three axes of space [@watson1966]. @woodcock1977 and
+@benn1994 indices are used to describe the fabric shape. Several authors
+*e.g.*, [@bertran1997; @bertran2002; @lenoble2004] built models to
+compare the fabric of archaeological assemblages to natural sedimentary
+processes and human processes (*i.e.*, fabric of knapping spots). The
+null hypothesis of an absence of perturbation is a fabric shape
+compatible with that of experimental knapping spots. These methods are
+described in the Mathematics section of this article.
 
@@
 interfaces to facilitate the task of non-programmer users
-[@mcpherron2018]. The following sections explain all the selected approaches and our
-proposals for processing the fabrics thanks to fabryka, from taking
-measurements in the field, through the statistical methods and graphical
-modes of representation used in the application (fig.
-\ref{fig:figure_presentation}). \bigbreak
+[@mcpherron2018]. The following sections explain all the selected
+approaches and our proposals for processing the fabrics thanks to
+fabryka, from taking measurements in the field, through the statistical
+methods and graphical modes of representation used in the application
+(fig. \ref{fig:figure_presentation}). \bigbreak
 
@@
 
-During excavation, the orientation and dip of elongated remains (*i.e.* a > 2 * b and with a > 2 cm) is
-measured on a representative set of artefacts or on the whole
-assemblage. The orientation (= bearing) of the lowest point is measured
-in the range [0; 360[ with North as the reference (*i.e.* orientation =
-0°). The dip (= plunge) is measured in the range [0; 90] with horizontal
-as a reference. The orientation [0; 360[ can be converted into axial
-data measured in the range [0; 180[ by subtracting 180° from the
-measurements in the range [180; 360].
+During excavation, the orientation and dip of elongated remains (*i.e.*
+a \> 2 \* b and with a \> 2 cm) is measured on a representative set of
+artefacts or on the whole assemblage. The orientation (= bearing) of the
+lowest point is measured in the range [0; 360[ with North as the
+reference (*i.e.* orientation = 0°). The dip (= plunge) is measured in
+the range [0; 90] with horizontal as a reference. The orientation [0;
+360[ can be converted into axial data measured in the range [0; 180[ by
+subtracting 180° from the measurements in the range [180; 360].
 
-Traditionally, the azimuth of the principal axis of an object is measured using a compass considering that the point with the lowest altitude indicates the orientation of the vector.
-The dip is measured using an inclinometer.
-The aim is to measure the orientation and dip of the elongaton axis of
-of objects ignoring its shape. To do this, the measuring device must be
-placed parallel to the elongation axis of the object.
-Measuring the fabric from its imprint or visible surface is biased by
-its shape, making impossible to assess its true dip. Since the
+Traditionally, the azimuth of the principal axis of an object is
+measured using a compass considering that the point with the lowest
+altitude indicates the orientation of the vector. The dip is measured
+using an inclinometer. The aim is to measure the orientation and dip of
+the elongaton axis of of objects ignoring its shape. To do this, the
+measuring device must be placed parallel to the elongation axis of the
+object. Measuring the fabric from its imprint or visible surface is
+biased by its shape, making impossible to assess its true dip. Since the
 theodolite was introduced to excavations in the 1990s, the fabrics is
-sometimes measured by recording the coordinates of the ends of objects [@mcpherron2005artifact].
-In the latter case, the rules of trigonometry are applied to convert the
-pair of coordinates into orientations and dips.
+sometimes measured by recording the coordinates of the ends of objects
+[@mcpherron2005artifact]. In the latter case, the rules of trigonometry
+are applied to convert the pair of coordinates into orientations and
+dips.
 
@@
     distribution of orientations. If the $p$ value returned by the
-    Rayleigh test is less than the threshold of 0.05, the null hypothesis
-    of a random sampling from a uniform distribution is rejected. The
-    formula for calculating the Vector Magnitude $L$ is as follows:
+    Rayleigh test is less than the threshold of 0.05, the null
+    hypothesis of a random sampling from a uniform distribution is
+    rejected. The formula for calculating the Vector Magnitude $L$ is as
+    follows:
 
@@
 The method consisting of doubling the value of the axial angles before
-calculating the Vector Magnitude $L$ and the Rayleigh test, as proposed by
-Krumbein [-@krumbein1939], tests the hypothesis of a bimodal
+calculating the Vector Magnitude $L$ and the Rayleigh test, as proposed
+by Krumbein [-@krumbein1939], tests the hypothesis of a bimodal
 distribution with a period of 90°. The hypothesis of a 90° period
@@
 varies [@curray1956; @bertran2002]. For example, when the Vector
-Magnitude $L$ is 25%, the Rayleigh test gives $p-values$ below a critical
-threshold of 0.05 for series with more than 50 measurements (fig. 2).
+Magnitude $L$ is 25%, the Rayleigh test gives $p-values$ below a
+critical threshold of 0.05 for series with more than 50 measurements
+(fig. 2).
 
@@
 
-The fabric shape can be described by calculating the ratios $r1$ = ln(E1/E2), $r2$ = ln(E2/E3) and $K$ = r1/r2 [@woodcock1977].
-When 0 \< $K$ \< 1, the fabric is planar.
-When 1 \< $K$ \< $+\infty$, the fabric is linear.
-The parameter $C$ = ln(E1/E3) is a measure of the strength of the preferred orientation [@woodcock1977].
-These indices can be plotted into a Woodcock diagram.
+The fabric shape can be described by calculating the ratios $r1$ =
+ln(E1/E2), $r2$ = ln(E2/E3) and $K$ = r1/r2 [@woodcock1977]. When 0 \<
+$K$ \< 1, the fabric is planar. When 1 \< $K$ \< $+\infty$, the fabric
+is linear. The parameter $C$ = ln(E1/E3) is a measure of the strength of
+the preferred orientation [@woodcock1977]. These indices can be plotted
+into a Woodcock diagram.
 
-The Benn indices [@benn1994] describe the fabric shape of a sample and can be plotted in a ternary diagram (*i.e.* Benn diagram).
-The elongation index EL is equal to $1-(E2/E1)$ and the isotropy index $IS$ is $E3/E1$.
-When $EL$ is high, the axes of the objects are grouped around an orientation.
-When $IS$ is high, the fabric is said to be isotropic meaning the axes of the objects are uniformly distributed in all directions.
-Finally, when the value $1-EL-IS$ is high, the fabric is said to be planar, with the axes of the objects grouped around a plane.
+The Benn indices [@benn1994] describe the fabric shape of a sample and
+can be plotted in a ternary diagram (*i.e.* Benn diagram). The
+elongation index EL is equal to $1-(E2/E1)$ and the isotropy index $IS$
+is $E3/E1$. When $EL$ is high, the axes of the objects are grouped
+around an orientation. When $IS$ is high, the fabric is said to be
+isotropic meaning the axes of the objects are uniformly distributed in
+all directions. Finally, when the value $1-EL-IS$ is high, the fabric is
+said to be planar, with the axes of the objects grouped around a plane.
 \bigskip
@@
 \ref{fig:figure_upload_your_data}E). It contains a data set coming from
-the Lower shelter of Le Moustier [@thomas2019; @texier2020]. The example files
-directly contain the right column names. You can use them directly in
-the application. Whatever the names of the columns in your file, you can
-select the appropriate columns in the column selector (fig.
+the Lower shelter of Le Moustier [@thomas2019; @texier2020]. The example
+files directly contain the right column names. You can use them directly
+in the application. Whatever the names of the columns in your file, you
+can select the appropriate columns in the column selector (fig.
 \ref{fig:figure_upload_your_data}F). Please keep only complete lines
@@
 Krumbein 1939) appear. If the angles were measured using a total station
-(*i.e.* two shots data or 2 by 2 coordinates of plotted pieces refits), the orientation and the dip are calculated. If only one set of coordinates is provided with angles, a second set of
-coordinates is calculated and displayed in the new table (fig.
+(*i.e.* two shots data or 2 by 2 coordinates of plotted pieces refits),
+the orientation and the dip are calculated. If only one set of
+coordinates is provided with angles, a second set of coordinates is
+calculated and displayed in the new table (fig.
 \ref{fig:figure_upload_your_data}G). You can download the new data by
@@
     the Benn indices
--   $IS$ and $EL$ are the Benn indices ratios standing
-    respectively for, 'isotropy' with $IS = E3/E1$ and 'elongation' with
-    $EL = 1-(E2/E1)$ [@benn1994].
+-   $IS$ and $EL$ are the Benn indices ratios standing respectively for,
+    'isotropy' with $IS = E3/E1$ and 'elongation' with $EL = 1-(E2/E1)$
+    [@benn1994].
 -   $K$ and $C$ are the @woodcock1977 indices.
@@
 
-Certain geological processes can have a different impact on objects depending on their
-physical properties (*e.g.* silex versus bones in a water flow as in the
-example files provided where bones have a strong preferential
-orientation and are much more 'linear' in the Benn diagram [@thomas2019; @texier2020]). The natural processes models that you can add to the Benn diagram are calculated using data mainly from
+Certain geological processes can have a different impact on objects
+depending on their physical properties (*e.g.* silex versus bones in a
+water flow as in the example files provided where bones have a strong
+preferential orientation and are much more 'linear' in the Benn diagram
+[@thomas2019; @texier2020]). The natural processes models that you can
+add to the Benn diagram are calculated using data mainly from
 [@bertran1997; @bertran2002; @bertran2006; @lenoble2004]. You will find
@@
 
-The 'spatialised' method generates often more
-linear fabrics in comparison with the 'classical' method. This can be
-explained by the fact that objects that are close in space are more
-likely to resemble each other than objects that are further in space. A difference between global fabric ("classical method", *i.e.*, one point per sample) and spatialised fabric 
-(series of n neighbors) in Benn diagrams marks a spatial heterogeneity that needs to be explored.
+The 'spatialised' method generates often more linear fabrics in
+comparison with the 'classical' method. This can be explained by the
+fact that objects that are close in space are more likely to resemble
+each other than objects that are further in space. A difference between
+global fabric ("classical method", *i.e.*, one point per sample) and
+spatialised fabric (series of n neighbors) in Benn diagrams marks a
+spatial heterogeneity that needs to be explored.
 
@@
 
-The contributors panel aims at citing all contributors of this project. \bigbreak
+The contributors panel aims at citing all contributors of this project.
+\bigbreak
 
-# Useful links and tips
+# Data, scripts, code, and supplementary information availability
 
--   use fabryka online: <https://marchaeologist.shinyapps.io/fabryka/> \smallbreak
--   source code: <https://github.com/marchaeologist/fabryka/tree/main> \smallbreak
--   Report bugs or any request: <https://github.com/marchaeologist/fabryka/issues> \smallbreak
--   github reprository doi of fabryka: <https://doi.org/10.5281/zenodo.15909666> \smallbreak
--   Zenodo repository of this article: <https://doi.org/10.5281/zenodo.16601423> \bigbreak
+-   use fabryka online: <https://marchaeologist.shinyapps.io/fabryka/>
+    \smallbreak
+-   source code: <https://github.com/marchaeologist/fabryka/tree/main>
+    \smallbreak
+-   Report bugs or any request:
+    <https://github.com/marchaeologist/fabryka/issues> \smallbreak
+-   github reprository doi of fabryka:
+    <https://doi.org/10.5281/zenodo.15909666> \smallbreak
+-   Zenodo repository of this article:
+    <https://doi.org/10.5281/zenodo.16758152> \bigbreak
 
@@
     <https://github.com/marchaeologist/fabryka> or download the code in
-"Code" and "Download ZIP" at the same address. Unzip the folder. This will give you the
-latest version of the source code.
+    "Code" and "Download ZIP" at the same address. Unzip the folder.
+    This will give you the latest version of the source code.
 
@@
 
-4.
-If you are using the application for the first time, you have to install several packages used in the application.
-Lists of all these packages and dependencies are available in a DESCRIPTION file and a DEPENDECIES file in the github repository.
-To install fabryka packages and dependencies, copy and paste the following code lines into the R console:
+4.  If you are using the application for the first time, you have to
+    install several packages used in the application. Lists of all these
+    packages and dependencies are available in a DESCRIPTION file and a
+    DEPENDECIES file in the github repository. To install fabryka
+    packages and dependencies, copy and paste the following code lines
+    into the R console:
 
@@
 
-5. The application can be launched by opening the app.R file in the main folder and pressing "Run App" or by running the following commands:
+5.  The application can be launched by opening the app.R file in the
+    main folder and pressing "Run App" or by running the following
+    commands:
 
@@
 author, with a few modifications. This work benefited from discussions
-with Jean-Pierre Texier, Arnaud Lenoble, Pascal Bertran and Dominique
+with Jean-Pierre Texier, Arnaud Lenoble, Pascal Bertran, Thomas Perrin and Dominique
 Todisco on issues related to site formation processes. The reference
 data used to build the models of the classical method were collected by
-P. Bertran throughout his career. Pascal Bertran provided us with all these
-data in order to feed the application. The development of the
-application also benefited from the help of Rémi Lemoy, François Baleux and
-Victor Baleux on mathematical issues. A first version of the application
-was tested by Emmanuel Discamps and Aurélien Royer. This project is
-supported by the "Service Valorisarion de la Recherche" of the
-University Toulouse Jean Jaurès. Finally, the constructive comments of
-Pascal Bertran, recommender of this article, and, of Frédéric
-Santos, Nicolas Frerebeau and Alfonso Benito-Calvo, reviewers of this
-article, considerably improved the manuscript and the web application. I
-would like to thank all these reasearchers for their advice. \bigbreak
+P. Bertran throughout his career. Pascal Bertran provided us with all
+these data in order to feed the application. The development of the
+application also benefited from the help of Rémi Lemoy, François Baleux
+and Victor Baleux on mathematical issues. A first version of the
+application was tested by Emmanuel Discamps and Aurélien Royer. This
+project is supported by the "Service Valorisarion de la Recherche" of
+the University Toulouse Jean Jaurès.
+A preprint version of this article (v3, https://doi.org/10.5281/zenodo.16601423) has been peer-reviewed and recommended by PCI Archaeo (https://doi.org/10.24072/pci.archaeo.100611).
+Comments made by Pascal Bertran, recommender of this article, and, by Frédéric Santos, Nicolas Frerebeau and Alfonso Benito-Calvo, reviewers of this article, considerably improved the manuscript and the application.
+I would like to thank all these reasearchers for their advice.
+\bigbreak
+
+# Funding
+
+This research was funded by the Service Valorisation de la Recherche of Toulouse - Jean Jaurès University. \bigbreak
 
@@
 
-The authors declare that they comply with the PCI rule of having no financial conflicts of interest in relation to the content of the article. \bigbreak
+The authors declare that they comply with the PCI rule of having no
+financial conflicts of interest in relation to the content of the
+article. \bigbreak
 
-# Bibliography
+# References
 
-The 'Bibliography' panel in the application contains the bibliography
+The 'Bibliography' panel in the application contains the references
 cited in the application. This article contains the same bibliography

### 2025-08-07 fabryka_pci/article.tex
peer review completed (PCI archaeology)

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=2.54cm]{geometry}
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\usepackage{graphicx}
\makeatletter
\newsavebox\pandoc@box
\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
  \sbox\pandoc@box{#1}%
  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
  \else\usebox{\pandoc@box}%
  \fi%
}
% Set default figure placement to htbp
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{titling}
\pretitle{\begin{flushright}
\includegraphics[width=2in,height=2in]{logo.png}\LARGE\\}
\posttitle{\end{flushright}}
\usepackage{amsmath}
\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhead[L]{fabryka}
\fancyhead[R]{R web application for fabric analysis}
\fancyfoot[C]{\thepage}
\usepackage{float}
\floatplacement{figure}{H}
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={ fabryka},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{\hfill\break
\textbf{fabryka}}
\usepackage{etoolbox}
\makeatletter
\providecommand{\subtitle}[1]{% add subtitle to \maketitle
  \apptocmd{\@title}{\par {\large #1 \par}}{}{}
}
\makeatother
\subtitle{A web application for easily analysing \& exploring the fabric
of archaeological assemblages}
\author{}
\date{\vspace{-2.5em}}

\begin{document}
\maketitle

\renewcommand{\[}{\begin{equation}}
\renewcommand{\]}{\end{equation}}

\textbf{Marc Thomas} \textsuperscript{1*}
(\url{https://orcid.org/0000-0002-8160-1910}) \smallbreak

\textsuperscript{1} TRACES, UMR 5608 - Université Toulouse Jean Jaurès,
5 Allée Antonio Machado, FR-31058 Toulouse Cedex 9 \smallbreak

\textsuperscript{*} Corresponding author \smallbreak

E-mail:
\href{mailto:marcthomas1@hotmail.fr}{\nolinkurl{marcthomas1@hotmail.fr}}
\smallbreak

\textbf{Key words}: R shiny web application, fabric, site formation
processes, spatial analysis, orientation and dip \bigbreak

\section{Abstract}\label{abstract}

This article presents fabryka 1.1, an open access R shiny web
application dedicated to the fabric analysis of archaeological or
geological assemblages. The fabric is a statistic made of the
orientation and dip of elongated particles and archaeological objects.
It provides information about the nature of the geological processes
involved in the formation of the deposits and their impact on the
preservation of the archaeological assemblages. This statistical
analysis is not easily accessible to non-specialists and is particularly
time-consuming, sometimes requiring the use of several software, some of
which are chargeable. fabryka uses classical analytical procedures
(Benn, rose, Schmidt and Woodcock diagrams and statistical tests on
orientations) for analysing the fabrics of objects and includes a
spatial method. New statistical methods and new ways of exploration are
proposed in the application and in this present article.\smallbreak

You can use fabryka by following this link
\url{https://marchaeologist.shinyapps.io/fabryka/} or by launching the
application in Rstudio thanks to its script. The fabryka application
files, made in R (\citeproc{ref-rcoreteam2025}{Team 2025}) by using the
R package \emph{shiny} (\citeproc{ref-chang2024}{Chang et al. 2024}),
and this article, written in rmarkdown
(\citeproc{ref-allaire2014}{Allaire et al. 2014};
\citeproc{ref-xie2018}{Xie, Allaire, and Grolemund 2018};
\citeproc{ref-xie2020}{Xie, Dervieux, and Riederer 2020}), are both
available online in a public github repository
(\url{https://github.com/marchaeologist/fabryka/tree/main}). \pagebreak

\section{Introduction}\label{introduction}

Fabric is made of the orientation and dip of elongated objects
(\citeproc{ref-bertran1997}{Bertran et al. 1997};
\citeproc{ref-bertran2002}{Bertran and Lenoble 2002};
\citeproc{ref-lenoble2004}{Lenoble and Bertran 2004}). It is a statistic
(\emph{i.e.}, it needs as much measurements as possible) that enables
the identification of sedimentary processes involved in the formation of
the deposits (\citeproc{ref-bertran1997}{Bertran et al. 1997};
\citeproc{ref-bertran2002}{Bertran and Lenoble 2002};
\citeproc{ref-lenoble2004}{Lenoble and Bertran 2004}) and their impact
on archaeological assemblages (\citeproc{ref-isaac1967}{Isaac 1967};
\citeproc{ref-schick1986}{Schick 1986}; \citeproc{ref-thomas2019}{Thomas
et al. 2019}). Objects abandoned by prehistoric people can be considered
as sedimentary particles and are affected by natural processes
(\citeproc{ref-schiffer1983}{Schiffer 1983}). As geological particles
are often not elongated enough, analysis of the fabric of sedimentary
deposits often depends mainly on archaeological material (bones and
lithics). This makes it possible to identify the sedimentary processes
to which they were subjected and to discuss their degree of spatial
reorganization. When possible, distinguishing sedimentary particles from
archaeological objects in the fabric analysis could allow to test
whether these two populations were affected by different processes
(\emph{e.g.}, the fabric of an undisturbed occupation covered by
deposits from a minor flood should reveal a distinct fabric from that of
the alluvial pebbles on which it rests).

One of the first archaeological application probably dates to the
experiments carried out by the South African G. Isaac (1967) on the
shores of Lake Magadi in Kenya. G. Isaac wanted to explain the role
played by hydrological processes on the bone accumulations found in the
Great Rift Valley. In November 1964, he placed several concentrations of
experimental material near a river branch. A few months later, heavy
rains had reactivated the valley's rivers. In addition to the
displacement or even loss of all or part of the material, G. Isaac
revealed a reorientation that was mainly perpendicular and, to a lesser
extent, parallel to the orientation of flow, with a slight upstream dip.
Since G. Isaac's pioneering work, there has been an abundant literature
devoted to the analysis of fabrics (\citeproc{ref-benn1994}{Benn 1994};
\citeproc{ref-lenoble2004}{Lenoble and Bertran 2004};
\citeproc{ref-mcpherron2018}{S. P. McPherron 2018}) allowing this method
to be increasingly performed on archaeological sites.

Numerical methods have been developed to measure the orientation and dip
of objects in the field (\citeproc{ref-mcpherron2005artifact}{S. J.
McPherron 2005}; \citeproc{ref-discamps2024}{Discamps et al. 2024}) or
from field archives such as plans and photos
(\citeproc{ref-de2013application}{Torre and Benito-Calvo 2013};
\citeproc{ref-sanchez2016assessment}{Sánchez-Romero et al. 2016}).

Two main families of statistical method are commonly used to
characterize the fabric of a given assemblage
(\citeproc{ref-bertran2002}{Bertran and Lenoble 2002};
\citeproc{ref-mcpherron2018}{S. P. McPherron 2018};
\citeproc{ref-de2021new}{Torre et al. 2021}). \newline The first methods
focus on the orientation only and aim to test the null hypothesis of a
random orientation of objects through several statistical tests
(\citeproc{ref-krumbein1939}{Krumbein 1939};
\citeproc{ref-curray1956}{Curray 1956}; \citeproc{ref-rao1967large}{Rao
1967}).\newline The second statistical method takes into account object
orientation and dip. Three normalized eigenvalues (E1, E2 and E3) are
calculated from the cosine formed by the axis of the artifact with the
three axes of space (\citeproc{ref-watson1966}{Watson 1966}). Woodcock
(\citeproc{ref-woodcock1977}{1977}) and Benn
(\citeproc{ref-benn1994}{1994}) indices are used to describe the fabric
shape. Several authors \emph{e.g.}, (\citeproc{ref-bertran1997}{Bertran
et al. 1997}; \citeproc{ref-bertran2002}{Bertran and Lenoble 2002};
\citeproc{ref-lenoble2004}{Lenoble and Bertran 2004}) built models to
compare the fabric of archaeological assemblages to natural sedimentary
processes and human processes (\emph{i.e.}, fabric of knapping spots).
The null hypothesis of an absence of perturbation is a fabric shape
compatible with that of experimental knapping spots. These methods are
described in the Mathematics section of this article.

Several software allow data processing, some of which are not freely
accessible, do not allow spatial exploration of the data or do not offer
interfaces to facilitate the task of non-programmer users
(\citeproc{ref-mcpherron2018}{S. P. McPherron 2018}). The following
sections explain all the selected approaches and our proposals for
processing the fabrics thanks to fabryka, from taking measurements in
the field, through the statistical methods and graphical modes of
representation used in the application
(fig.~\ref{fig:figure_presentation}). \bigbreak

\begin{figure} [h]
\centering
  \includegraphics[width=15cm]{figure/presentation_applications.png}
  \caption{fabryka creation scheme and organisation of the article}
\label{fig:figure_presentation}
\end{figure}

\section{Field data acquisition}\label{field-data-acquisition}

During excavation, the orientation and dip of elongated remains
(\emph{i.e.} a \textgreater{} 2 * b and with a \textgreater{} 2 cm) is
measured on a representative set of artefacts or on the whole
assemblage. The orientation (= bearing) of the lowest point is measured
in the range {[}0; 360{[} with North as the reference (\emph{i.e.}
orientation = 0°). The dip (= plunge) is measured in the range {[}0;
90{]} with horizontal as a reference. The orientation {[}0; 360{[} can
be converted into axial data measured in the range {[}0; 180{[} by
subtracting 180° from the measurements in the range {[}180; 360{]}.

Traditionally, the azimuth of the principal axis of an object is
measured using a compass considering that the point with the lowest
altitude indicates the orientation of the vector. The dip is measured
using an inclinometer. The aim is to measure the orientation and dip of
the elongaton axis of of objects ignoring its shape. To do this, the
measuring device must be placed parallel to the elongation axis of the
object. Measuring the fabric from its imprint or visible surface is
biased by its shape, making impossible to assess its true dip. Since the
theodolite was introduced to excavations in the 1990s, the fabrics is
sometimes measured by recording the coordinates of the ends of objects
(\citeproc{ref-mcpherron2005artifact}{S. J. McPherron 2005}). In the
latter case, the rules of trigonometry are applied to convert the pair
of coordinates into orientations and dips.

Recently, speleologist have created a digital inclinometer (Leica
DistoX310 modified to DistoX2; \url{https://paperless.bheeb.ch/}) which
can be used to measure fabrics (\citeproc{ref-discamps2024}{Discamps et
al. 2024}). The device stores and associates data with the order of
measurement. When the device is tilted downwards, the orientation of the
lowest point is measured and, the recorded dip is negative. When the
device is tilted upwards, the orientation of the highest point is
recorded, and the dip is positive. To obtain data that is suitable for
statistical analysis, in the first case the absolute value of the dip
must be retained. In the second case (\emph{i.e.} positive dip), if the
measured orientation is in the interval {[}0; 180{]}, 180° must be added
to the orientation, and if the measured orientation is in the interval
{]}180; 360{]}, 180° must be subtracted from the orientation.

When possible, preferring the use of a digital inclinometer avoids many
errors during measurement and data recording. On the other hand, for the
reasons given above, the use of a theodolite to measure fabrics is not
recommended as it results in inaccurate dip measurements. As well,
measuring orientations and dip in sections leads to sampling biases.

A sample corresponds to all the orientation and dip measurements taken
within a Lithostratigraphic Unit (LU) or a pedostratigraphic unit. These
units seem to be the most coherent entity which needs to be explored
because they are defined according to geological criteria, which a
priori guarantee a certain coherence in terms of the formation dynamics
we are seeking to approach. In fabryka, analysis are constrained within
samples unless the user decides otherwise in the spatial exploration
panel. \bigbreak

\section{Mathematics}\label{mathematics}

\subsection{Classical method}\label{classical-method}

Two main statistical family methods are used:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  The first considers the orientation only. The calculation of the
  Vector Magnitude \(L\) (expressed as a percentage) by Curray's
  (\citeproc{ref-curray1956}{1956}) formula tests the hypothesis of an
  unimodal distribution of orientations. If the \(p\) value returned by
  the Rayleigh test is less than the threshold of 0.05, the null
  hypothesis of a random sampling from a uniform distribution is
  rejected. The formula for calculating the Vector Magnitude \(L\) is as
  follows:
\end{enumerate}

Curray's formula to calculate \(L\) is:

\[L = \frac{r*100}{n}\] with:

\[r = \sqrt{\sum_{i=\alpha}^n \sin{(2\alpha)} + \sum_{i=\alpha}^n \cos{(2\alpha)}}\]
\smallbreak where \(n\) is the number of measures and \(\alpha\), is the
axial data (0 to 180°) \smallbreak

The Rayleigh test is performed thanks to the formula:

\[p = e^{(-(n*L^2))*10^-4}\]

The method consisting of doubling the value of the axial angles before
calculating the Vector Magnitude \(L\) and the Rayleigh test, as
proposed by Krumbein (\citeproc{ref-krumbein1939}{1939}), tests the
hypothesis of a bimodal distribution with a period of 90°. The
hypothesis of a 90° period bimodal distribution of a sample of
orientations is retained when \(p\) \textless0.05
(\citeproc{ref-curray1956}{Curray 1956}). The sensitivity of the
Rayleigh test to the sample size \(n\) prevents samples from being
compared with each other when \(n\) varies
(\citeproc{ref-curray1956}{Curray 1956};
\citeproc{ref-bertran2002}{Bertran and Lenoble 2002}). For example, when
the Vector Magnitude \(L\) is 25\%, the Rayleigh test gives \(p-values\)
below a critical threshold of 0.05 for series with more than 50
measurements (fig.~2).

The Rao (\citeproc{ref-rao1967large}{1967}) spacing test assesses the
uniformity of circular data. It determines whether the data show
significant directionality (\emph{i.e.} the null hypothesis is a uniform
distribution). It is widely used to detect multimodal directionality.

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{1}
\tightlist
\item
  The second considers the orientation and dip of the objects. The
  normalised eigenvalues \(E1\), \(E2\) and \(E3\) are calculated from
  the cosine formed by the axis of the objects with the three axes of
  space (\citeproc{ref-watson1966}{Watson 1966}).
\end{enumerate}

The fabric shape can be described by calculating the ratios \(r1\) =
ln(E1/E2), \(r2\) = ln(E2/E3) and \(K\) = r1/r2
(\citeproc{ref-woodcock1977}{Woodcock 1977}). When 0 \textless{} \(K\)
\textless{} 1, the fabric is planar. When 1 \textless{} \(K\)
\textless{} \(+\infty\), the fabric is linear. The parameter \(C\) =
ln(E1/E3) is a measure of the strength of the preferred orientation
(\citeproc{ref-woodcock1977}{Woodcock 1977}). These indices can be
plotted into a Woodcock diagram.

The Benn indices (\citeproc{ref-benn1994}{Benn 1994}) describe the
fabric shape of a sample and can be plotted in a ternary diagram
(\emph{i.e.} Benn diagram). The elongation index EL is equal to
\(1-(E2/E1)\) and the isotropy index \(IS\) is \(E3/E1\). When \(EL\) is
high, the axes of the objects are grouped around an orientation. When
\(IS\) is high, the fabric is said to be isotropic meaning the axes of
the objects are uniformly distributed in all directions. Finally, when
the value \(1-EL-IS\) is high, the fabric is said to be planar, with the
axes of the objects grouped around a plane. \bigskip

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{article_files/figure-latex/rayleigh-1.pdf}}
\caption{Evolution of the p-value returned by the Rayleigh test as a
function of the Vector Magnitude `L' and the sample size `n'
\label{rayleigh}}
\end{figure}

\subsection{Spatialised method}\label{spatialised-method}

To distinguish different statistical populations, a spatial method
proposed by McPherron (\citeproc{ref-mcpherron2018}{2018}) is used.
Thanks to this method it is possible to calculate the Benn indices and
the Vector Magnitude \(L\) (formula (\textbf{3.}) modified from
McPherron (\citeproc{ref-mcpherron2018}{2018}) R script) for each
remains, and its \(n\) nearest neighbors in 3 dimensions (McPherron
(\citeproc{ref-mcpherron2018}{2018}) R script, modified, formula
(\textbf{4.})). The analysis of the fabric as proposed by McPherron
(\citeproc{ref-mcpherron2018}{2018}) has the advantage of spatialising
the measurements. It is possible to localize the statistical signal of
fabrics by varying the number of measurements sampled to calculate the
Benn indices. The choice of about 40 to 60 measurements (each remain and
its 40 to 60 nearest neighbors) to calculate the Benn indices allows
local particularities to be observed, and above all, it allows the
Rayleigh test to be carried out in its field of application (stability
of the test between around 40 to 50 measurements after Bertran and
Lenoble (\citeproc{ref-bertran2002}{2002})).

By avoiding the instability of the Rayleigh test, the rejection rate
\(Tr\) of the null hypothesis provides a more reliable description of
the distribution of orientations in a sample. In addition, it allows
comparisons between samples despite varying sample size.

The rejection ratio \(Tr\) is performed thanks to the formula:

\[Tr = \frac{(n~times~p.value < 0.05)* 100}{n}\]

\section{Graphical representation}\label{graphical-representation}

Several methods of representation are used. Rose diagrams summaries the
orientation and dip measurements for each sample. Benn indices are
projected in a Benn diagram and measures (orientation and dip) are
projected into Schmidt diagrams. These figures are available for both
`classical' and `spatialised' methods.

As submited by S. P. McPherron (\citeproc{ref-mcpherron2018}{2018}), a
spatial investigation of the fabric is proposed in a dedicated panel.
Benn (\citeproc{ref-benn1994}{1994}) indices calculated for each spatial
series of remains are projected within a Benn diagram in which each pole
(isotropic, linear, and planar) is colored respectively in red, green
and blue. The RGB color code is reused to materialize the fabric of the
remains on spatial projections (plan and section views) of the material
along the axes of the excavation grid. \bigbreak

\section{Interface and user
interactivity}\label{interface-and-user-interactivity}

The next sections provide a description and a tutorial of fabryka. All
the figures in this section were produced using the example data set
provided in the application (\emph{i.e.} ``Case 3: Angles from DistoX2
with coordinates).''

\subsection{Upload data}\label{upload-data}

The simplest method to upload data in the application is to name the
column names directly, as shown in the example files. Enter your data in
a csv file, adding the columns `id', `code' and `sample'. If the columns
don't exist, fill them in with the same letter. By doing so, you do not
need to select your variables in the main panel. The application
correctly reads the data file directly.

First, upload your .csv file in the left sidebar panel (fig.
\ref{fig:figure_upload_your_data}A). Pay attention to your decimal and
column delimiters (fig.~\ref{fig:figure_upload_your_data}B and C).

fabryka supports three angle measurement methods:

\begin{itemize}
\tightlist
\item
  angles taken with a compass and an inclinometer
\item
  angles taken with a DistoX2
\item
  angles taken with two shots at the total station.
\end{itemize}

Five cases (\emph{i.e.} forms of data) are covered by the application
(fig. \ref{fig:figure_upload_your_data}D):

\begin{itemize}
\tightlist
\item
  check `Case 1: Only angles' if you have in your data set the columns
  `id', `sample', `code', `orientation' and `dip'.
\item
  check `Case 2: Only angles from DistoX2' if you have in your data set
  the same columns that above but your data comes from a DistoX2 and
  needs to be converted.
\item
  check `Case 3: Angles from DistoX2 with coordinates' if you have in
  your data set the same columns that above (Case 2, data from DistoX2)
  with a single set of coordinates (x, y, z).
\item
  check `Case 4: Angles and coordinates' if your data has the same form
  that above (Case 3) but is coming from a compass and an inclinometer.
\item
  check `Case 5: Two shots data without angles' if you have the columns
  `id', `sample', `code' and the angles have been measured with a total
  station (\emph{i.e.} two shots data).
\end{itemize}

According to your method of angle measurement and the presence of
coordinates you must select the right case. Please refer to the example
files in the left side bar panel (fig.
\ref{fig:figure_upload_your_data}E). It contains a data set coming from
the Lower shelter of Le Moustier (\citeproc{ref-thomas2019}{Thomas et
al. 2019}; \citeproc{ref-texier2020}{Texier et al. 2020}). The example
files directly contain the right column names. You can use them directly
in the application. Whatever the names of the columns in your file, you
can select the appropriate columns in the column selector (fig.
\ref{fig:figure_upload_your_data}F). Please keep only complete lines
(\emph{i.e.} no empty modalities, so no line without a fabric
measurement). When a data table with new columns appears below the
column selector, you are ready to use the application.

If your data comes from a DistoX2, the columns `orientation' (=
`bearing', \emph{i.e.} 0 to 360°) and `dip' (= `plunge') are
automatically corrected in the new table. Whatever the form of your data
set, two new columns `orientation\_pi' (0 to 180° angles) and
`angle\_double' (see Krumbein 1939) appear. If the angles were measured
using a total station (\emph{i.e.} two shots data or 2 by 2 coordinates
of plotted pieces refits), the orientation and the dip are calculated.
If only one set of coordinates is provided with angles, a second set of
coordinates is calculated and displayed in the new table (fig.
\ref{fig:figure_upload_your_data}G). You can download the new data by
clicking the `copy', the `csv' or the `pdf' buttons. \bigbreak

\subsection{Classical method}\label{classical-method-1}

The classical method (fig.~\ref{fig:figure_classical_method}) can be
computed for all sources of data and absence/presence of coordinates
(\emph{i.e.} all cases described above).

\begin{figure} [H]
\centering
  \includegraphics[width=16cm]{figure/upload_your_data.jpg}
  \caption{A. Browse your .csv file, B. choose your column separator, C. choose your decimal separator, D. check the right form of your dataset, E. to do so, have a look of example files, F. select the column needed in the application, G. a new dataset with new columns appears, your are ready to explore your data.}
\label{fig:figure_upload_your_data}
\end{figure}

This panel includes:

\begin{itemize}
\tightlist
\item
  Benn diagrams with possible integration of Bertran and Lenoble
  (\citeproc{ref-bertran2002}{2002}) model (fig.
  \ref{fig:figure_classical_method}D)
\item
  Rose diagrams for axial data and dip (fig.
  \ref{fig:figure_classical_method}G)
\item
  Schmidt diagrams (fig.~\ref{fig:figure_classical_method}H)
\item
  Woodcock diagrams
\item
  Summary data table with Benn indices, additional statistics, and
  orientation statistical tests.
\end{itemize}

In this last table (fig.~\ref{fig:figure_classical_method}I):

\begin{itemize}
\tightlist
\item
  \(N\) stands for number of measures in the samples
\item
  \(E1\), \(E2\) and \(E3\) are the eigenvalues
  (\citeproc{ref-watson1966}{Watson 1966}). They correspond to the sum
  of the cosines formed by the axis of the objects and the three axis of
  the space. They are used to compute the Benn indices
\item
  \(IS\) and \(EL\) are the Benn indices ratios standing respectively
  for, `isotropy' with \(IS = E3/E1\) and `elongation' with
  \(EL = 1-(E2/E1)\) (\citeproc{ref-benn1994}{Benn 1994}).
\item
  \(K\) and \(C\) are the Woodcock (\citeproc{ref-woodcock1977}{1977})
  indices.
\item
  \(L\) is the Vector Magnitude (strength of the preferred orientation)
  and \(R.p\) is the p-value result of the Rayleigh test.
\item
  \(L.double\) and \(R.p.double\) are the same statistics that above for
  double angles (\citeproc{ref-krumbein1939}{Krumbein 1939}).
\item
  \(Rao_p\) is the p\_value of the Rao
  (\citeproc{ref-rao1967large}{1967}) test.
\end{itemize}

You can control the data you plot and the data appearing in the figures
and the summary data table by changing the checkboxes filters (`filter
by sample' and `filter by code') and the slider (\emph{i.e.} minimum
number of measures) in the left sidebar panel (fig.
\ref{fig:figure_classical_method}A, B and C).

\begin{figure} [H]
\centering
  \includegraphics[width=16cm]{figure/classical_method.jpg}
  \caption{A. filter by sample, B. if you want to filter by codes, uncheck 'no code distinction', C. manage the minimum sample size to perform analysis, D. buld your model thanks to Bertran and Lenoble (2002) data, E. and F. download figures by selecting a file format and pressing the 'download' button, G. Rose diagrams subpanel, H. Schmidt diagram subpanel, I. Summary table with additional statistics}
\label{fig:figure_classical_method}
\end{figure}

Certain geological processes can have a different impact on objects
depending on their physical properties (\emph{e.g.} silex versus bones
in a water flow as in the example files provided where bones have a
strong preferential orientation and are much more `linear' in the Benn
diagram (\citeproc{ref-thomas2019}{Thomas et al. 2019};
\citeproc{ref-texier2020}{Texier et al. 2020})). The natural processes
models that you can add to the Benn diagram are calculated using data
mainly from (\citeproc{ref-bertran1997}{Bertran et al. 1997},
\citeproc{ref-bertran2006}{2006}; \citeproc{ref-bertran2002}{Bertran and
Lenoble 2002}; \citeproc{ref-lenoble2004}{Lenoble and Bertran 2004}).
You will find the reference model data (with corresponding literature
references) which is used in the application in the `Model reference
data' subpanel. In the rose diagram, axial data is considered (and not
orientations) as the Rayleigh test is perform on axial data
(\citeproc{ref-curray1956}{Curray 1956}) in the summary table. You can
download all the figures by selecting a file format (.pdf, .jpeg and
.png) and pressing the `download' button (fig.
\ref{fig:figure_classical_method}E and F).

\subsection{Spatialised method}\label{spatialised-method-1}

This panel includes (fig.~\ref{fig:figure_spatialised_method}):

\begin{itemize}
\tightlist
\item
  Benn diagrams with Benn indices computed with nearest neighbors points
  (\citeproc{ref-mcpherron2018}{S. P. McPherron 2018})
\item
  Spatial projections of the points according to their fabric shape
\item
  Summary data tables with orientation statistical tests, one at the
  scale of each spatial series (each object and its \(n\) nearest
  neighbors fabric = one row) and one at the scale of the sample (fig.
  \ref{fig:figure_spatialised_method}F and G).
\end{itemize}

A table similar to the one done in the Classical method is proposed and
contains the same statistics .

At the spatial series scale (\emph{i.e.} each points and \(n\)
neighbors), statistics are provided in a table (Spatial series data
table subpanel) with the Benn indices, the orientation rates \(L\),
\(L.double\) and \(p-values\) returned by the Rayleigh tests. At the
sample scale, mean and standard deviations values of Benn indices, mean
\(L\) and its corresponding Rayleigh tests means (mean values of all
spatial series in the sample) and rejections ratios \(Tr\) (formula
\textbf{4.}) are given in the subpanel `Summary table of samples'.

In the projections, the orientation of the objects can be materialized
by sticks (by unchecking the corresponding box in the left sidebar
panel, fig.~\ref{fig:figure_spatialised_method}D) which color is
representative of the fabric of the series formed by the object and its
\(n\) nearest neighbors. You can also add orthophoto in the background
of projections (fig.~\ref{fig:figure_spatialised_method}E). In the left
sidebar panel, you can control the samples considered in the analysis by
selecting the sample explored or by distinguishing the code used
(\emph{e.g.} flint versus bone,
fig.~\ref{fig:figure_spatialised_method}A and B). You can also use the
slider to change the number of neighbors searched for around each point
(fig.~\ref{fig:figure_spatialised_method}C). Be careful, this will alter
the results obtained in the Benn diagram and Rayleigh tests. \bigbreak

\subsection{Spatial exploration}\label{spatial-exploration}

This panel includes (fig.~\ref{fig:figure_spatial_exploration}):

\begin{itemize}
\tightlist
\item
  A spatial exploration of your data starting from a Benn diagram
\item
  A spatial exploration of your data starting from the spatial
  projections (fig.~\ref{fig:figure_spatial_exploration}J)
\item
  Summary data tables with orientation statistical tests (fig.
  \ref{fig:figure_spatial_exploration}I and K).
\end{itemize}

In this panel, you can perform a spatial analysis of your data starting
from a Benn diagram or from projections. In both cases, you need to
select the data you want to explore more locally in a new analysis. From
the projections, you can select the points you want to explore directly.
In the Benn diagram, you need to select the lasso selector (fig.
\ref{fig:figure_spatial_exploration}D). Pay attention to the number of
points selected and check that the new sample contains enough points to
carry out the new analysis. If this is not the case, you can lower the
slider in the left sidebar panel to perform the analysis with fewer
neighbors (fig.~\ref{fig:figure_spatial_exploration}E). You can download
the new samples you have selected with spatial statistics from the
corresponding subpanels (fig.~\ref{fig:figure_spatial_exploration}I and
K). \bigbreak

\begin{figure} [h]
\centering
  \includegraphics[width=16cm]{figure/spatialised_method.jpg}
  \caption{A. filter by sample, B. if you want to filter by code, uncheck 'no code distinction', C. choose the n nearest neighbors to perform the statistics, D. make sticks in spatial projections, E. upload a georeferenced orthophoto and check 'include' orthophoto to display it, F. new dataset at the spatial series scale, G. summary table at the sample scale.}
\label{fig:figure_spatialised_method}
\end{figure}

\subsection{Tutorial}\label{tutorial}

The `Tutorial' panel contains a summary of the mathematics, statistics
and methods used in fabryka and a short tutorial directly accessible
from the application. \bigbreak

\begin{figure} [h]
\centering
  \includegraphics[width=16cm]{figure/spatial_exploration.jpg}
  \caption{A. if you want to filter by sample, uncheck 'No sample distinction' B. if you want to filter by code, uncheck 'no code distinction', C. choose the n nearest neighbors to perform the statistics (first Benn diagram), E. choose the n nearest neighbors to perform the statistics on the sampled data (second Benn diagram), F. make sticks in spatial projections, G. you can upload a georeferenced orthophoto and check 'include' orthophoto to display it, H. a new spatial exploration of the fabric is performed, I. new dataset at the selected spatial series scale, I. a similar spatial exploration is possible from the projections, K. new dataset at the selected spatial series scale.}
\label{fig:figure_spatial_exploration}
\end{figure}

\subsection{New reference models}\label{new-reference-models}

The `spatialised' method generates often more linear fabrics in
comparison with the `classical' method. This can be explained by the
fact that objects that are close in space are more likely to resemble
each other than objects that are further in space. A difference between
global fabric (``classical method'', \emph{i.e.}, one point per sample)
and spatialised fabric (series of n neighbors) in Benn diagrams marks a
spatial heterogeneity that needs to be explored.

Also, the structure of the point pattern is evocative of its fabric.
Thus, the dispersion of the points depends on the shape of the fabric.
Two extremes appear: the most isotropic samples are those with the most
dispersed point patterns (\emph{i.e.} the standard deviations are high)
while the most linear samples are those with the least dispersed point
patterns. Planar samples highlight point patterns with intermediate
dispersion. The explanation of this phenomenon is intuitive: the more
similar the fabrics of objects are, the closer they are in the Benn
diagram and the more linear they are. To reflect this phenomenon, the
standard deviations of the Benn indices are calculated and aim to
describe the dispersion point patterns in the Benn diagrams.

Global planar or isotropic fabrics can sometimes mask local anisotropy.
Within certain samples, groups of objects form a dense cluster in the
point pattern. This is expressed either by a locally more linear fabric,
or by a planar fabric with a strong bimodal orientation.

For the reasons given above, we wish to create new models that
considered the particularities of the spatial method. A panel was
therefore set up to collect new data describing the fabric of
sedimentary processes in order to build a new ``spatialised'' reference
model (\emph{cf.} the ``classical'' model proposed by Bertran and
Lenoble (\citeproc{ref-bertran2002}{2002})). This work is in progress.
This panel will be updated in the next version of the application.

\bigbreak

\subsection{Contributors}\label{contributors}

The contributors panel aims at citing all contributors of this project.
\bigbreak

\section{Data, scripts, code, and supplementary information
availability}\label{data-scripts-code-and-supplementary-information-availability}

\begin{itemize}
\tightlist
\item
  use fabryka online: \url{https://marchaeologist.shinyapps.io/fabryka/}
  \smallbreak
\item
  source code: \url{https://github.com/marchaeologist/fabryka/tree/main}
  \smallbreak
\item
  Report bugs or any request:
  \url{https://github.com/marchaeologist/fabryka/issues} \smallbreak
\item
  github reprository doi of fabryka:
  \url{https://doi.org/10.5281/zenodo.15909666} \smallbreak
\item
  Zenodo repository of this article:
  \url{https://doi.org/10.5281/zenodo.16758152} \bigbreak
\end{itemize}

\textbf{If you are preparing a mission in a field without Internet
access:}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Install \href{https://www.r-project.org}{R} and
  \href{https://posit.co/download/rstudio-desktop/}{Rstudio Desktop}.
\item
  Clone the fabryka repository from GitHub:
  \url{https://github.com/marchaeologist/fabryka} or download the code
  in ``Code'' and ``Download ZIP'' at the same address. Unzip the
  folder. This will give you the latest version of the source code.
\item
  Open fabryka.Rproj
\item
  If you are using the application for the first time, you have to
  install several packages used in the application. Lists of all these
  packages and dependencies are available in a DESCRIPTION file and a
  DEPENDECIES file in the github repository. To install fabryka packages
  and dependencies, copy and paste the following code lines into the R
  console:
\end{enumerate}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{fabryka\_packages }\OtherTok{=} \FunctionTok{c}\NormalTok{(}\StringTok{"shiny"}\NormalTok{, }\StringTok{"dplyr"}\NormalTok{,}\StringTok{"ggplot2"}\NormalTok{,}\StringTok{"plotly"}\NormalTok{, }\StringTok{"ggsci"}\NormalTok{, }\StringTok{"ggtern"}\NormalTok{, }\StringTok{"circular"}\NormalTok{, }
                     \StringTok{"CircStats"}\NormalTok{, }\StringTok{"Ternary"}\NormalTok{, }\StringTok{"RStoolbox"}\NormalTok{, }\StringTok{"shinyjs"}\NormalTok{, }\StringTok{"rmarkdown"}\NormalTok{, }
                     \StringTok{"shinythemes"}\NormalTok{,  }\StringTok{"ggalt"}\NormalTok{, }\StringTok{"DT"}\NormalTok{, }\StringTok{"raster"}\NormalTok{)}

\NormalTok{is\_installed }\OtherTok{=} \FunctionTok{sapply}\NormalTok{(fabryka\_packages, require, }\AttributeTok{character.only=}\NormalTok{T)}

\FunctionTok{sapply}\NormalTok{(fabryka\_packages[}\SpecialCharTok{!}\NormalTok{is\_installed], install.packages)}
\end{Highlighting}
\end{Shaded}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{4}
\tightlist
\item
  The application can be launched by opening the app.R file in the main
  folder and pressing ``Run App'' or by running the following commands:
\end{enumerate}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{pkgload}\SpecialCharTok{::}\FunctionTok{load\_all}\NormalTok{(}\StringTok{"."}\NormalTok{)}

\FunctionTok{fabryka}\NormalTok{()}
\end{Highlighting}
\end{Shaded}

\section{Acknowledgements}\label{acknowledgements}

The writing of the script was greatly facilitated by the work published
by Shannon McPherron (2018). It includes some functions published by the
author, with a few modifications. This work benefited from discussions
with Jean-Pierre Texier, Arnaud Lenoble, Pascal Bertran, Thomas Perrin
and Dominique Todisco on issues related to site formation processes. The
reference data used to build the models of the classical method were
collected by P. Bertran throughout his career. Pascal Bertran provided
us with all these data in order to feed the application. The development
of the application also benefited from the help of Rémi Lemoy, François
Baleux and Victor Baleux on mathematical issues. A first version of the
application was tested by Emmanuel Discamps and Aurélien Royer. This
project is supported by the ``Service Valorisarion de la Recherche'' of
the University Toulouse Jean Jaurès. A preprint version of this article
(v3, \url{https://doi.org/10.5281/zenodo.16601423}) has been
peer-reviewed and recommended by PCI Archaeo
(\url{https://doi.org/10.24072/pci.archaeo.100611}). Comments made by
Pascal Bertran, recommender of this article, and, by Frédéric Santos,
Nicolas Frerebeau and Alfonso Benito-Calvo, reviewers of this article,
considerably improved the manuscript and the application. I would like
to thank all these reasearchers for their advice. \bigbreak

\section{Funding}\label{funding}

This research was funded by the Service Valorisation de la Recherche of
Toulouse - Jean Jaurès University. \bigbreak

\section{Conflict of interest
disclosure}\label{conflict-of-interest-disclosure}

The authors declare that they comply with the PCI rule of having no
financial conflicts of interest in relation to the content of the
article. \bigbreak

\section{References}\label{references}

The `Bibliography' panel in the application contains the references
cited in the application. This article contains the same bibliography
with some additional references. \bigbreak

\phantomsection\label{refs}
\begin{CSLReferences}{1}{0}
\bibitem[\citeproctext]{ref-allaire2014}
Allaire, Joseph J, Yihui Xie, Christophe Dervieux, Jonathan McPherson,
Javier Luraschi, Kevin Ushey, Aron Atkins, et al. 2014. {``Rmarkdown:
Dynamic Documents for r.''} \emph{(No Title)}.

\bibitem[\citeproctext]{ref-benn1994}
Benn, Douglas. 1994. {``Fabric Shape and the Interpretation of
Sedimentary Fabric Data.''} \emph{Journal of Sedimentary Research} 64
(4a): 910915.

\bibitem[\citeproctext]{ref-bertran2006}
Bertran, Pascal, Jean-Guillaume Bordes, Antoine Barré, Arnaud Lenoble,
and Vincent Mourre. 2006. {``Fabrique d'amas de Débitage: Données
Expérimentales.''} \emph{Bulletin de La Société Préhistorique
Française}, 3347.

\bibitem[\citeproctext]{ref-bertran1997}
Bertran, Pascal, Bernard Hetu, Jean-Pierre Texier, and Henk Van Steijn.
1997. {``Fabric Characteristics of Subaerial Slope Deposits.''}
\emph{Sedimentology} 44 (1): 116.

\bibitem[\citeproctext]{ref-bertran2002}
Bertran, Pascal, and Arnaud Lenoble. 2002. {``Fabriques Des Niveaux
Archéologiques: Méthode Et Premier Bilan Des Apports à l{'}étude
Taphonomique Des Sites Paléolithiques.''} \emph{PALEO. Revue
d'archéologie Préhistorique}, no. 14: 1328.

\bibitem[\citeproctext]{ref-chang2024}
Chang, Winston, Joe Cheng, J. J. Allaire, Carson Sievert, Barret
Schloerke, Yihui Xie, Jeff Allen, Jonathan McPherson, Alan Dipert, and
Barbara Borges. 2024. \emph{Shiny: Web Application Framework for r}.
\url{https://CRAN.R-project.org/package=shiny}.

\bibitem[\citeproctext]{ref-curray1956}
Curray, Joseph R. 1956. {``The Analysis of Two-Dimensional Orientation
Data.''} \emph{The Journal of Geology} 64 (2): 117131.

\bibitem[\citeproctext]{ref-discamps2024}
Discamps, Emmanuel, Marc Thomas, Jean-Philippe Faivre, and Brad Gravina.
2024. {``Protocole de Fouille Sur Les Sites Paléolithiques:
Justification Des Besoins Et Modalités de Mise En Øeuvre à Combe-Grenal
Et à l{'}abri Inférieur Du Moustier (Dordogne, France).''} \emph{Gallia
Préhistoire}, no. 64.

\bibitem[\citeproctext]{ref-isaac1967}
Isaac, Glynn L. 1967. {``Towards the Interpretation of Occupation
Debris: Some Experiments and Observations.''} \emph{Kroeber
Anthropological Society Papers} 37 (37): 3157.

\bibitem[\citeproctext]{ref-krumbein1939}
Krumbein, William C. 1939. {``Preferred Orientation of Pebbles in
Sedimentary Deposits.''} \emph{The Journal of Geology} 47 (7): 673706.

\bibitem[\citeproctext]{ref-lenoble2004}
Lenoble, Arnaud, and Pascal Bertran. 2004. {``Fabric of Palaeolithic
Levels: Methods and Implications for Site Formation Processes.''}
\emph{Journal of Archaeological Science} 31 (4): 457469.

\bibitem[\citeproctext]{ref-mcpherron2005artifact}
McPherron, Shannon JP. 2005. {``Artifact Orientations and Site Formation
Processes from Total Station Proveniences.''} \emph{Journal of
Archaeological Science} 32 (7): 1003--14.

\bibitem[\citeproctext]{ref-mcpherron2018}
McPherron, Shannon P. 2018. {``Additional Statistical and Graphical
Methods for Analyzing Site Formation Processes Using Artifact
Orientations.''} \emph{PloS One} 13 (1): e0190195.

\bibitem[\citeproctext]{ref-rao1967large}
Rao, JS. 1967. {``Large Sample Tests for the Homogeneity of Angular
Data.''} \emph{Sankhya Ser B} 28: 172--74.

\bibitem[\citeproctext]{ref-sanchez2016assessment}
Sánchez-Romero, Laura, Alfonso Benito-Calvo, Alfredo Pérez-González, and
Manuel Santonja. 2016. {``Assessment of Accumulation Processes at the
Middle Pleistocene Site of Ambrona (Soria, Spain). Density and
Orientation Patterns in Spatial Datasets Derived from Excavations
Conducted from the 1960s to the Present.''} \emph{PloS One} 11 (12):
e0167595.

\bibitem[\citeproctext]{ref-schick1986}
Schick, Kathy Diane. 1986. {``Stone Age Sites in the Making: Experiments
in the Formation and Transformation of Archaeological Occurrences.''}
\emph{(No Title)}.

\bibitem[\citeproctext]{ref-schiffer1983}
Schiffer, Michael B. 1983. {``Toward the Identification of Formation
Processes.''} \emph{American Antiquity} 48 (4): 675706.

\bibitem[\citeproctext]{ref-rcoreteam2025}
Team, R Core. 2025. \emph{R: A Language and Environment for Statistical
Computing}. Vienna, Austria: R Foundation for Statistical Computing.
\url{https://www.R-project.org/}.

\bibitem[\citeproctext]{ref-texier2020}
Texier, Jean-Pierre, Emmanuel Discamps, Brad Gravina, and Marc Thomas.
2020. {``Les Dépôts de Remplissage de l{'}abri Inférieur Du Moustier
(Dordogne, France): Lithostratigraphie, Processus de Formation Et
Évolution Du Système Géomorphologique.''} \emph{PALEO. Revue
d'archéologie Préhistorique}, no. 30-2: 320345.

\bibitem[\citeproctext]{ref-thomas2019}
Thomas, Marc, Emmanuel Discamps, Brad Gravina, and Jean-Pierre Texier.
2019. {``Analyse Taphonomique Et Spatiale de Palimpsestes d'occupations
Moustériennes de l'abri Inférieur Du Moustier (Dordogne, France).''}
\emph{PALEO. Revue d'archéologie Préhistorique}, no. 30-1: 278299.

\bibitem[\citeproctext]{ref-de2013application}
Torre, Ignacio de la, and Alfonso Benito-Calvo. 2013. {``Application of
GIS Methods to Retrieve Orientation Patterns from Imagery; a Case Study
from Beds i and II, Olduvai Gorge (Tanzania).''} \emph{Journal of
Archaeological Science} 40 (5): 2446--57.

\bibitem[\citeproctext]{ref-de2021new}
Torre, Ignacio de la, Alfonso Benito-Calvo, Carmen Martin-Ramos, Lindsay
J McHenry, Rafael Mora, Jackson K Njau, Michael C Pante, Ian G
Stanistreet, and Harald Stollhofen. 2021. {``New Excavations in the MNK
Skull Site, and the Last Appearance of the Oldowan and Homo Habilis at
Olduvai Gorge, Tanzania.''} \emph{Journal of Anthropological
Archaeology} 61: 101255.

\bibitem[\citeproctext]{ref-watson1966}
Watson, Geoffrey S. 1966. {``The Statistics of Orientation Data.''}
\emph{The Journal of Geology} 74 (5, Part 2): 786797.

\bibitem[\citeproctext]{ref-woodcock1977}
Woodcock, NH. 1977. {``Specification of Fabric Shapes Using an
Eigenvalue Method.''} \emph{Geological Society of America Bulletin} 88
(9): 12311236.

\bibitem[\citeproctext]{ref-xie2018}
Xie, Yihui, Joseph J Allaire, and Garrett Grolemund. 2018. \emph{R
Markdown: The Definitive Guide}. Chapman; Hall/CRC.

\bibitem[\citeproctext]{ref-xie2020}
Xie, Yihui, Christophe Dervieux, and Emily Riederer. 2020. \emph{R
Markdown Cookbook}. Chapman; Hall/CRC.

\end{CSLReferences}

\end{document}
